---
title: "Lecture [10b] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1 (Tidy Selection)

a.    Use tidy selection to extract all variables that have names ending with "_color" from the `starwars` tibble.

b.    Use tidy selection to extract the `name` variable and all variables from the `species` variable to the last variable in the `starwars` tibble.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
#| message: false

library(tidyverse)
starwars |> 
  select(ends_with("_color"))
```

**Answer (b)**

```{r}
starwars |> 
  select(name, species:last_col())
```

</blockquote></details></p>

## Question 2 (Across)

a.    Use `across()` to apply the `str_to_upper()` function to the `manufacturer` and `model` variables of the `mpg` tibble.

b.    Use `across()` to summarize the median of every numeric variable in the `diamonds` tibble.


<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
mpg |> 
  mutate(
    across(
      .cols = c(manufacturer, model), 
      .fns = str_to_upper
    )
  )
```

**Answer (b)**

```{r}
diamonds |> 
  summarize(
    across(
      .cols = where(is.numeric), 
      .fns = median
    )
  )
```


</blockquote></details></p>

## Question 3 (Scale Scores)

a.    Add a new column to the `diamonds` tibble called `volume` that contains the rectangular volume of each diamond (i.e., the product or multiplication of the `x`, `y`, and `z` variables)

b.    Add a new column to the `diamonds` tibble called `mdim` that contains the average dimension size of each diamond (i.e., the mean of the `x`, `y`, and `z` variables)

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

```{r}
diamonds |> 
  mutate(volume = x * y * z)
```

**Answer (b)**

```{r}
diamonds |> 
  mutate(mdim = rowMeans(pick(x, y, z)))
```

**Alternate Answer (b)**

```{r}
diamonds |> 
  # Create row identifier
  mutate(row_id = row_number()) |> 
  # Mutate while grouping by it
  mutate(
    mdim = mean(c(x, y, z)),
    .by = row_id
  )
```

</blockquote></details></p>

## Challenge Question

a.    Look at the code below and figure out how it works.

b.    Adapt this code to create four new variables called `carat_minimum`, `carat_maximum`, `price_minimum`, and `price_maximum`.

```{r}
diamonds |> 
  summarize(
    across(
      .cols = c(carat, price),
      .fns = list(m = mean, s = sd),
      .names = "{.fn}_{.col}"
    )
  )
```

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
diamonds |> 
  summarize(
    across(
      .cols = c(carat, price),
      .fns = list(minimum = min, maximum = max),
      .names = "{.col}_{.fn}"
    )
  )
```

</blockquote></details></p>


---

*Note that there is no need to turn in Activities. These are just for practice!*
