---
title: "Lecture [11a] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1

Examine the `table5` tibble (included in \{tidyverse\}) and tidy it up using `separate()` and `unite()`. End with variables for country, four-digit year (YYYY), number of tuberculosis cases, and population.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false

library(tidyverse)
table5 |> 
  unite(
    col = "year",
    century, 
    year, 
    sep = ""
  ) |> 
  separate_wider_delim(
    col = rate, 
    delim = "/",
    names = c("cases", "population")
  )
```

</blockquote></details></p>


## Question 2

Reshape the `billboard` tibble (included in \{tidyverse\}) to a long format such that each row represents a single track’s ranking for a single week (variables: `artist`, `track`, `date.entered`, `week`, `rank`). For a challenge, remove the prefix from your final week variable and convert it into a numeric variable.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
billboard |> 
  pivot_longer(
    cols = starts_with("wk"), 
    names_to = "week", 
    values_to = "rank",
    names_prefix = "wk", # Challenge
    names_transform = parse_number # Challenge
  )
```

</blockquote></details></p>


## Question 3

Reshape the `table2` tibble (included in \{tidyverse\}) so that each row represents a single country in a single year (Hint: variables should be `country`, `year`, `cases`, and `population`).

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
table2 |> 
  pivot_wider(names_from = type, values_from = count)
```

</blockquote></details></p>


## Optional Challenge

Reshape the `us_rent_income` tibble (included in \{tidyverse\}) so that each row represents a single US state (Hint: Variables should be something like: `GEOID`, `NAME`, `estimate_income`, `estimate_rent`, `moe_income`, and `moe_rent`)

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
us_rent_income |> 
  pivot_wider(
    names_from = variable, 
    values_from = c(estimate, moe)
  )
```

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
