---
title: "Lecture [11b] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1

Install and load the \{Lahman\} package, which contains data about baseball teams and players.

a. Explore how the `People`, `Batting`, and `Salaries` tables are connected.

b. How would you join this data to compare the salaries of right-handed and left-handed batters? Do so using a \{tidyverse\} function and select only the relevant variables.

*Note: These tables are data frames not tibbles (so they will print a **lot** of rows). To fix this, you can use `as_tibble()` to convert them to tibbles.*

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**

All three tables are connected via the `playerID` variable.

**Answer (b)**

```{r}
#| warning: false
#| message: false

library(tidyverse)
library(Lahman) # must be installed first

joined_data <- 
  inner_join(People, Salaries, by = "playerID") |>  
  select(playerID, nameFirst, nameLast, bats, yearID, teamID, salary) |>
  as_tibble() |> 
  print()
```


</blockquote></details></p>

## Question 2

a. Join the `band_members` and `band_instruments` tibbles (included in \{tidyverse\}) using all four basic types of joins. Compare the results.

b. Now join the `band_members` and `band_instruments2` tibbles in a way that adds the instrument played by each name in `band_members.`

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Answer (a)**
```{r}
left_join(band_members, band_instruments, by = "name")
right_join(band_members, band_instruments, by = "name")
inner_join(band_members, band_instruments, by = "name")
full_join(band_members, band_instruments, by = "name")
```

**Answer (b)**
```{r}
left_join(band_members, band_instruments2, by = c("name" = "artist"))
```



</blockquote></details></p>

## Question 3

Explore what happens when you join two tables with duplicated/repeated keys.

```{r}
x <- tibble(
  key = c(1, 2, 2, 3),
  val_x = c("x1", "x2", "x3", "x4")
)
y <- tibble(
  key = c(1, 2, 2, 3),
  val_y = c("y1", "y2", "y3", "y4")
)
```


<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
left_join(x, y, by = "key")
```

The key gets duplicated for each pairwise combination.

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
