---
title: "Lecture [12a] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1

Create a ggplot from `mpg` that maps `displ` to x and `hwy` to y and includes a point geom. Facet the plot such that there are seven columns showing `class` levels and three rows showing `drv` levels.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false

library(tidyverse)
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point() + 
  facet_grid(drv~class)
```


</blockquote></details></p>

## Question 2

Create a stacked bar chart from `mpg` that shows the proportion of each `class` of car that uses each type of drive train (`drv`). Use this plot to answer: which class of car uses only front-wheel drive (i.e., `"f"`) and which class of car is the most balanced in terms of using different types of drive trains? *Hint:* All bars should be the same height.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
mpg |> 
  ggplot(aes(x = class, fill = drv)) + 
  geom_bar(position = "fill")
```

All **minivans** in this dataset use the front-wheel drive type of drive train, and the **subcompact** class is the most balanced.
</blockquote></details></p>

## Question 3

Create a ggplot from `mpg` that maps `hwy` to x and `manufacturer` to y; use `stat_summary()` to plot a pointrange geom for each manufacturer that puts the point at the median value and extends the range to the minimum and maximum values. Use this plot to answer: which two manufacturers exceeded 25 mpg highway fuel efficiency for all their cars in this dataset?

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
mpg |> 
  ggplot(aes(x = hwy, y = manufacturer)) +
  stat_summary(
    fun = median,
    fun.min = min,
    fun.max = max,
    geom = "pointrange"
  )

# Hidden shortcut function :)
mpg |> 
  ggplot(aes(x = hwy, y = manufacturer)) +
  stat_summary(
    fun.data = median_hilow,
    geom = "pointrange"
  )
```
**Pontiac** and **Honda** both had minimum `hwy` values (i.e., lower ranges) above 25.

</blockquote></details></p>

## Challenge

Recreate the figure below from the `economics` tibble (included in \{tidyverse\}). *Hint:* This will require some reshaping as well as faceting.

```{r}
#| echo: false

economics |> 
  pivot_longer(
    pce:unemploy, 
    names_to = "variable", 
    values_to = "value"
  ) |> 
  ggplot(aes(x = date, y = value)) + 
  geom_line() +
  facet_grid(variable~., scales = "free_y")
```

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| eval: false

economics |> 
  pivot_longer(
    pce:unemploy, 
    names_to = "variable", 
    values_to = "value"
  ) |> 
  ggplot(aes(x = date, y = value)) + 
  geom_line() +
  facet_grid(variable~., scales = "free_y")
```

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
