---
title: "Lecture [12b] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1

I am interested in buying a used SUV from 2008. Filter the `mpg` dataset to only include cars from that year and then create a jittered scatterplot with `hwy` on x and `cty` on y. Highlight in red the cars that are from the "suv" class.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false

library(tidyverse)
library(gghighlight)

mpg |> 
  filter(year == 2008) |> 
  ggplot(aes(x = hwy, y = cty)) +
  geom_jitter(color = "red") +
  gghighlight(class == "suv")
```


</blockquote></details></p>

## Question 2

Recreate this figure from `mpg` using \{ggplot2\} and \{patchwork\}. *Hint:* The geoms (left-to-right) are density, bar, density, boxplot, boxplot.

*Bonus:* Make the top row half the height of the bottom row and add tags "a" to "e".

```{r}
#| message: false
#| echo: false

library(patchwork)
p1 <- ggplot(mpg, aes(x = hwy)) + geom_density()
p2 <- ggplot(mpg, aes(x = drv)) + geom_bar()
p3 <- ggplot(mpg, aes(x = cty)) + geom_density()
p4 <- ggplot(mpg, aes(x = drv, y = hwy)) + geom_boxplot()
p5 <- ggplot(mpg, aes(x = drv, y = cty)) + geom_boxplot()
((p1 + p2 + p3) / (p4 + p5))
```

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false

library(patchwork)
p1 <- ggplot(mpg, aes(x = hwy)) + geom_density()
p2 <- ggplot(mpg, aes(x = drv)) + geom_bar()
p3 <- ggplot(mpg, aes(x = cty)) + geom_density()
p4 <- ggplot(mpg, aes(x = drv, y = hwy)) + geom_boxplot()
p5 <- ggplot(mpg, aes(x = drv, y = cty)) + geom_boxplot()
((p1 + p2 + p3) / (p4 + p5))
```

**Bonus**

```{r}
((p1 + p2 + p3) / (p4 + p5)) + 
  plot_layout(heights = c(1, 2)) +
  plot_annotation(tag_levels = "a")
```

</blockquote></details></p>

## Question 3

Create a document in PowerPoint (or equivalent program, e.g., Google Slides) and draw a box where you would put a figure. Examine that box's size and save a ggplot in that same size. Insert your image into PowerPoint.

*Bonus:* Adjust the font size of your plot to be more readable in PowerPoint.

```{r}
#| eval: false

# Simple example plot
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
```


<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| eval: false

p <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
ggsave("act3a.png", plot = p, width = 5, height = 3, units = "in")
```

**Bonus**

```{r}
#| eval: false

p <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() +
  theme_grey(base_size = 18)
ggsave("act3b.png", plot = p, width = 5, height = 3, units = "in")
```

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
