---
lecture: "12a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} |
{{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 24))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Understand how R handles strings and special characters

- Combine, separate, and extract parts of text data

- Use regular expressions to detect and replace patterns within strings
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Working with Strings

## Text Data in R {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Question:** *How does R handle text?*

-   Text data is stored as **strings** (character vectors)
-   We create strings by wrapping text in double quotes (`"as so"`) or single quotes (`'as so'`)

::: {.fragment .mt1}
-   We will use the \{stringr\} package (part of the \{tidyverse\}). All of its functions start with `str_` and are built specifically to handle messy text
:::

::: {.fragment .mt1}
-   We will learn the basics of regular expressions (regex) to search for complex patterns in text
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/text_box.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Creating Strings & Escapes

::: {.fsmaller}
**Question:** *How do we define text data in R?*
:::

```{r}
x <- "We usually create strings by wrapping text in double quotes."
writeLines(x)
```

:::{.fragment}
```{r}
y <- "Backslashes (\\) \"escape\" special symbols, treating them as characters."
writeLines(y)
```
:::
:::{.fragment}
```{r}
z <- "Backslashes can also create: \nNewlines (\\n) and \ttabs (\\t)"
writeLines(z)
```
:::

## Combining Strings

::: {.fsmaller}
**Question:** *How do we stick multiple strings together?*
:::

```{r}
#| warning: false
df <- 
  tibble(
    first = c("John", "Jane", "Alice"),
    last = c("Doe", "Smith", "Johnson")
  ) |> 
  print()
```

## Combining Strings

::: {.fsmaller}
**Question:** *How can we flexibly combine strings?*

Yes, `str_glue()` lets us insert R variables directly into a string using curly braces `{}`. This is often much easier to read than `str_c()`.
:::

```{r}
df |> 
  mutate(
    greeting = str_glue("Hello, {first} {last}! Welcome back.")
  )
```

## Flattening Strings

::: {.fsmaller}
**Question:** *How do we collapse a whole column into a single string?*

We can use `str_flatten()` to combine a vector of multiple strings into one single string. We can also specify a separator.
:::

```{r}
df |> 
  summarize(
    all_names = str_flatten(first, collapse = ", ")
  )
```

## Real-World Application

::: {.fsmaller}
**Question:** *Which carriers operate out of each NYC airport?*
:::

```{r}
data("flights", package = "nycflights13")
flights |> 
  summarize(
    .by = origin,                   # For each origin airport...
    carriers = 
      carrier |>                    # Find its active carriers
      unique() |>                   # Remove all duplicates
      sort() |>                     # Sort them alphanumerically
      str_flatten(collapse = ", ")  # Flatten them into one string
  )
```

# Extracting & Separating

## Subsetting Strings

::: {.fsmaller}
**Question:** *How do we extract a specific range of characters?*
:::

```{r}
# Hex color codes use #RRGGBB for red, green, and blue intensities
df <- 
  tibble(
    hex = c("#FF0000", "#00FF00", "#0000FF")
  ) |> 
  print()
```

## Subsetting Strings

::: {.fsmaller}
We use `str_sub()`, which takes a `start` and `end` position. We can even use negative numbers to count backward from the end of the string.
:::

```{r}
df |> 
  mutate(
    red = str_sub(hex, start = 2, end = 3),     # 2 from left to 3 from left 
    green = str_sub(hex, start = -4, end = -3), # 4 from right to 3 from right
    blue = str_sub(hex, start = 6, end = -1)    # 6 from left to 1 from right
  )
```

## Separating Columns

::: {.fsmaller}
**Question:** *How do we split one messy column into multiple clean ones?*
:::

```{r}
df <- 
  tibble(
    id = 1:3,
    blood_pressure = c("120/80", "135/90", "110/70")
  ) |> 
  print()
```

## Separating Columns

```{r}
df |> 
  separate_wider_delim(
    cols = blood_pressure,              # original/messy col name
    delim = "/",                        # character to separate on
    names = c("systolic", "diastolic")  # new col names to create
  )
```

## Lengthening Columns

::: {.fsmaller}
**Question:** *Can we separate into rows instead of columns?*
:::

```{r}
df <- 
  tibble(
    name = c("Alice", "Bob"),
    classes = c("Math, Science", "History, Art, Math")
  ) |> 
  print()
```

## Lengthening Columns

```{r}
df |> 
  separate_longer_delim(
    cols = classes, 
    delim = ", "
  )
```

## Real-World Application

::: {.fsmaller}
**Question:** *Why is this useful in data science?*
:::

```{r}
df <- tibble(sku = c("US-APP-2023-11", "CA-BAN-2024-01"))
```

:::{.fragment}
```{r}
df |> 
  separate_wider_delim(
    cols = sku, 
    names = c("country", "product", "year", "month"), 
    delim = "-"
  )
```
:::

# Regular Expressions

## Intro to Regex {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Question:** *What if our target isn't a fixed position or a simple delimiter?*

-   [Regular Expressions]{.b .blue} (regex) are a concise language for describing patterns in strings. They are very powerful but can look like gibberish

::: {.fragment .mt1}
-   Most \{stringr\} functions take in a regex pattern
-   We will use `str_view()` to preview what our regex is matching. Matches will be surrounded by angle brackets: `<...>`
-   Then we will learn how to apply them
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/search.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Exact Matching

::: {.fsmaller}
The simplest regex pattern is just an exact match of characters.
:::

```{r}
x <- c("apple", "banana", "pear", "pineapple")

# Show me elements that contain "apple"
str_view(x, "apple")
```

## Anchors

::: {.fsmaller}
**Question:** *What if we only want matches at the start or end of a string?*

We use anchors. `^` matches the start of the string, and `$` matches the end.
:::

```{r}
# Show me elements that START with "a"
str_view(x, "^a")

# Show me elements that END with "e"
str_view(x, "e$")
```

## Regex Building Blocks

::: {.fsmaller}
**Question:** *How do we match flexible patterns instead of exact matches?*
:::

::: {.fsmaller}
-   **Wildcards (`.`):** Matches *any* single character
-   **Character Classes (`[]`):** Matches any *one* of the characters listed
-   **Shortcuts:** Built-in classes for common patterns
    + e.g., `\\d` (digits) and `\\s` (spaces)
-   **Quantifiers:** Control how many times a pattern repeats
    + `{n}`: matches exactly *n* times
    + `+`: matches one or more times

:::

## Examples

```{r}
x <- c("sun", "sand", "soon", "July 4", "2026")
```

:::{.columns .mt1}
:::{.column}
```{r}
# Single Wildcard
str_view(
  string = x, 
  pattern = "s.n"
)
```
:::{.fragment}
```{r}
# One or More Wildcards
str_view(
  string = x, 
  pattern = "s.+n"
)
```
:::
:::
:::{.column}
:::{.fragment}
```{r}
# Spaces
str_view(
  string = x, 
  pattern = "\\s"
)
```
:::
:::{.fragment}
```{r}
# Exactly 4 digits
str_view(
  string = x, 
  pattern = "\\d{4}"
)
```
:::
:::
:::

## Detecting Patterns

::: {.fsmaller}
**Question:** *How do we filter a dataset based on text patterns?*
:::

```{r}
df <- 
  tibble(name = c("Ann", "Andy", "Amanda", "Bob", "Cindy")) |> 
  print()
```

## Detecting Patterns

::: {.fsmaller}
We can use `str_detect()` to ask if the pattern was detected
:::

:::{.columns}
:::{.column}
```{r}
# Is "an" or "An" detected?
df |> 
  mutate(
    match = str_detect(name, "[aA]n")
  )
```
:::
:::{.column .fragment}
```{r}
# Retain row if it is detected
df |>
  filter(
    str_detect(name, "[aA]n")
  )
```

:::
:::

## Replacing Patterns

::: {.fsmaller}
**Question:** *How do we fix or replace specific text elements?*
:::

```{r}
(x <- "We often utilize fancy words when we could utilize simple ones.")
```

:::{.fragment}
```{r}
# Replace the FIRST instance of the pattern only
str_replace(x, pattern = "utilize", replacement = "use")
```
:::

:::{.fragment}
```{r}
# Replace all instances of the pattern
str_replace_all(x, pattern = "utilize", replacement = "use")
```
:::

## Removing Patterns

::: {.fsmaller}
**Question:** *How do we delete parts of a string?*
:::

```{r}
(x <- c("(555) 123-4567", "800-555-0199"))
```

:::{.fragment}
```{r}
# Remove any of the following: [...]
#   Parentheses: ()
#   Hyphens: \\-
#   Spaces: \\s
str_remove_all(x, "[()\\-\\s]")
```
:::

## Removing Whitespace

```{r}
x <- "\t  Sometimes strings have\n   too   much white space "
writeLines(x)
```

:::{.fragment}
```{r}
# Forcibly remove all whitespace
str_remove_all(x, "\\s")
```
:::

:::{.fragment}
```{r}
# Remove outer whitespace and compress inner whitespace
str_squish(x)
```
:::

## Real-World Application

::: {.fsmaller}
**Question:** *Why is this useful in data science?*

Data entered by humans in free-text fields is notoriously messy. We can string these tools together to sanitize it.
:::

```{r}
df <- 
  tibble(
    email = c("jane AT uni DOT edu", "\njohn@gmail.com ", 
              "educator@khan.org", "SMITH@  college.edu ")
  ) |> 
  print()
```

## Real-World Application

```{r}
df |> 
  mutate(
    clean_email = 
      email |> 
      str_squish() |>               # Collapse whitespace
      str_replace(" AT ", "@") |>   # Replace AT with @
      str_replace(" DOT ", ".") |>  # Replace DOT with .
      str_to_lower() |>             # Make all lowercase
      str_remove_all("\\s"),        # Remove all spaces
    dotedu = str_detect(clean_email, "\\.edu$")
  )
```
