---
title: "Week 12 Assignment"
format: 
  html:
    css: ../../styles.css
    toc: false
---

![](../../img/programming_2780E3.svg){width="200"}

## Instructions

*Create a new Quarto document (HTML format), clear out the boilerplate code/text, and use this document to store your answers to each of the six questions below. Be sure to enable `embed-resources` using the header below. If your answer involves writing text, then use markdown to write 2-4 sentences per question; if your answer involves running code, then use one or more code chunks.*

```markdown
---
title: "Week 12 Assignment"
author: "Your Name"
embed-resources: true
---
```

*Please clearly mark which part of your document is answering which question (e.g., using markdown headings) and remove any "scratch work" or failed attempts from your document before submitting it. Finally, note that many essay-style questions will require you to read between the lines and use your imagination (as opposed to just looking up the answer in the lecture).*

## Question 1 (Combining and Flattening Strings)

**Background:** R provides multiple ways to combine text, depending on whether you want to stitch columns together row-by-row or collapse an entire column into a single string.

**a) Conceptual**

What is the conceptual difference between `str_glue()` and `str_flatten()`? How does their typical use case within a `mutate()` step versus a `summarize()` step differ?

**b) Applied**

Load the `msleep` dataset from \{ggplot2\} (included in \{tidyverse\}). First, use `filter()` to keep only the animals where `vore` is `"carni"` (carnivores). Next, use `summarize()` and `str_flatten()` to create a single string containing all the carnivore names separated by a comma and a space (e.g., "Cheetah, Northern fur seal, ..."). Print the resulting dataset.

## Question 2 (Separating and Cleaning Text)

**Background:** Data often arrives in messy formats with multiple pieces of information crammed into a single string or cluttered with unnecessary words.

**a) Conceptual**

Regular expressions (regex) use special characters to define patterns. Explain the difference between `.` and `\\.` in a regex pattern. Why do we need the double backslash when looking for a literal period in R?

**b) Applied**

Run the provided code chunk below to create the `support_tickets` dataset. First, use `separate_wider_delim()` to split the `issue` column into two new columns named `"status_code"` and `"message"` using `" - "` as the delimiter. Then, use `mutate()` and `str_remove_all()` to strip the words `"Error code: "` and `"Warning: "` from the `status_code` column so only the numbers remain. Print the new dataset.

```{r}
#| message: false

library(tidyverse)
support_tickets <- tibble(
  ticket_id = 101:103, 
  issue = c("Error code: 404 - Not Found", 
            "Warning: 502 - Bad Gateway", 
            "Error code: 500 - Internal Server Error")
)
```

## Question 3 (Reordering Factors)

**Background:** By default, R will plot categorical variables in alphabetical order. We can use factors to reorder these categories by their frequency or by the values of another variable to make our plots easier to read.

**a) Conceptual**

What is the functional difference between `fct_infreq()` and `fct_reorder()`? Give a brief scenario of when you would choose one over the other.

**b) Applied**

Load the `msleep` dataset from \{ggplot2\}. Calculate the average `sleep_total` for each biological `order` using `summarize()` and the `.by` argument. Be sure to name the new column `mean_sleep` and drop any missing observations. Next, use `mutate()` and `fct_reorder()` to reorder the `order` factor based on the average sleep time. Finally, pipe this into `ggplot()` to create a horizontal bar chart (`geom_col()`) showing `mean_sleep` on the x-axis and `order` on the y-axis.

## Question 4 (Modifying and Collapsing Levels)

**Background:** Real-world categorical data often contains redundant levels, typos, or categories that are too small to be analyzed individually. Modifying factor levels helps us tidy up this data.

**a) Conceptual**

When dealing with a categorical variable that contains many rare categories, why might a data scientist choose to use `fct_lump_n()` or `fct_lump_prop()` prior to plotting or statistical modeling?

**b) Applied**

Load the `gss_cat` dataset from \{forcats\}. The `rincome` (respondent income) column contains many levels, some of which represent missing or non-response data. Use `mutate()` and `fct_collapse()` to combine the levels `"No answer"`, `"Don't know"`, `"Refused"`, and `"Not applicable"` into a single new level called `"Unknown"`. Then, pipe the result into `count()` to verify the new distribution of `rincome`.

## Question 5 (Parsing and Formatting Dates)

**Background:** Dates stored as character strings are difficult to sort and plot correctly. Parsing them into Date objects allows R to understand the actual calendar values.

**a) Conceptual**

Explain the difference between `ymd()`, `mdy()`, and `dmy()`. Why is it safer to parse dates using these \{lubridate\} functions rather than keeping them as character strings?

**b) Applied**

Run the provided code chunk below to create the `checkouts` dataset, which contains European-style dates. Use `mutate()` and the correct \{lubridate\} function to convert the `due_date` column from a character string into a proper Date object. Then, use `mutate()` again with `wday()` to create a new column called `due_day` that contains the abbreviated label of the day of the week the book is due (e.g., "Mon", "Tue"). Set `label = TRUE` to ensure it returns the name instead of a number.

```{r}
checkouts <- tibble(
  book = c("Dune", "1984", "Beloved"), 
  due_date = c("15-08-2026", "02-11-2026", "25-12-2026")
)
```

## Question 6 (Calculating Time Spans)

**Background:** R provides several ways to calculate the time span between two dates. Choosing the right tool depends on whether you are tracking human calendar time or exact scientific time.

**a) Conceptual**

In \{lubridate\}, what is the fundamental difference between a "Period" (like `days()`) and a "Duration" (like `ddays()`)? Briefly describe a situation where using one instead of the other would lead to an incorrect calculation.

**b) Applied**

Run the provided code to create the `missions` dataset, logging the launch and splashdown times of historical spaceflights. First, use `mutate()` and `interval()` to calculate the exact span of time between `launch` and `splashdown` for each mission, saving it to a new column called `mission_length`. Next, create another column called `duration_days` by dividing your `mission_length` interval by `ddays(1)` to see exactly how many 24-hour duration periods the mission lasted. Print the resulting dataset.

```{r}
missions <- tibble(
  mission = c("Apollo 11", "Apollo 13", "Artemis 1"),
  launch = ymd_hms(c("1969-07-16 13:32:00", "1970-04-11 19:13:00", "2022-11-16 06:47:44")),
  splashdown = ymd_hms(c("1969-07-24 16:50:35", "1970-04-17 18:07:41", "2022-12-11 17:40:30"))
)
```

---

✅ Done!
*Please save and then render your document to an HTML file. Check it over to make sure it is complete (and rendered properly) and then upload this file to Canvas by the deadline listed on the syllabus. Note that, if any of your code chunks has an error in it, the HTML file will not render properly (this is why removing scratch work is important). If you receive an error message upon clicking "Render" about a missing file, this actually means that there is an error in one of your chunks.*
