---
lecture: "12b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} |
{{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 24))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Understand the difference between character strings and factors

- Control the order of categorical data in plots and tables

- Recode, collapse, and "lump" factor levels to simplify messy data
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Creating Factors

## What are Factors? {.fs70}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Question:** *Why not just use strings for categories?*

-   [Factors]{.b .blue} are used for categorical variables with a fixed and known set of possible values (***levels***)
-   Each level can also have a nice display ***label***
-   Unlike strings, factors allow you to control the ***order*** of categories and check for invalid values

::: {.fragment .mt1}
-   We use the {forcats} package (part of the {tidyverse}) to work with them
-   Most functions in this package start with `fct_`
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/theater.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Creating Factors

::: {.fs80}
**Question:** *How do we convert raw data into a factor?*
:::

```{r}
# Raw data as numbers (where 1=Female, 2=Male and 11 is a typo)
sex <- c(2, 1, 1, 2, 11)
```

:::{.fragment}
```{r}
# Convert to factor so R knows that only 1 and 2 were valid
fct_sex <- factor(sex, levels = c(1, 2))
fct_sex
```
:::

:::{.fragment}
```{r}
# Each level can be given a string to serve as its "label"
factor(sex, levels = c(1, 2), labels = c("Female", "Male"))
```
:::

## Factors vs. Strings

::: {.fs80}
**Question:** *What happens if we don't specify levels?*
:::

```{r}
# Raw data as strings
birth_months <- c("Jan", "Feb", "Feb", "Jan", "Apr", "Mar", "Apr")

# Default levels = observed values in alphanumeric order
factor(birth_months)
```

:::{.fragment}
```{r}
# Explicitly define levels and order (including unobserved ones)
factor(birth_months, levels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", 
                                "Jul", "Aug", "Sep", "Oct", "Nov", "Dec"))
```
:::

## A Safer Alternative

::: {.fs80}
**Question:** *Are there better tools for creating factors than base R?*

The {forcats} package provides `fct()`. It is safer than `factor()` because it errors if you include a value not in the levels, and it orders by first appearance.
:::

```{r}
#| error: true

# Base R factor() silently creates NAs for typos
factor(c("Dec", "Apr", "Jam", "Mar"), levels = c("Jan", "Feb", "Mar", "Apr", "Dec"))

# forcats::fct() catches the error for us
fct(c("Dec", "Apr", "Jam", "Mar"), levels = c("Jan", "Feb", "Mar", "Apr", "Dec"))
```

# Ordering Levels

## Introducing the GSS Dataset

::: {.fs80}
To explore reordering and modifying factors, we will use the `gss_cat` dataset. This is a sample of data from the General Social Survey.
:::

```{r}
#| message: false

library(tidyverse)
data("gss_cat", package = "forcats")
glimpse(gss_cat)
```

## Order by Appearance

::: {.fs80}
**Question:** *Can I order my levels by which appear first in the data?*
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

gss_cat |> 
  mutate(
    marital = fct_inorder(marital)
  ) |> 
  ggplot(aes(y = marital)) + 
  geom_bar()
```


## Order by Frequency

::: {.fs80}
**Question:** *How can I order my levels by their number of observations?*
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

gss_cat |> 
  mutate(
    marital = fct_infreq(marital)
  ) |> 
  ggplot(aes(y = marital)) + 
  geom_bar()
```

## Reversing Order

::: {.fs80}
**Question:** *Can I quickly reverse the order of my factors?*
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

gss_cat |> 
  mutate( 
    marital = 
      marital |> 
      fct_infreq() |> 
      fct_rev()
  ) |> 
  ggplot(aes(y = marital)) + 
  geom_bar()
```

## Creating a Summary Table

::: {.fs80}
Let's calculate the average age for each marital status group to play with.
:::

```{r}
#| message: false

gss_age <- gss_cat |> 
  drop_na(age) |> 
  summarize(
    .by = marital,
    mean_age = mean(age)
  ) |> 
  print()
```

## Order by Value

::: {.fs80}
Order by another variable in the dataset
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

# Order by average age
gss_age |> 
  mutate(
    marital = fct_reorder(
      marital,  # factor to reorder
      mean_age  # values to order by
    )
  ) |> 
  ggplot(
    aes(
      x = mean_age, 
      y = marital
    )
  ) + 
  geom_col()
```

## Manual Reordering

::: {.fs80}
Manually bring one or more variables to the front / bottom
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

gss_age |> 
  mutate(
    marital = fct_relevel(
      marital,
      "No answer", # first
      "Divorced"   # second
    )
  ) |> 
  ggplot(aes(
    x = mean_age,
    y = marital
  )) + 
  geom_col()
```

# Modifying Levels

## Why Modify Levels?

::: {.columns .pv4}

::: {.column width="60%"}
- Sometimes the data you inherit has messy categories
- You might need to combine small, specific groups into larger, more general ones for analysis
- Categorical variables with too many levels can cause issues
- Modifying levels allows us to tidy up before plotting or modeling
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/tools.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Recoding Levels

::: {.fs80}
**Question:** *How do we rename levels to make them clearer?*
:::

```{r}
#| warning: false
#| output-location: column

gss_cat |> 
  mutate(partyid = fct_recode(partyid,
    "rep, strong" = "Strong republican",
    "rep, weak"  = "Not str republican",
    "ind, rep" = "Ind,near rep",
    "ind"  = "Independent",
    "ind, dem" = "Ind,near dem",
    "dem, weak"  = "Not str democrat",
    "dem, strong" = "Strong democrat"
  )) |> 
  count(partyid) |> 
  print(n = 10)
```

## Collapsing Levels

::: {.fs80}
**Question:** *How do we group many small categories into a few large ones?*
:::

```{r}
gss_cat |> 
  mutate(party_simple = fct_collapse(partyid,
    other = c("No answer", "Don't know", "Other party"),
    rep   = c("Strong republican", "Not str republican"),
    ind   = c("Ind,near rep", "Independent", "Ind,near dem"),
    dem   = c("Not str democrat", "Strong democrat")
  )) |> 
  count(party_simple, sort = TRUE)
```

## Many Small Groups

```{r}
gss_cat |> ggplot(aes(y = relig)) + geom_bar()
```


## Lumping Small Groups

::: {.fs80}
**Question:** *What if we only care about the most common categories?*
:::

```{r}
gss_cat |> 
  mutate(
    # Keep the 5 largest religions and lump all others
    relig_lumped = fct_lump_n(relig, n = 5)
  ) |> 
  count(relig_lumped, sort = TRUE)
```

## Other Ways to Lump

::: {.fs80}
**Question:** *What if I want to lump based on proportions rather than a fixed count?*
:::

```{r}
gss_cat |> 
  mutate(
    # Keep religions that appear at least 5% of the time
    relig_prop = fct_lump_prop(relig, prop = 0.05)
  ) |> 
  count(relig_prop, sort = TRUE)
```
