---
title: "Week 12: Specialize"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [12a] Strings

### Topics
- Understand how R handles strings and special characters
- Combine, separate, and extract parts of text data
- Use regular expressions to detect and replace patterns within strings

### Readings
- R4DS (2E) Chapter 14: [Strings](https://r4ds.hadley.nz/strings.html){target="_blank"}
- R4DS (2E) Chapter 15: [Regular Expressions](https://r4ds.hadley.nz/regexps.html){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[12a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Combining and Extracting Text

Using the `starwars` dataset (included in \{dplyr\}), let's generate descriptive text for each character.

- First, use `select()` to keep just the `name`, `species`, and `homeworld` columns.
- Use `str_glue()` inside a `mutate()` step to create a new column called `description` that dynamically inserts the character's information into the string.
- Create another column called `home_abbr` that uses `str_sub()` to extract only the first three letters of the character's homeworld.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)

data("starwars", package = "dplyr")

starwars |> 
  select(name, species, homeworld) |> 
  mutate(
    description = str_glue("{name} is a {species} from {homeworld}."),
    home_abbr = str_sub(homeworld, start = 1, end = 3)
  )
```

{{< dstop >}}

#### 2) Separating Product Codes

Data often comes with multiple pieces of information stored in a single column. Let's practice splitting them apart using a simulated inventory dataset.

- First, run the code provided below to create the `inventory` dataset.
- Use `separate_wider_delim()` to split the `item` column into three separate columns: `"furniture"`, `"wood_type"`, and `"price"`.
- Set the `delim` argument to the hyphen character.

```{r}
inventory <- tibble(
  item = c("Chair-Oak-199", "Table-Pine-299", "Desk-Birch-399")
)
```

{{< dstart summary="Answer key" >}}

```{r}
inventory |> 
  separate_wider_delim(
    cols = item,
    delim = "-",
    names = c("furniture", "wood_type", "price")
  )
```

{{< dstop >}}

#### 3) Regular Expressions

Let's use regular expressions (regex) to find specific patterns in text and clean them up using building blocks like anchors and character classes.

- Start with the `starwars` dataset and use `filter()` with `str_detect()` to find all characters whose name starts with either "A" or "S". *Hint:* Use an anchor and a character class.
- Use `mutate()` and `str_remove_all()` to create a new column called `no_spaces` that removes all whitespace from the names.
- Select only the `name` and `no_spaces` columns to verify your work.

{{< dstart summary="Answer key" >}}

```{r}
starwars |> 
  filter(str_detect(name, "^[AS]")) |> 
  mutate(
    no_spaces = str_remove_all(name, "\\s")
  ) |> 
  select(name, no_spaces)
```

{{< dstop >}}

## [12b] Factors

### Topics
- Understand the difference between character strings and factors
- Control the order of categorical data in plots and tables
- Recode, collapse, and "lump" factor levels to simplify messy data

### Readings
- R4DS (2E) Chapter 16: [Factors](https://r4ds.hadley.nz/factors.html){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[12b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Ordering Factors by Value

Using the `mpg` dataset (included in \{ggplot2\}), let's create a bar chart that is ordered meaningfully instead of alphabetically.

- First, calculate the average engine size (`displ`) for each `manufacturer` using `summarize()` and the `.by` argument. Name the new column `mean_displ`.
- Use `mutate()` and `fct_reorder()` to reorder the `manufacturer` factor based on the calculated `mean_displ`.
- Pipe the result into `ggplot()` and create a horizontal bar chart (`geom_col()`) with `mean_displ` on the x-axis and `manufacturer` on the y-axis.
- Look at the plot to answer: which manufacturers make the largest and smallest engines on average?

{{< dstart summary="Answer key" >}}

```{r}
mpg |> 
  summarize(
    .by = manufacturer, 
    mean_displ = mean(displ)
  ) |> 
  mutate(
    manufacturer = fct_reorder(manufacturer, mean_displ)
  ) |> 
  ggplot(aes(x = mean_displ, y = manufacturer)) + 
  geom_col()
```

{{< dstop >}}

#### 2) Lumping Rare Categories

When a categorical variable has too many levels, it can be helpful to group the smallest ones together.

- Use the `starwars` dataset and drop any rows with a missing `species` value.
- Use `mutate()` and `fct_lump_n()` to create a new column called `species_lumped`.
- Keep only the top 3 most common species and lump all others into an "Other" category.
- Pipe the result into `count()` to see the tallies for your new column.

{{< dstart summary="Answer key" >}}

```{r}
starwars |> 
  drop_na(species) |> 
  mutate(species_lumped = fct_lump_n(species, n = 3)) |> 
  count(species_lumped, sort = TRUE)
```

{{< dstop >}}

#### 3) Collapsing Factor Levels

Sometimes we want to manually group specific categories together for a broader analysis

- Using the `diamonds` dataset, create a new column called `cut_simplified` using `fct_collapse()`.
- Collapse "Ideal" and "Premium" into a single level called "Excellent".
- Collapse "Very Good", "Good", and "Fair" into a single level called "Standard".
- Run `count()` on your new `cut_simplified` column to see the results.

{{< dstart summary="Answer key" >}}

```{r}
diamonds |> 
  mutate(
    cut_simplified = fct_collapse(cut,
      "Excellent" = c("Ideal", "Premium"),
      "Standard"  = c("Very Good", "Good", "Fair")
    )
  ) |> 
  count(cut_simplified)
```

{{< dstop >}}

## [12c] Dates & Times

### Topics
- Parse strings and build dates from individual components
- Extract, round, and modify date-time components
- Understand periods, durations, and intervals for time spans

### Readings
- R4DS (2E) Chapter 17: [Dates and times](https://r4ds.hadley.nz/datetimes.html){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[12c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Parsing Dates from Strings

Let's convert human-readable text into proper machine-readable dates.

- Run the code below to load a small dataset of historical events.
- Use `mutate()` and the appropriate \{lubridate\} parsing function to convert the `date_str` column into a proper Date object. Name the new column `actual_date`.

```{r}
events <- tibble(
  event = c("Apollo 11 Landing", "First Flight", "Y2K Bug"),
  date_str = c("July 20, 1969", "December 17, 1903", "January 1, 2000")
)
```

{{< dstart summary="Answer key" >}}

```{r}
events |> 
  mutate(
    actual_date = mdy(date_str)
  )
```

{{< dstop >}}

#### 2) Extracting Components

Using the `economics` dataset (including with \{ggplot2\}), let's figure out which month historically has the highest average unemployment.

- Use `mutate()` and `month()` to extract the month from the `date` column.
- Set `label = TRUE` so you get the actual month label (e.g., "Jan") instead of numbers.
- Next, `summarize()` the average of the `unemploy` column grouped by your new month variable using `.by`.
- Finally, `arrange()` the results in descending order to see the top months.

{{< dstart summary="Answer key" >}}

```{r}
economics |> 
  mutate(obs_month = month(date, label = TRUE)) |> 
  summarize(
    .by = obs_month,
    avg_unemploy = mean(unemploy)
  ) |> 
  arrange(desc(avg_unemploy))
```

{{< dstop >}}

#### 3) Calculating Time Spans

Let's calculate exact lifespans using intervals and durations.

- Run the code below to create a small dataset of early US Presidents.
- Use `mutate()` and the `interval()` function to calculate the exact span of time between their `born` and `died` dates. Name this column `life_span`.
- Create a final column called `age_years` by dividing the `life_span` interval by `years(1)` to see their age at death.

```{r}
presidents <- tibble(
  name = c("George Washington", "Abraham Lincoln"),
  born = ymd(c("1732-02-22", "1809-02-12")),
  died = ymd(c("1799-12-14", "1865-04-15"))
)
```

{{< dstart summary="Answer key" >}}

```{r}
presidents |> 
  mutate(
    life_span = interval(start = born, end = died),
    age_years = life_span / years(1)
  )
```

{{< dstop >}}

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../11/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 11</a>

  <a href="../13/index.html" class="btn btn-primary" role="button">Week 13&nbsp;&nbsp;&raquo;</a>
</div>
```
