---
lecture: "05b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| message: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(patchwork)
set_theme(theme_gray(base_size = 30))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Combine layers of overlapping geometric objects

- Learn aesthetic setting, mapping, and grouping
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Setup

```{r}
#| message: false

library(tidyverse)
data("mpg", package = "ggplot2")

glimpse(mpg)
```

# Layering

## Layering {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   ggplot2 uses a [layered]{.b .blue} grammar of graphics
    -   We can keep [stacking geoms]{.b .green} on top

::: {.fragment .mt1}
-   Layering adds a lot of possibilities
    -   We can convey [more complex]{.b .green} ideas
    -   We can learn more about our data
:::

::: {.fragment .mt1}
-   But we can still describe these graphics
    -   Just [describe each layer]{.b .green} in turn
    -   And describe the layers' ordering
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/layers.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## The Smooth Geom

```{r}
#| layout-ncol: 2
#| message: false

# Left: Point geom plots all the observations
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()

# Right: Smooth geom plots a local line-of-best-fit
ggplot(mpg, aes(x = displ, y = hwy)) + geom_smooth()
```

## Overlapping the Layers

```{r}
#| message: false
#| warning: false
#| fig-height: 6
#| fig-width: 18

# Adding additional geoms stacks their layers on top
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point() + 
  geom_smooth()
```

## Economics Dataset

```{r}
data("economics", package = "ggplot2")

glimpse(economics)
```


## Line Geoms for Time Series

```{r}
#| layout-ncol: 2
#| message: false

# Left
ggplot(economics, aes(x = date, y = unemploy)) + geom_point()

# Right
ggplot(economics, aes(x = date, y = unemploy)) + geom_line()
```

## Adding Reference Lines

```{r}
#| fig-height: 6
#| fig-width: 18

# Add a horizontal reference line at y=10000
ggplot(economics, aes(x = date, y = unemploy)) + 
  geom_hline(yintercept = 10000) +
  geom_line()
```


# Aesthetics

## Aesthetics {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Aesthetics]{.b .blue} control a geom's appearance
    -   points: *color, shape, size, ...*
    -   smooths: *color, linetype, linewidth, ...*

::: {.fragment .mt1}
-   There are two ways to control aesthetics:
    -   You can [set it]{.b .blue} to a [value]{.b .green}
        -   It will take on *one* value only
        -   e.g., all points = red
    -   You can [map it]{.b .blue} to a [variable]{.b .green}
        -   It will take on *different* values
        -   e.g., sedans = red, trucks = blue

:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/watercolor.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Setting Color

```{r}
#| fig-height: 6
#| fig-width: 18

# Set color inside geom_*() to apply it to all points
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(color = "blue")
```

## Mapping Color

```{r}
#| fig-height: 6
#| fig-width: 18

# Map color to a variable within aes() to have it vary
ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point()
```

## Pitfall: Setting in aes

```{r}
#| fig-height: 6
#| fig-width: 18

# Don't try to set an aesthetic inside aes()
ggplot(mpg, aes(x = displ, y = hwy, color = "blue")) + 
  geom_point()
```

## Pitfall: Mapping in geom

```{r}
#| error: true
#| fig-height: 6
#| fig-width: 18

# Don't try to map an aesthetic outside of aes()
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = class)
```


## Setting Shape

```{r}
#| fig-height: 6
#| fig-width: 18

# Set the shape inside geom_*() to "triangle"
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(shape = "triangle")
```

## Mapping Shape

```{r}
#| fig-height: 6
#| fig-width: 18

# Map shape to the drv variable inside aes()
ggplot(mpg, aes(x = displ, y = hwy, shape = drv)) + 
  geom_point()
```

## Redundant Mapping

```{r}
#| fig-height: 6
#| fig-width: 18

# Map both shape and color to the drv variable
ggplot(mpg, aes(x = displ, y = hwy, shape = drv, color = drv)) + 
  geom_point()
```

## Other Useful Aesthetics {.smaller}

::: {.columns .pv4}
::: {.column width="50%"}
-   [size]{.b .blue}
    -   **Geoms:** `point`, `text`
    -   **Action:** Controls point or font size

-   [alpha]{.b .blue}
    -   **Geoms:** All
    -   **Action:** Controls transparency<br>(0 = invisible, 1 = opaque)
    
:::
::: {.column width="50%"}
-   [linewidth]{.b .blue}
    -   **Geoms:** `line`, `smooth`
    -   **Action:** Controls line thickness

-   [linetype]{.b .blue}
    -   **Geoms:** `line`, `smooth`
    -   **Action:** Controls line pattern<br>(solid, dashed, dotted)

:::
:::

## Aesthetics in Action

```{r}
#| output-location: column
#| message: false
#| fig-height: 7
#| fig-width: 7

ggplot(
  mpg, 
  aes(x = displ, y = hwy)
) +
  geom_point(
    size = 4,
    color = "coral",
    shape = "square",
    alpha = 0.5
  ) +
  geom_smooth(
    color = "steelblue",
    linewidth = 1.5,
    linetype = "dashed"
  )
```

## Aesthetics in Action

```{r}
#| fig-height: 6
#| fig-width: 18

ggplot(economics, aes(x = date, y = unemploy)) + 
  geom_hline(yintercept = 10000, color = "coral", 
             linetype = "dashed", linewidth = 2) +
  geom_line(linewidth = 1)
```

# Grouping

## Implicit vs. Explicit Grouping {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   **Grouping** controls how functions (like `geom_smooth` or `geom_line`) analyze your data.
    -   Do we calculate *one* trend or *many*?

::: {.fragment .mt1}
-   **Implicit Grouping** (via `color`, `shape`, etc.)
    -   Mapping a variable to appearance *automatically* groups the data.
    -   **Result:** Separate geoms + a Legend.
:::

::: {.fragment .mt1}
-   **Explicit Grouping** (via `group`)
    -   Mapping a variable to `group` groups the data *without* changing appearance.
    -   **Result:** Separate geoms + No Legend.
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/hub.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Smooth (No Grouping)

```{r}
#| message: false
#| fig-height: 6
#| fig-width: 18

# No color mapping leads to a smooth for all datapoints
ggplot(mpg, aes(x = displ, y = hwy)) + geom_smooth()
```

## Smooth (Implicit Grouping)

```{r}
#| message: false
#| fig-height: 6
#| fig-width: 18

# Mapping color to drv leads to a separate smooth for each group
ggplot(mpg, aes(x = displ, y = hwy, color = drv)) + geom_smooth()
```

## ChickWeight Dataset

```{r}
data("ChickWeight", package = "datasets")

glimpse(ChickWeight)
```

## Line Series (No Grouping)

```{r}
#| message: false
#| fig-height: 6
#| fig-width: 18

# Without grouping, we get an incomprehensible line for all Chicks
ggplot(ChickWeight, aes(x = Time, y = weight)) + geom_line()
```

## Line Series (Explicit Grouping)

```{r}
#| message: false
#| fig-height: 6
#| fig-width: 18

# But with explicit grouping, we get a separate line per Chick
ggplot(ChickWeight, aes(x = Time, y = weight, group = Chick)) + 
  geom_line(alpha = 0.3)
```
