---
title: "Week 05: Visualize"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [05a] Graphics / Scatterplots

### Topics
- Learn the principles that govern data visualization
- Create and save scatterplots using `ggplot()` and `ggsave()`

### Readings
- R4DS (2E) Section 1.1: [Introductions](https://r4ds.hadley.nz/data-visualize.html#introduction){target="_blank"}
- R4DS (2E) Section 1.2: [First Steps](https://r4ds.hadley.nz/data-visualize.html#first-steps){target="_blank"}
- R4DS (2E) Section 1.3: [ggplot2 calls](https://r4ds.hadley.nz/data-visualize.html#sec-ggplot2-calls){target="_blank"}
- R4DS (2E) Section 1.6: [Saving your plots](https://r4ds.hadley.nz/data-visualize.html#sec-ggsave){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[05a] Slideshow">
</iframe>
```

::: {.fs80}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Scatterplots

Create a new scatterplot using the `mpg` dataset (from the \{ggplot2\} package).
 
* Map **Highway MPG (`hwy`)** to the x-axis.
* Map **City MPG (`cty`)** to the y-axis.
* Ensure the points are displayed as geometric objects.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)
data("mpg", package = "ggplot2")
p <- 
  ggplot(
    data = mpg,
    mapping = aes(x = hwy, y = cty)
  ) +
  geom_point()
p
```

{{< dstop >}}

#### 2) Saving Graphics

Using `ggsave()`, perform the following steps:

a.  Save the plot from question 1 as a **PNG** file named "plot_mpg.png" with a width of 8 inches and a height of 6 inches.
b.  Save the plot from question 1 as a **PDF** file named "plot_mpg.pdf" with the same dimensions.
c.  Open both files on your computer. Zoom in closely on the points in both files and describe the difference in quality between the raster (PNG) and vector (PDF) versions.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
#| eval: false

ggsave(
  filename = "plot_mpg.png",
  plot = p,
  width = 8,
  height = 6,
  units = "in"
)
```

**Part (b)**

```{r}
#| eval: false

ggsave(
  filename = "plot_mpg.pdf",
  plot = p,
  width = 8,
  height = 6,
  units = "in"
)
```

{{< dstop >}}

## [05b] Layers / Aesthetics

### Topics
- Combine layers of overlapping geometric objects
- Learn aesthetic setting, mapping, and grouping

### Readings
- R4DS (2E) Section 1.2.4: [Adding aesthetics and layers](https://r4ds.hadley.nz/data-visualize.html#sec-adding-aesthetics-layers){target="_blank"}
- R4DS (2E) Section 9.2: [Aesthetic mappings](https://r4ds.hadley.nz/layers.html#aesthetic-mappings){target="_blank"}
- R4DS (2E) Section 9.3: [Geometric objects](https://r4ds.hadley.nz/layers.html#sec-geometric-objects){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[05b] Slideshow">
</iframe>
```

::: {.fs80}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Layers and Settings

a.  **Layering Geoms:** The `pressure` dataset (from the "datasets" package) contains data on the relationship between temperature and vapor pressure of mercury. Create a plot to visualize this.
    * Map **Temperature (`temperature`)** to the x-axis and **Pressure (`pressure`)** to the y-axis.
    * Add a layer of points using `geom_point()`.
    * Add a layer of lines using `geom_line()`.
b.  **Setting Aesthetics:** Modify the plot from Part (a) to change the static appearance.
    * Keep the points layer, but set the size of the points to 3.
    * Modify the line layer: set the color to "blue" and the linetype to "dashed".

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
#| message: false

library(tidyverse)
data("pressure", package = "datasets")

pressure |>
  ggplot(aes(x = temperature, y = pressure)) +
  geom_point() +
  geom_line()
```

**Part (b)**

```{r}
pressure |>
  ggplot(aes(x = temperature, y = pressure)) +
  geom_point(size = 3) +
  geom_line(color = "blue", linetype = "dashed")
```

{{< dstop >}}

#### 2) Mappings and Grouping

a.  **Mapping Aesthetics:** Create a scatterplot using the `mpg` dataset (from the "ggplot2" package) to see if fuel efficiency data is clustered by cylinder count.
    * Map **City MPG (`cty`)** to the x-axis and **Highway MPG (`hwy`)** to the y-axis.
    * Map **Cylinders (`cyl`)** to the **color** aesthetic. (Note: You may want to wrap `cyl` in `factor()` so R treats it as categories rather than a continuous number).
    * Add a layer of points and set their `size` to 3.
b.  **Explicit Grouping:** The `Orange` dataset (from the "datasets" package) tracks the growth of 5 specific orange trees. Create a line plot to visualize this growth.
    * Map **Age (`age`)** to the x-axis and **Circumference (`circumference`)** to the y-axis.
    * Map **Tree (`Tree`)** to the `group` aesthetic.
    * Add a layer of lines using `geom_line()`.
    * *Bonus:* What happens if you remove the `group` aesthetic?

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
data("mpg", package = "ggplot2")

mpg |>
  ggplot(aes(x = cty, y = hwy, color = factor(cyl))) +
  geom_point(size = 3)
```

**Part (b)**

```{r}
data("Orange", package = "datasets")

Orange |>
  ggplot(aes(x = age, y = circumference, group = Tree)) +
  geom_line()
```

{{< dstop >}}


## [05c] Distributions / Relationships

### Topics
- Visualize the distribution of a single variable
- Visualize relationships between several variables

### Readings
- R4DS (2E) Section 1.4: [Visualizing distributions](https://r4ds.hadley.nz/data-visualize.html#visualizing-distributions){target="_blank"}
- R4DS (2E) Section 1.5: [Visualizing relationships](https://r4ds.hadley.nz/data-visualize.html#visualizing-relationships){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[05c] Slideshow">
</iframe>
```

::: {.fs80}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Distributions

a) Create a histogram using the `mpg` dataset (from the \{ggplot2\} package) to visualize the distribution of City MPG.
    + Map **City MPG (`cty`)** to the x-axis.
    + Ensure the bars are displayed as geometric objects using `geom_histogram()`.
    + Set the `binwidth` argument to 2.
b) Create a bar chart using the `mpg` dataset to visualize the distribution of Vehicle Class.
    + Map **Vehicle Class (`class`)** to the x-axis.
    + Ensure the bars are displayed as geometric objects using `geom_bar()`.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
#| message: false

library(tidyverse)
data("mpg", package = "ggplot2")

mpg |>
  ggplot(aes(x = cty)) +
  geom_histogram(binwidth = 2)
```

**Part (b)**

```{r}
mpg |>
  ggplot(aes(x = class)) +
  geom_bar()
```

{{< dstop >}}

#### 2) Relationships

a) Create a boxplot using the `mpg` dataset to visualize the relationship between vehicle class and Highway MPG.
    + Map **Vehicle Class (`class`)** to the x-axis.
    + Map **Highway MPG (`hwy`)** to the y-axis.
    + Ensure the distribution summaries are displayed as geometric objects using `geom_boxplot()`.

b) Create a stacked bar chart using the `mpg` dataset to visualize the relationship between vehicle class and drive train.
    + Map **Vehicle Class (`class`)** to the x-axis.
    + Map **Drive Train (`drv`)** to the fill aesthetic.
    + Ensure the bars are displayed as geometric objects using `geom_bar()`.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
mpg |>
  ggplot(aes(x = class, y = hwy)) +
  geom_boxplot()
```

**Part (b)**

```{r}
mpg |>
  ggplot(aes(x = class, fill = drv)) +
  geom_bar()
```

{{< dstop >}}

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../04/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 04</a>

  <a href="../06/index.html" class="btn btn-primary" role="button">Week 06&nbsp;&nbsp;&raquo;</a>
</div>
```
