---
title: "Week 05: Visualize"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

:::{.callout-caution title="Work in Progress"}
This page is still under development.
:::

## [05a] Graphics / Scatterplots

### Topics
- Learn the principles that govern data visualization
- Create and save scatterplots using `ggplot()` and `ggsave()`

### Readings
- R4DS (2E) Section 1.1: [Introductions](https://r4ds.hadley.nz/data-visualize.html#introduction){target="_blank"}
- R4DS (2E) Section 1.2: [First Steps](https://r4ds.hadley.nz/data-visualize.html#first-steps){target="_blank"}
- R4DS (2E) Section 1.3: [ggplot2 calls](https://r4ds.hadley.nz/data-visualize.html#sec-ggplot2-calls){target="_blank"}
- R4DS (2E) Section 1.6: [Saving your plots](https://r4ds.hadley.nz/data-visualize.html#sec-ggsave){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[05a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and then press 'F' on your keyboard to view the slides full screen.
:::

### Practice

#### 1) Scatterplots

Create a new scatterplot using the `mpg` dataset (from the \{ggplot2\} package).
 
* Map **Highway MPG (`hwy`)** to the x-axis.
* Map **City MPG (`cty`)** to the y-axis.
* Ensure the points are displayed as geometric objects.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)
data("mpg", package = "ggplot2")
p <- 
  ggplot(
    data = mpg,
    mapping = aes(x = hwy, y = cty)
  ) +
  geom_point()
p
```

{{< dstop >}}

## 2) Saving Graphics

Using `ggsave()`, perform the following steps:

a.  Save the plot from question 1 as a **PNG** file named "plot_mpg.png" with a width of 8 inches and a height of 6 inches.
b.  Save the plot from question 1 as a **PDF** file named "plot_mpg.pdf" with the same dimensions.
c.  Open both files on your computer. Zoom in closely on the points in both files and describe the difference in quality between the raster (PNG) and vector (PDF) versions.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
#| eval: false

ggsave(
  filename = "plot_mpg.png",
  plot = p,
  width = 8,
  height = 6,
  units = "in"
)
```

**Part (b)**

```{r}
#| eval: false

ggsave(
  filename = "plot_mpg.pdf",
  plot = p,
  width = 8,
  height = 6,
  units = "in"
)
```

{{< dstop >}}

## [05b] Distributions / Relationships

### Topics
- Visualize the distribution of a single variable
- Visualize relationships between several variables

### Readings
- R4DS (2E) Section 1.4: [Visualizing distributions](https://r4ds.hadley.nz/data-transform.html#introduction){target="_blank"}
- R4DS (2E) Section 1.5: [Visualizing relationships](https://r4ds.hadley.nz/data-visualize.html#visualizing-distributions){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[05b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and then press 'F' on your keyboard to view the slides full screen.
:::

### Practice



## [05c] Aesthetics / Geometrics

### Topics
- Learn when to "set" versus "map" an aesthetic value
- Combine layers of geometrics objects (points and smooths)

### Readings
- R4DS (2E) Section 9.2: [Aesthetic mappings](https://r4ds.hadley.nz/layers.html#aesthetic-mappings){target="_blank"}
- R4DS (2E) Section 9.3: [Geometric objects](https://r4ds.hadley.nz/layers.html#sec-geometric-objects){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[05c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and then press 'F' on your keyboard to view the slides full screen.
:::

### Practice


<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../04/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 04</a>

  <a href="../06/index.html" class="btn btn-primary" role="button">Week 06&nbsp;&nbsp;&raquo;</a>
</div>
```
