---
title: "Week 05: Visualize"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

:::{.callout-caution title="Work in Progress"}
This page is still under development.
:::

## [05a] Graphics / Scatterplots

### Topics
- Learn the principles that govern data visualization
- Create and save scatterplots using `ggplot()` and `ggsave()`

### Readings
- R4DS (2E) Section 1.1: [Introductions](https://r4ds.hadley.nz/data-visualize.html#introduction){target="_blank"}
- R4DS (2E) Section 1.2: [First Steps](https://r4ds.hadley.nz/data-visualize.html#first-steps){target="_blank"}
- R4DS (2E) Section 1.3: [ggplot2 calls](https://r4ds.hadley.nz/data-visualize.html#sec-ggplot2-calls){target="_blank"}
- R4DS (2E) Section 1.6: [Saving your plots](https://r4ds.hadley.nz/data-visualize.html#sec-ggsave){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[05a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Scatterplots

Create a new scatterplot using the `mpg` dataset (from the \{ggplot2\} package).
 
* Map **Highway MPG (`hwy`)** to the x-axis.
* Map **City MPG (`cty`)** to the y-axis.
* Ensure the points are displayed as geometric objects.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)
data("mpg", package = "ggplot2")
p <- 
  ggplot(
    data = mpg,
    mapping = aes(x = hwy, y = cty)
  ) +
  geom_point()
p
```

{{< dstop >}}

#### 2) Saving Graphics

Using `ggsave()`, perform the following steps:

a.  Save the plot from question 1 as a **PNG** file named "plot_mpg.png" with a width of 8 inches and a height of 6 inches.
b.  Save the plot from question 1 as a **PDF** file named "plot_mpg.pdf" with the same dimensions.
c.  Open both files on your computer. Zoom in closely on the points in both files and describe the difference in quality between the raster (PNG) and vector (PDF) versions.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
#| eval: false

ggsave(
  filename = "plot_mpg.png",
  plot = p,
  width = 8,
  height = 6,
  units = "in"
)
```

**Part (b)**

```{r}
#| eval: false

ggsave(
  filename = "plot_mpg.pdf",
  plot = p,
  width = 8,
  height = 6,
  units = "in"
)
```

{{< dstop >}}

## [05b] Aesthetics / Layers

### Topics
- Learn when to "set" versus "map" an aesthetic value
- Combine layers of geometric objects (points and smooths)

### Readings
- R4DS (2E) Section 1.2.4: [Adding aesthetics and layers](https://r4ds.hadley.nz/data-visualize.html#sec-adding-aesthetics-layers){target="_blank"}
- R4DS (2E) Section 9.2: [Aesthetic mappings](https://r4ds.hadley.nz/layers.html#aesthetic-mappings){target="_blank"}
- R4DS (2E) Section 9.3: [Geometric objects](https://r4ds.hadley.nz/layers.html#sec-geometric-objects){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[05b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice



## [05c] Distributions / Relationships

### Topics
- Visualize the distribution of a single variable
- Visualize relationships between several variables

### Readings
- R4DS (2E) Section 1.4: [Visualizing distributions](https://r4ds.hadley.nz/data-visualize.html#visualizing-distributions){target="_blank"}
- R4DS (2E) Section 1.5: [Visualizing relationships](https://r4ds.hadley.nz/data-visualize.html#visualizing-relationships){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[05c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Distributions

a) Create a histogram using the `mpg` dataset (from the \{ggplot2\} package) to visualize the distribution of City MPG.
    + Map **City MPG (`cty`)** to the x-axis.
    + Ensure the bars are displayed as geometric objects using `geom_histogram()`.
    + Set the `binwidth` argument to 2.
b) Create a bar chart using the `mpg` dataset to visualize the distribution of Vehicle Class.
    + Map **Vehicle Class (`class`)** to the x-axis.
    + Ensure the bars are displayed as geometric objects using `geom_bar()`.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
#| message: false

library(tidyverse)
data("mpg", package = "ggplot2")

mpg |>
  ggplot(aes(x = cty)) +
  geom_histogram(binwidth = 2)
```

**Part (b)**

```{r}
mpg |>
  ggplot(aes(x = class)) +
  geom_bar()
```

{{< dstop >}}

#### 2) Relationships

a) Create a boxplot using the `mpg` dataset to visualize the relationship between vehicle class and Highway MPG.
    + Map **Vehicle Class (`class`)** to the x-axis.
    + Map **Highway MPG (`hwy`)** to the y-axis.
    + Ensure the distribution summaries are displayed as geometric objects using `geom_boxplot()`.

b) Create a stacked bar chart using the `mpg` dataset to visualize the relationship between vehicle class and drive train.
    + Map **Vehicle Class (`class`)** to the x-axis.
    + Map **Drive Train (`drv`)** to the fill aesthetic.
    + Ensure the bars are displayed as geometric objects using `geom_bar()`.

{{< dstart summary="Answer key" >}}

**Part (a)**

```{r}
mpg |>
  ggplot(aes(x = class, y = hwy)) +
  geom_boxplot()
```

**Part (b)**

```{r}
mpg |>
  ggplot(aes(x = class, fill = drv)) +
  geom_bar()
```

{{< dstop >}}

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../04/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 04</a>

  <a href="../06/index.html" class="btn btn-primary" role="button">Week 06&nbsp;&nbsp;&raquo;</a>
</div>
```
