---
lecture: "13a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(patchwork)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 20))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Create multiple subplots using `facet_wrap()` and `facet_grid()`

- Use position adjustments to stack, fill, and dodge bar charts

- Address overplotting in scatterplots using jittering and hex-binning
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Faceting

## Faceting {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Faceting** creates multiple subplots
    -   Each subplot shows a different subset of the data and is excellent for exploring conditional relationships

::: {.fragment .mt1}
-   There are two main types of faceting:
    -   **Wrap:** Creates a ribbon of plots
    -   **Grid:** Creates a matrix of plots
    
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/grid.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Setup for Faceting

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

data("penguins", package = "datasets")
# Drop rows with NAs
penguins2 <- drop_na(penguins) 

# Create a base plot to add to
p <- 
  ggplot(
    penguins2, 
    aes(x = body_mass, y = flipper_len)
  ) + 
  geom_point()
p
```


## Facet Wrap

```{r}
#| fig-width: 12
#| fig-height: 7

# Create a subplot for each species
p + facet_wrap(facets = vars(species))
```

## Layout Options

```{r}
#| fig-width: 12
#| fig-height: 7

# Control the number of columns and rows to wrap across
p + facet_wrap(facets = vars(species), ncol = 2, nrow = 2)
```

## Multiple Variables

```{r}
#| fig-width: 12
#| fig-height: 7

# Wrap across combinations of species and sex
p + facet_wrap(facets = vars(species, sex))
```

## Freeing Scales

```{r}
#| fig-width: 12
#| fig-height: 7

# While wrapping, give each *subplot* its own scales
p + facet_wrap(facets = vars(species, sex), scales = "free")
```

## Facet Grid

```{r}
#| fig-width: 12
#| fig-height: 7

# Create a sex-by-species matrix (or grid) of subplots
p + facet_grid(rows = vars(sex), cols = vars(species))
```

## Facet Labellers

```{r}
#| fig-width: 12
#| fig-height: 7

# Label each facet "strip" using both the variable name and value
p + facet_grid(rows = vars(sex), cols = vars(species), labeller = label_both)
```

## Freeing Scales

```{r}
#| fig-width: 12
#| fig-height: 7

# While gridding, give each *row* and each *column* its own scales
p + facet_grid(rows = vars(sex), cols = vars(species), scales = "free")
```

# Positions

## Position Adjustments {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   Geoms often overlap with many groups
-   **Position adjustments** dictate how ggplot2 should handles this overlap
    -   We can stack items on top of one another
    -   We can place them side-by-side
    -   We can stretch them to a common size

::: {.fragment .mt1}
-   This is useful for bar and column charts
-   It is critical for aligning text labels
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/stacked.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## The Stacking Default

```{r}
#| fig-width: 12
#| fig-height: 7

# The default for geom_bar is `position = "stack"`
ggplot(penguins2, aes(x = species, fill = island)) + geom_bar(color = "black")
```

## Proportional Bars

```{r}
#| fig-width: 12
#| fig-height: 7

# We can use `position = "fill"` to turn them into proportions
ggplot(penguins, aes(x = species, fill = island)) +  
  geom_bar(color = "black", position = "fill") + labs(y = "proportion")
```

## Dodging Bars

```{r}
#| fig-width: 12
#| fig-height: 7

# We can use `position = "dodge"` to put them side-by-side
ggplot(penguins, aes(x = species, fill = island)) +
  geom_bar(color = "black", position = "dodge")
```

## Plotting from Summaries

```{r}
# To explicitly plot the zeros, we can use a two-step process
penguins_sum <- 
  penguins2 |> 
  count(species, island, .drop = FALSE) |> 
  print()
```

## Dodging Columns

```{r}
#| fig-width: 12
#| fig-height: 7

# Now we map the y-axis to `n` and use geom_col()
ggplot(penguins_sum, aes(x = species, y = n, fill = island)) +
  geom_col(color = "black", position = "dodge")
```


## Overplotting

```{r}
#| fig-width: 12
#| fig-height: 7

# Multiple overlapping geoms can lead to "overplotting"
data("mpg", package = "ggplot2")
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point(alpha = 0.5)
```

## Jittering Points

```{r}
#| fig-width: 12
#| fig-height: 7

# Jittering is pushing each point slightly in a random direction
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(alpha = 0.5, position = "jitter")
```

## Jitter Magnitude

```{r}
#| fig-width: 12
#| fig-height: 7

# We can push farther but pushing too far may become misleading
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(alpha = 0.5, position = position_jitter(width = .5, height = .5))
```

## Extreme Overplotting

```{r}
#| fig-width: 12
#| fig-height: 7

# With lots and lots of points, jittering is unlikely to help
data("diamonds", package = "ggplot2")
ggplot(diamonds, aes(x = carat, y = price)) + geom_point(alpha = 0.5)
```


## Binning into Hexes

```{r}
#| fig-width: 12
#| fig-height: 7

# In such cases, we can count points within hexagonal bins
library(hexbin)
ggplot(diamonds, aes(x = carat, y = price)) + geom_hex()
```


## Discrete Overplotting

```{r}
#| fig-width: 12
#| fig-height: 7

# With discrete x, we want to jitter horizontally only
ggplot(mpg, aes(x = drv, y = hwy)) + geom_point()
```

## Horizontal Jitter

```{r}
#| fig-width: 12
#| fig-height: 7

# Set height to 0 to only push left or right
ggplot(mpg, aes(x = drv, y = hwy)) + 
  geom_point(position = position_jitter(width = 0.25, height = 0))
```

## Discrete Overplotting

```{r}
#| fig-width: 12
#| fig-height: 7

## Beeswarm plots can also handle discrete overplotting well
library(ggbeeswarm)
ggplot(mpg, aes(x = drv, y = hwy)) + geom_beeswarm()
```
