---
lecture: "13b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(maps)
library(patchwork)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 20))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Extract and draw polygon map data in ggplot2

- Add points and text labels to existing maps

- Join data with map data and make complex maps
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Drawing Maps

## Where Do Maps Come From? {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   To draw a map, we need a dataset of geographic coordinates
    -   These coordinates define the boundary lines of countries, states, or counties
-   The `map_data()` function from the \{maps\} package provides easy access
    -   It returns a data frame of longitude and latitude points
-   We will then plot the region shapes using `geom_polygon()`
    
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/globe.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Extracting State Data

```{r}
library(tidyverse)
library(maps)

# Extract the coordinates for the kansas map
kansas <- map_data("county", "kansas") |>
  as_tibble() |> print(n = 5)
```

## Understanding Map Data

- **long** and **lat**: The x and y coordinates of the boundary lines
- **group**: A unique identifier for each polygon (e.g., a specific island or continuous landmass)
- **order**: The sequence in which the points should be connected to draw the shape correctly
- **region**: The name of the geographic area (e.g., state)
- **subregion:** The name of the area portion (e.g., county)

## Drawing Polygons

```{r}
#| fig-width: 12
#| fig-height: 5

ggplot(kansas, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black")
```

## Fixing the Distortion

```{r}
#| fig-width: 12
#| fig-height: 5

ggplot(kansas, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "white", color = "black") + coord_quickmap()
```

# Wrangling Map Data

## Pulling All States

```{r}
# Extract the coordinates for all US states
state_map <- map_data("state") |> 
  as_tibble() |> print(n = 10)
```

## Drawing the US Map

```{r}
#| fig-width: 12
#| fig-height: 5

ggplot(state_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "steelblue", color = "black", linewidth = 0.5) +
  coord_quickmap()
```

## Filtering Regions

```{r}
#| fig-width: 12
#| fig-height: 5

# We can use standard dplyr tools to map specific regions
midwest_map <- state_map |> 
  filter(region %in% c("kansas", "missouri", "iowa", "nebraska"))

ggplot(midwest_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "wheat", color = "black") +
  coord_quickmap()
```

# Points and Labels

## Overlaying Data {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   Maps are most useful when they provide context for other data
-   Because maps in ggplot2 are just standard geoms, we can easily layer other geoms on top of them
    -   `geom_point()` is great for cities, events, or locations
    -   `geom_text()` or `geom_label()` can identify specific places
    
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/translate.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## A City Dataset

```{r}
# Let's create a small dataset of cities
cities <- tribble(
  ~city, ~long, ~lat,
  "Lawrence", -95.2353, 38.9717,
  "St. Louis", -90.1994, 38.6270,
  "Des Moines", -93.6091, 41.5868,
  "Omaha", -95.9345, 41.2565
)
cities
```

## Separating Aesthetics

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

p_midwest <- 
  ggplot() +
  geom_polygon(
    data = midwest_map, 
    mapping = aes(
      x = long, 
      y = lat, 
      group = group
    ),
    fill = "white", 
    color = "black"
  ) +
  coord_quickmap()
p_midwest
```


## Adding Points

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

p_midwest +
  geom_point(
    data = cities,
    aes(x = long, y = lat),
    color = "red", 
    size = 4
  )
```

## Adding Labels

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

p_midwest +
  geom_point(
    data = cities,
    aes(x = long, y = lat),
    color = "red", 
    size = 3
  ) +
  geom_text(
    data = cities,
    aes(
      x = long, 
      y = lat, 
      label = city
    ),
    nudge_y = 0.3, # shift up
    fontface = "bold"
  )
```


# Choropleth Maps

## What is a Choropleth? {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   A **choropleth map** uses color or shading to represent statistical data across geographic regions
    -   For example, coloring states based on their population density or average temperature
-   To create one, we must merge our geographic polygon data with our statistical data using a join

:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/crayons.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Preparing Statistical Data

```{r}
#| message: false
#| replace_path: ["../../data/", ""]

arrests <- read_csv("../../data/arrests.csv")
arrests
```

## Joining the Data

```{r}
# We use left_join() to attach the statistics to the map boundaries
state_joined <- 
  state_map |> 
  left_join(arrests, by = "region")

state_joined
```

## Creating the Choropleth

```{r}
#| fig-width: 12
#| fig-height: 5

# We map the fill aesthetic to one of our statistical variables
ggplot(state_joined, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = Murder), color = "white", linewidth = 0.3) +
  coord_quickmap()
```

## Polishing the Map

```{r}
#| output-location: slide
#| fig-width: 12
#| fig-height: 5

ggplot(
  state_joined, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    aes(fill = Murder), 
    color = "white",
    linewidth = 0.3
  ) +
  coord_quickmap() +
  scale_fill_viridis_c(option = "magma") + # use better colors
  theme_void() + # remove the axes and gridlines
  labs(
    title = "US Murder Arrest Rates",
    fill = "Rate per 100k"
  )
```

## Summary

-   **Coordinates:** Alter how space is mapped using `coord_fixed()`, `coord_polar()`, and `coord_quickmap()`
-   **Polygons:** Build basic maps using `map_data()` and `geom_polygon()`, ensuring `group` is mapped properly
-   **Layers:** Add context to maps by layering `geom_point()` and `geom_text()` from separate data sources
-   **Choropleths:** Color geographic regions by joining statistical data to spatial boundary data
