---
lecture: "13b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(maps)
library(patchwork)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 20))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Understand how ggplot2 handles coordinate systems

- Extract and draw polygon map data in ggplot2

- Add points, text, and colors to existing maps
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Clipping vs. Zooming

## Original

```{r}
#| fig-width: 12
#| fig-height: 7

# Original plot with full data
p_orig <- ggplot(mpg, aes(x = displ, y = hwy)) + geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "red", linewidth = 2)
p_orig
```

## Clipping

```{r}
#| fig-width: 12
#| fig-height: 7

# Using lims drops points outside this range and then RE-FITS the smooth
p_orig + lims(x = c(5.0, 6.5), y = c(10, 30))
```

## Zooming

```{r}
#| fig-width: 12
#| fig-height: 7

# Using coord limits zooms in WITHOUT re-fitting the smooth
p_orig + coord_cartesian(xlim = c(5.0, 6.5), ylim = c(10, 30))
```

# Aspect Ratios

## Freed Aspect Ratio

```{r}
#| fig-width: 12
#| fig-height: 7

# Generate the path of a perfect circle
t <- seq(0, 2*pi, length.out = 100)
circle <- tibble(x = sin(t), y = cos(t))

# By default, a change of 1 on x and y can be different distances
ggplot(circle, aes(x, y)) + geom_path()
```

## Fixed Aspect Ratio

```{r}
#| fig-width: 12
#| fig-height: 7

# An aspect ratio of 1 ensures these changes have the same distance
ggplot(circle, aes(x, y)) + geom_path() + coord_fixed(ratio = 1)
```

# Polar Coordinates

## Filled Bar Charts

```{r}
#| fig-width: 12
#| fig-height: 7

ggplot(penguins, aes(y = 1, fill = sex)) + 
  geom_bar(position = "fill") + labs(y = "proportion")
```

## Filled Pie Charts

```{r}
#| fig-width: 12
#| fig-height: 5

ggplot(penguins, aes(x = 1, fill = sex)) + 
  geom_bar(position = "fill") +   
  coord_polar(theta = "y") + 
  theme_void(base_size = 24)
```
# Drawing Maps

## Where Do Maps Come From? {.fsmaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   To draw a map, we need a dataset of geographic coordinates
    -   These coordinates define the boundary lines of countries, states, or counties
-   The `map_data()` function from the `maps` package provides easy access to these datasets
    -   It returns a data frame of longitude and latitude points
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/globe.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Extracting World Data

```{r}
# Extract the coordinates for the world map
world_map <- map_data("world")
world_map
```

## Understanding Map Data

- **long** and **lat**: The x and y coordinates of the boundary lines
- **group**: A unique identifier for each polygon (e.g., a specific island or continuous landmass)
- **order**: The sequence in which the points should be connected to draw the shape correctly
- **region**: The name of the geographic area (e.g., country name)

## Drawing Polygons

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# We use geom_polygon() to draw shapes
# The group aesthetic is critical so R knows which points belong together
ggplot(
  world_map, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    fill = "lightgray", 
    color = "white",
    linewidth = 0.2
  )
```

## Fixing the Distortion

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# The previous map looked stretched!
# coord_quickmap() sets the aspect ratio correctly for maps based on latitude
ggplot(
  world_map, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    fill = "lightgray", 
    color = "white",
    linewidth = 0.2
  ) +
  coord_quickmap()
```

# Regional Maps

## Extracting State Data

```{r}
# Extract the coordinates for US states
state_map <- map_data("state")
state_map
```

## Drawing the US Map

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

ggplot(
  state_map, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    fill = "steelblue", 
    color = "black",
    linewidth = 0.5
  ) +
  coord_quickmap()
```

## Filtering Regions

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# We can use standard dplyr tools to map specific regions
midwest_map <- 
  state_map |> 
  filter(
    region %in% c("kansas", "missouri", "iowa", "nebraska")
  )

ggplot(
  midwest_map, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    fill = "palegreen", 
    color = "black"
  ) +
  coord_quickmap()
```

# Points and Labels

## Overlaying Data {.fsmaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   Maps are most useful when they provide context for other data
-   Because maps in ggplot2 are just standard geoms, we can easily layer other geoms on top of them
    -   `geom_point()` is great for cities, events, or locations
    -   `geom_text()` or `geom_label()` can identify specific places
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/stacked.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## A City Dataset

```{r}
# Let's create a small dataset of cities
cities <- tribble(
  ~city, ~long, ~lat,
  "Lawrence", -95.2353, 38.9717,
  "St. Louis", -90.1994, 38.6270,
  "Des Moines", -93.6091, 41.5868,
  "Omaha", -95.9345, 41.2565
)
cities
```

## Preventing Aesthetic Inheritance

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# Best practice: keep the global ggplot() empty and map aesthetics within each geom.
# This prevents geom_point from looking for a 'group' variable in the cities data!
p_midwest <- 
  ggplot() +
  geom_polygon(
    data = midwest_map, 
    aes(x = long, y = lat, group = group),
    fill = "white", color = "black"
  ) +
  coord_quickmap()

p_midwest
```


## Adding Points

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# Now we overlay the points using our cities data
p_midwest +
  geom_point(
    data = cities,
    aes(x = long, y = lat),
    color = "red", 
    size = 4
  )
```

## Adding Labels

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# We can nudge the text slightly so it doesn't overlap the dot
p_midwest +
  geom_point(
    data = cities,
    aes(x = long, y = lat),
    color = "red", size = 3
  ) +
  geom_text(
    data = cities,
    aes(x = long, y = lat, label = city),
    nudge_y = 0.3,
    fontface = "bold"
  )
```


# Choropleth Maps

## What is a Choropleth? {.fsmaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   A **choropleth map** uses color or shading to represent statistical data across geographic regions
    -   For example, coloring states based on their population density or average temperature
-   To create one, we must merge our geographic polygon data with our statistical data using a join
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/grid.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Preparing Statistical Data

```{r}
# We will use the built-in USArrests dataset
# The map_data region names are lowercase, so our data needs to match!
arrests <- 
  USArrests |> 
  rownames_to_column("region") |> 
  mutate(region = tolower(region))

arrests
```

## Joining the Data

```{r}
# We use left_join() to attach the statistics to the map boundaries
state_joined <- 
  state_map |> 
  left_join(arrests, by = "region")

state_joined
```

## Creating the Choropleth

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# We map the fill aesthetic to one of our statistical variables
ggplot(
  state_joined, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    aes(fill = Murder), 
    color = "white",
    linewidth = 0.3
  ) +
  coord_quickmap()
```

## Polishing the Map

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# We can use theme_void() to remove the axes and gridlines
# and scale_fill_viridis_c() for an accessible color palette
ggplot(
  state_joined, 
  aes(x = long, y = lat, group = group)
) +
  geom_polygon(
    aes(fill = Murder), 
    color = "white",
    linewidth = 0.3
  ) +
  coord_quickmap() +
  scale_fill_viridis_c(option = "magma") +
  theme_void() +
  labs(
    title = "US Murder Arrest Rates",
    fill = "Rate per 100k"
  )
```

## Summary

-   **Coordinates:** Alter how space is mapped using `coord_fixed()`, `coord_polar()`, and `coord_quickmap()`
-   **Polygons:** Build basic maps using `map_data()` and `geom_polygon()`, ensuring `group` is mapped properly
-   **Layers:** Add context to maps by layering `geom_point()` and `geom_text()` from separate data sources
-   **Choropleths:** Color geographic regions by joining statistical data to spatial boundary data
