---
lecture: "13c"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(patchwork)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 24))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Apply complete themes to rapidly change the look of plots

- Customize individual theme elements for fine-grained control

- Combine multiple plots into complex layouts using patchwork
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Themes

## Themes {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- Themes let you control [non-data elements]{.b .blue}
  - Font size and color
  - Ticks and gridlines
  - Legend position

::: {.fragment .mt1}
- Complete themes change many elements
  - `theme_*()`
:::

::: {.fragment .mt1}
- Individual elements can also be changed
  - `theme(...)`
  - `element_*()`
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/pantone.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Setup for Themes

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

library(tidyverse)
# Create a base plot to add themes to
p <- 
  ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter(
    aes(color = drv, shape = drv),
    size = 3, alpha = 0.8
  ) + 
  labs(
    x = "Engine displacement (L)",
    y = "Highway MPG",
    color = "Drive",
    shape = "Drive",
    title = "Fuel Efficiency"
  )

p
```

## Classic Theme

```{r}
#| fig-width: 12
#| fig-height: 5

# theme_classic() removes gridlines and gives a clean axis
p + theme_classic(base_size = 24)
```

## Black and White Theme

```{r}
#| fig-width: 12
#| fig-height: 5

# theme_bw() adds a border and keeps major gridlines
p + theme_bw(base_size = 24)
```

## Minimal Theme

```{r}
#| fig-width: 12
#| fig-height: 5

# theme_minimal() removes the background box for a modern look
p + theme_minimal(base_size = 24)
```

## Legend Position

```{r}
#| fig-width: 12
#| fig-height: 5

# Move the legend to the bottom, top, left, or "none"
p + theme(legend.position = "bottom")
```

## Legend Coordinates

```{r}
#| fig-width: 12
#| fig-height: 5

# Move the legend inside the plot using relative coordinates (0 to 1)
p + theme(legend.position = c(0.85, 0.8))
```

## Text Elements

```{r}
#| fig-width: 12
#| fig-height: 5

# Use element_text() to change font properties
p + theme(plot.title = element_text(face = "bold", color = "red"))
```

## Line Elements

```{r}
#| fig-width: 12
#| fig-height: 5

# Use element_line() to modify gridlines and axes
p + theme(panel.grid.major = element_line(linetype = "dashed", color = "red"))
```

## Blanking

```{r}
#| fig-width: 12
#| fig-height: 5

# Use element_blank() to remove specific features entirely
p + theme(panel.grid.minor = element_blank(), axis.ticks.x = element_blank())
```

## Reusable Themes

```{r}
#| output-location: column
#| fig-width: 9
#| fig-height: 7

# Store your custom theme in an object
# Then apply it to multiple plots
my_theme <- 
  theme_minimal(base_size = 30) +
  theme(
    legend.position = "bottom",
    plot.title = 
      element_text(face = "bold"),
    panel.grid.minor = element_blank()
  )

p + my_theme
```

# Layouts

## Layouts {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- We may want to combine multiple plots into one
  - e.g., for communication and publishing
  - e.g., to combine related "subfigures"
  - e.g., to create complex figure layouts

::: {.fragment .mt1}
- To do so, we can use the `{patchwork}` package
  - This redefines the `+`, `|`, `/` operators
  - We use these to combine ggplot objects
  - We can control layout and annotations
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/quilt.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Setup for Patchwork

```{r}
#| echo: true

# Create example plots to work with
p1 <- ggplot(mpg, aes(x = displ)) + 
  geom_histogram(bins = 10, fill = "steelblue", color = "white") + 
  labs(x = "Engine Displacement", y = NULL)

p2 <- ggplot(mpg, aes(x = hwy)) + 
  geom_histogram(bins = 10, fill = "seagreen", color = "white") + 
  labs(x = "Highway MPG", y = NULL)

p3 <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(alpha = 0.5) +
  labs(x = "Engine Displacement", y = "Highway MPG")
```

## Combining Plots: Basic Addition

```{r}
#| fig-width: 12
#| fig-height: 6

# The + operator attempts to lay out the plots logically
library(patchwork)

p1 + p2
```

## Controlling the Layout

```{r}
#| fig-width: 12
#| fig-height: 6

# Use plot_layout() to explicitly state columns or rows
p1 + p2 + p3 + plot_layout(ncol = 3)
```

## Adding Spacers

```{r}
#| fig-width: 12
#| fig-height: 6

# Use plot_spacer() to leave empty areas in the grid
p1 + plot_spacer() + p2 + plot_layout(ncol = 3)
```

## Side-by-Side

```{r}
#| fig-width: 12
#| fig-height: 6

# The | operator explicitly places plots side-by-side
p1 | p2 | p3
```

## Above-and-Below

```{r}
#| fig-width: 12
#| fig-height: 7

# The / operator explicitly places plots above and below
p1 / p3
```

## Mixing and Matching

```{r}
#| fig-width: 12
#| fig-height: 7

# Parentheses determine nesting and grouping
(p1 | p2) / p3
```

## Relative Sizes

```{r}
#| fig-width: 12
#| fig-height: 7

# Pass a vector to widths or heights to adjust proportions
(p1 | p2) / p3 + plot_layout(heights = c(1, 2))
```

## Overall Annotations

```{r}
#| fig-width: 12
#| fig-height: 7

# plot_annotation() adds a master title, subtitle, or caption
(p1 | p2) / p3 + 
  plot_annotation(
    title = "Vehicle Engine Size and Fuel Economy",
    caption = "Data from fueleconomy.gov"
  )
```

## Automated Tagging

```{r}
#| fig-width: 12
#| fig-height: 7

# tag_levels automatically labels subfigures (e.g., A, B, C or 1, 2, 3)
(p1 | p2) / p3 + 
  plot_annotation(tag_levels = "A")
```
