---
title: "Week 13: Visualize"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [13a] Positions

### Topics
- Create multiple subplots using `facet_wrap()` and `facet_grid()`
- Use position adjustments to stack, fill, and dodge bar charts
- Address overplotting in scatterplots using jittering and hex-binning

### Readings
- R4DS (2E) Section 9.4: [Facets](https://r4ds.hadley.nz/layers.html#facets){target="_blank"}
- R4DS (2E) Section 9.6: [Position Adjustments](https://r4ds.hadley.nz/layers.html#position-adjustments){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[13a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice
#### 1) Faceting Sleep Patterns

Using the `msleep` dataset (from \{ggplot2\}), let's explore how different types of "vores" (carnivores, herbivores, etc.) sleep.

- Create a scatterplot with `bodywt` on the x-axis and `sleep_total` on the y-axis.
- Use `facet_wrap()` to create subplots based on the `vore` column.
- Because mammals have very different weight ranges, set `scales = "free_x"` within the facet function so each subplot has its own x-axis scale.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)

data("msleep", package = "ggplot2")

# We drop NAs in vore to keep the facets clean
msleep |> 
  drop_na(vore) |> 
  ggplot(aes(x = bodywt, y = sleep_total)) +
  geom_point() +
  facet_wrap(facets = vars(vore), scales = "free_x")
```

{{< dstop >}}

#### 2) Comparing Metro Areas Across States

Let's use the `midwest` dataset (from \{ggplot2\}) to see the distribution of counties that are considered "metropolitan" across different states.

- Create a bar chart (`geom_bar`) with `state` on the x-axis.
- Map the `fill` aesthetic to `inmetro` (using `factor()` to treat it as categorical).
- Use `position = "fill"` to show the **proportion** (rather than the count) of metro vs. non-metro counties in each state.
- Update the y label to "proportion" and the fill label to "Is Metro?"

{{< dstart summary="Answer key" >}}

```{r}
data("midwest", package = "ggplot2")

ggplot(midwest, aes(x = state, fill = factor(inmetro))) +
  geom_bar(position = "fill") +
  labs(y = "proportion", fill = "Is Metro?")
```

{{< dstop >}}

#### 3) Handling Overlapping Geyser Data

The `faithful` dataset (from \{datasets\}) contains many overlapping points. Let's practice two different ways to make the data density clearer.

- First, create a standard scatterplot of `eruptions` (x) vs `waiting` (y).
- Create a second version of the plot that uses jittering and alpha-transparency to reveal hidden points.
- Create a third version using `geom_hex()` to bin the points into count-based shapes. *Note: You will need the {hexbin} library installed.*

{{< dstart summary="Answer key" >}}

```{r}
#| layout-nth: 1
#| fig-width: 10

library(hexbin)

data("faithful", package = "datasets")

# Version 1: Jittering
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_point(position = "jitter", alpha = 0.5)

# Version 2: Hex Binning
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_hex()
```

{{< dstop >}}

## [13b] Maps

### Topics
- Extract and draw polygon map data in ggplot2
- Add points and text labels to existing maps
- Join data with map data and make complex maps

### Readings
- EGDA (3E) Chapter 6: [Maps](https://ggplot2-book.org/maps.html){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[13b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Mapping the World

Let's practice drawing a map of the entire world using the \{maps\} package.

- Extract the coordinates for the world map using `map_data("world")` and convert it to a tibble.
- Create a ggplot mapping the `long` (x), `lat` (y), and `group` aesthetics.
- Add the map polygons using `geom_polygon()`, setting the fill to "lightgreen" and the outline color to "black".
- Apply `coord_quickmap()` to fix the geographic distortion.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
library(maps)

# Extract the world map data
world_map <- map_data("world")

# Draw the map
ggplot(world_map, aes(x = long, y = lat, group = group)) +
  geom_polygon(fill = "lightgreen", color = "black") +
  coord_quickmap()
```

{{< dstop >}}

#### 2) Pinpointing Global Cities

Now let's add some points to our world map to represent major global cities. 

- Run the provided code to create a small dataset of three cities.
- Start with your world map code from the previous question, but move the mapping aesthetics into the `geom_polygon()` layer so they do not conflict with the points.
- Add a `geom_point()` layer using the new `cities` dataset.
- Map `long` to x and `lat` to y for the points, and set their color to "red" with a size of 3.

```{r}
cities <- tribble(
  ~city, ~long, ~lat,
  "Tokyo", 139.6917, 35.6895,
  "London", -0.1276, 51.5072,
  "New York", -74.0060, 40.7128
)
```

{{< dstart summary="Answer key" >}}

```{r}
ggplot() +
  geom_polygon(
    data = world_map, 
    mapping = aes(x = long, y = lat, group = group),
    fill = "lightgreen", 
    color = "black"
  ) +
  geom_point(
    data = cities,
    mapping = aes(x = long, y = lat),
    color = "red", 
    size = 3
  ) +
  coord_quickmap()
```

{{< dstop >}}

#### 3) Creating a Global Choropleth

Let's visualize some simulated statistical data across different countries to create a choropleth map.

- Run the provided code to load a small dataset of arbitrary values for five countries.
- Join the `world_map` dataset with `country_data` using a `left_join()`. Match them by the "region" column.
- Create a map using `geom_polygon()` where the `fill` aesthetic is mapped to the new `value` column.
- Set the polygon outline color to "black" with a linewidth of 0.3.
- Keep `coord_quickmap()` to maintain the correct proportions.

```{r}
country_data <- tribble(
  ~region, ~value,
  "Canada", 10,
  "Brazil", 25,
  "Australia", 40,
  "China", 80,
  "Algeria", 15
)
```

{{< dstart summary="Answer key" >}}

```{r}
# Join the data
world_joined <- world_map |> 
  left_join(country_data, by = "region")

# Draw the choropleth
ggplot(world_joined, aes(x = long, y = lat, group = group)) +
  geom_polygon(aes(fill = value), color = "black", linewidth = 0.3) +
  coord_quickmap()
```

{{< dstop >}}

## [13c] Themes

### Topics
- Apply complete themes to rapidly change the look of plots
- Customize individual theme elements for fine-grained control
- Combine multiple plots into complex layouts using patchwork

### Readings
- R4DS (2E) Section 11.5: [Themes](https://r4ds.hadley.nz/communication.html#sec-themes){target="_blank"}
- R4DS (2E) Section 11.6: [Layout](https://r4ds.hadley.nz/communication.html#layout){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[13c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Applying Complete Themes and Moving Legends

Let's practice applying a built-in theme and adjusting the legend position for a cleaner look.

- Run the provided code to create a base plot `p_base` using the `mpg` dataset.
- Apply the `theme_minimal()` complete theme to `p_base`.
- Use the `theme()` function to move the legend to the "bottom" of the plot.

```{r}
library(tidyverse)

p_base <- ggplot(mpg, aes(x = displ, y = hwy, color = class)) +
  geom_point(size = 3) +
  labs(title = "Engine Size vs. Highway MPG")
```

{{< dstart summary="Answer key" >}}

```{r}
p_base +
  theme_minimal() +
  theme(legend.position = "bottom")
```

{{< dstop >}}

#### 2) Customizing Theme Elements

Complete themes are great, but sometimes you need fine-grained control over specific components like text formatting and gridlines.

- Start with the `p_base` plot from the previous question.
- Use `theme()` and `element_text()` to make the plot title bold (`face = "bold"`).
- Use `element_blank()` to completely remove the minor gridlines (`panel.grid.minor`).
- Use `element_line()` to make the major gridlines (`panel.grid.major`) dashed.

{{< dstart summary="Answer key" >}}

```{r}
p_base +
  theme(
    plot.title = element_text(face = "bold"),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_line(linetype = "dashed")
  )
```

{{< dstop >}}

#### 3) Combining Plots with Patchwork

When presenting data, you often need to combine multiple related subfigures into one cohesive layout. 

- Run the provided code to create three simple plots (`p1`, `p2`, `p3`) and load the patchwork library.
- Use patchwork operators to place `p1` and `p2` side-by-side on the top row, and place `p3` below them. Use parentheses to group the top row.
- Add an overall `plot_annotation()` with the title "Vehicle Exploration" and automatically tag the subfigures using `tag_levels = "A"`.

```{r}
library(patchwork)

p1 <- ggplot(mpg, aes(x = class)) + 
  geom_bar(fill = "steelblue")

p2 <- ggplot(mpg, aes(x = drv)) + 
  geom_bar(fill = "seagreen")

p3 <- ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point(alpha = 0.5)
```

{{< dstart summary="Answer key" >}}

```{r}
#| fig-width: 12
#| fig-height: 7

(p1 | p2) / p3 +
  plot_annotation(
    title = "Vehicle Exploration",
    tag_levels = "A"
  )
```

{{< dstop >}}

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../12/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 12</a>

  <a href="../14/index.html" class="btn btn-primary" role="button">Week 14&nbsp;&nbsp;&raquo;</a>
</div>
```
