---
title: "Week 13: Visualize"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [13a] Positions

### Topics
- Create multiple subplots using `facet_wrap()` and `facet_grid()`
- Use position adjustments to stack, fill, and dodge bar charts
- Address overplotting in scatterplots using jittering and hex-binning

### Readings
- R4DS (2E) Section 9.4: [Facets](https://r4ds.hadley.nz/layers.html#facets){target="_blank"}
- R4DS (2E) Section 9.6: [Position Adjustments](https://r4ds.hadley.nz/layers.html#position-adjustments){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[13a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice
#### 1) Faceting Sleep Patterns

Using the `msleep` dataset (from \{ggplot2\}), let's explore how different types of "vores" (carnivores, herbivores, etc.) sleep.

- Create a scatterplot with `bodywt` on the x-axis and `sleep_total` on the y-axis.
- Use `facet_wrap()` to create subplots based on the `vore` column.
- Because mammals have very different weight ranges, set `scales = "free_x"` within the facet function so each subplot has its own x-axis scale.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)

data("msleep", package = "ggplot2")

# We drop NAs in vore to keep the facets clean
msleep |> 
  drop_na(vore) |> 
  ggplot(aes(x = bodywt, y = sleep_total)) +
  geom_point() +
  facet_wrap(facets = vars(vore), scales = "free_x")
```

{{< dstop >}}

#### 2) Comparing Metro Areas Across States

Let's use the `midwest` dataset (from \{ggplot2\}) to see the distribution of counties that are considered "metropolitan" across different states.

- Create a bar chart (`geom_bar`) with `state` on the x-axis.
- Map the `fill` aesthetic to `inmetro` (using `factor()` to treat it as categorical).
- Use `position = "fill"` to show the **proportion** (rather than the count) of metro vs. non-metro counties in each state.
- Update the y label to "proportion" and the fill label to "Is Metro?"

{{< dstart summary="Answer key" >}}

```{r}
data("midwest", package = "ggplot2")

ggplot(midwest, aes(x = state, fill = factor(inmetro))) +
  geom_bar(position = "fill") +
  labs(y = "proportion", fill = "Is Metro?")
```

{{< dstop >}}

#### 3) Handling Overlapping Geyser Data

The `faithful` dataset (from \{datasets\}) contains many overlapping points. Let's practice two different ways to make the data density clearer.

- First, create a standard scatterplot of `eruptions` (x) vs `waiting` (y).
- Create a second version of the plot that uses jittering to reveal hidden points.
- Create a third version using `geom_hex()` to bin the points into count-based shapes. *Note: You will need the {hexbin} library installed.*

{{< dstart summary="Answer key" >}}

```{r}
#| layout-nth: 1
#| fig-width: 10

library(hexbin)

data("faithful", package = "datasets")

# Version 1: Jittering
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_point(position = "jitter", alpha = 0.5) +
  labs(title = "Jittered Points")

# Version 2: Hex Binning
ggplot(faithful, aes(x = eruptions, y = waiting)) +
  geom_hex() +
  labs(title = "Hexagonal Binning")
```

{{< dstop >}}

## [13b] Coordinates

### Topics
- A
- B
- C

### Readings
- R4DS (2E) Section 9.5: [Statistical Transformations](https://r4ds.hadley.nz/layers.html#statistical-transformations){target="_blank"}
- R4DS (2E) Section 9.7: [Coordinate Systems](https://r4ds.hadley.nz/layers.html#coordinate-systems){target="_blank"}
- EGDA (3E) Chapter 6: [Maps](https://ggplot2-book.org/maps.html){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[13b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice
TBA

## [13c] Themes

### Topics
- A
- B
- C

### Readings
- R4DS (2E) Section 11.5: [Themes](https://r4ds.hadley.nz/communication.html#sec-themes){target="_blank"}
- R4DS (2E) Section 11.6: [Layout](https://r4ds.hadley.nz/communication.html#layout){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[13c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice
TBA

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../12/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 12</a>

  <a href="../14/index.html" class="btn btn-primary" role="button">Week 14&nbsp;&nbsp;&raquo;</a>
</div>
```
