---
title: "Lecture [03a] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1

Which of the following commands do you think will create an error in R? Why?

a) `score@T1 <- 3.2`
b) `score_at_T1 <- 3.2`
c) `score at T1 <- 3.2`
d) `1_score <- 3.2`
e) `ScoreAtTime1 <- 3.2`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R. *Bonus:* Can you come up with a better name for this variable?

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

| Code | Result (reason) |
|------|-----------------|
|`score@T1 <- 3.2`|🛑 Error (`@` not allowed in name)|
|`score_at_T1 <- 3.2`|✅ No Error|
|`score at T1 <- 3.2`|🛑 Error (spaces not allowed in name)|
|`1_score <- 3.2`|🛑 Error (name can't start with a number)|
|`ScoreAtTime1 <- 3.2`|✅ No Error|

</blockquote></details></p>

## Question 2

a) Use the `cos()` function to calculate the cosine of 3 and assign it to a variable called `y`. Print the value of `y` to the console. 
b) Use a function to round `y` to the nearest whole number.
c) Use a function to round `y` to two decimal places.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| collapse: true

# Part (a)
y <- cos(3)
y

# Part (b)
round(y)

# Part (c)
round(y, digits = 2)
```

</blockquote></details></p>

## Question 3

What do you expect the result to be for the following commands?

a) `sqrt(4 + 5)`
b) `sqrt(4) + 5`

Make your best guesses just by looking at the commands and then check your guesses by running the commands in R.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| collapse: true

# Part (a): Add then square root, same as sqrt(9)
sqrt(4 + 5)

# Part (b): Square root then add, same as 2 + 5
sqrt(4) + 5 
```

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
