---
title: "Lecture [06a] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1 (Code Style)

Apply the tidyverse style guide principles to clean up the following code:

```{r}
#| eval: false

library(tidyverse)
df <- mpg |> filter(year==2008) |> select(manufacturer , model, class, displ, hwy ) |> 
  mutate (manufacturer=str_to_title(manufacturer), model = str_to_title ( model ),class = str_to_title(class))
ggplot(data = df, mapping = aes(x = hwy, y = class, fill = class)) + 
geom_vline(xintercept=25, color = "grey40", linewidth = 1) +
geom_violin(color = "black",linewidth = 2 / 3,draw_quantiles = c(0.25,0.50, 0.75),scale = "count"
  )
```


<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

*If you didn't enjoy this question, imagine what grading in this course is like and let that motivate you to style your code nicely when submitting assignments in the future!*

```{r}
#| echo: true
#| message: false
#| fig-width: 5
#| fig-height: 5

library(tidyverse)
mpg2008 <- 
  mpg |> 
  filter(year == 2008) |> 
  select(manufacturer, model, class, displ, hwy) |> 
  mutate(
    manufacturer = str_to_title(manufacturer),
    model = str_to_title(model),
    class = str_to_title(class)
  )
ggplot(mpg2008, aes(x = hwy, y = class, fill = class)) + 
  geom_vline(xintercept = 25, color = "grey40", linewidth = 1) +
  geom_violin(
    color = "black",
    linewidth = 2 / 3,
    draw_quantiles = c(0.25, 0.50, 0.75),
    scale = "count"
  )
```

</blockquote></details></p>

## Question 2 (Chunk Options)

a) Adjust your code for Q1 to hide the welcome message from the \{tidyverse\} package.

b) Adjust your code for Q1 to make the plot 5 inches tall and 5 inches wide.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Part (a)**

Add `#| message: false` to the top of the chunk

**Part (b)**

Add `#| fig-height: 5` and `#| fig-width: 5` to the top of the chunk

</blockquote></details></p>

## Question 3 (More Quarto)

a) Embed the following online video in your document:<br />
<https://www.youtube.com/embed/hW7DW9NIO9M>

b) Use MathJax to add an equation (in display mode) to show Einstein's famous [mass-energy equivalence equation](https://en.wikipedia.org/wiki/Mass%E2%80%93energy_equivalence): $E=mc^2$

c) Change your document to use the `solar` HTML theme.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

**Part (a)**

`{{{< video https://www.youtube.com/embed/hW7DW9NIO9M >}}}`

{{< video https://www.youtube.com/embed/hW7DW9NIO9M >}}

**Part (b)**

`$$E = m c^{2}$$`

$$E = m c^{2}$$

**Part (c)**

```yaml
format:
  html:
    theme: solar
```

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
