---
title: "Lecture [06b] Activity"
---

![](../../img/programmer_2780E3.svg){.hero}

## Question 1 (Scales)

a) Use the `penguins` dataset from the \{palmerpenguins\} package to create a plot that maps `species` to the x-axis and `body_mass_g` to the y-axis. Add a violin geom and a jitter geom (with 0.25 width). For the jitter geom only, map `sex` to color.

b) Configure the y axis to range from 2500 to 6500.

c) Configure the x-axis to use the scientific names of each species (see table below)

d) *Bonus:* Suppress all messages and warnings from this chunk.

| Common Name | Scientific Name  |
|-------------|------------------|
| Adelie      | *P. adeliae*     |
| Chinstrap   | *P. antarcticus* |
| Gentoo      | *P. papua*       |

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false
#| warning: false

library(tidyverse)
library(palmerpenguins)

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_violin() + 
  geom_jitter(width = 0.25, aes(color = sex)) +
  scale_y_continuous(limits = c(2500, 6500)) +
  scale_x_discrete(
    labels = c(
      "Adelie" = "P. adeliae",
      "Chinstrap" = "P. antarcticus",
      "Gentoo" = "P. papua"
    )
  )
```

</blockquote></details></p>


## Question 2 (Labels)

Modify your final plot from Question 1 in the following ways:

a) Give the x-axis, y-axis, and the color scale better labels (with Title Case)

b) Give the plot a descriptive title about what the graphic is about (generally) and a subtitle that describes which sex-species combination had the highest body mass.

c) Add a caption stating the source of this data (e.g., the package name)

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false
#| warning: false

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_violin() + 
  geom_jitter(width = 0.25, aes(color = sex)) +
  scale_y_continuous(limits = c(2500, 6500)) +
  scale_x_discrete(
    labels = c(
      "Adelie" = "P. adeliae",
      "Chinstrap" = "P. antarcticus",
      "Gentoo" = "P. papua"
    )
  ) +
  labs(
    x = "Penguin Species",
    y = "Body Mass (g)",
    color = "Sex",
    title = "Body Mass by Species and Sex of Penguins at Palmer Station",
    subtitle = "The penguins with the highest body mass were males of the P. papua species",
    caption = "Data from the {palmerpenguins} R package"
  )
```


</blockquote></details></p>

## Question 3 (Themes)

Modify your final plot from Question 2 in the following ways:

a) Apply the "classic" complete theme using `theme_classic()` and change the base font size to 10.

b) Adjust the `legend.background` element to be rectangle with `color = "black"`.

c) Adjust the `axis.text.x` element to be text with `face = "italic"`.

<p><details>
<summary>Click here for the answer key</summary>
<blockquote>

```{r}
#| message: false
#| warning: false

ggplot(penguins, aes(x = species, y = body_mass_g)) +
  geom_violin() + 
  geom_jitter(width = 0.25, aes(color = sex)) +
  scale_y_continuous(limits = c(2500, 6500)) +
  scale_x_discrete(
    labels = c(
      "Adelie" = "P. adeliae",
      "Chinstrap" = "P. antarcticus",
      "Gentoo" = "P. papua"
    )
  ) +
  labs(
    x = "Penguin Species",
    y = "Body Mass (g)",
    color = "Sex",
    title = "Body Mass by Species and Sex of Penguins at Palmer Station",
    subtitle = "The penguins with the highest body mass were males of the P. papua species",
    caption = "Data from the {palmerpenguins} R package"
  ) +
  theme_classic(base_size = 10) +
  theme(
    axis.text.x = element_text(face = "italic"),
    legend.background = element_rect(color = "black")
  )
```

</blockquote></details></p>

---

*Note that there is no need to turn in Activities. These are just for practice!*
