---
lecture: "14a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 20))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Evaluate comparisons and combine multiple conditions using Boolean logic

- Summarize logical vectors to count occurrences and calculate proportions

- Control data transformations conditionally using `if_else()` and `case_when()`
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/padlock.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Comparisons &<br> Boolean Logic

## Logical Vectors {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   Before we can build conditional statements, we need to understand **logicals**
-   Logicals have three possible values:
    -   `TRUE`
    -   `FALSE`
    -   `NA` (missing)
-   We rarely type these by hand. Instead, we generate them using comparisons.

:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/check.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Relational Operators {.fs80}

Numerical only comparisons:

-   `x > 1` (is x greater than 1?)
-   `x >= 1` (is x greater than or equal to 1?)
-   `x < 1` (is x less than 1?)
-   `x <= 1` (is x less than or equal to 1?)

Numerical or character comparisons:

-   `x == 1` (is x equal to 1?)
-   `x != 1` (is x not equal to 1?)
-   `x %in% c(1, 2, 3)` (is x equal to 1, 2, or 3?)

## Comparisons in Action

```{r}
#| echo: true
library(tidyverse)
data("mpg", package = "ggplot2")

# Create a logical vector asking: is highway mpg greater than 30?
mpg |> 
  mutate(good_mileage = hwy > 30) |> 
  select(manufacturer, model, year, trans, hwy, good_mileage)
```

## Boolean Algebra {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   We can test multiple conditions at once
-   **Boolean algebra** allows us to combine multiple logical vectors
    -   `&` (AND):<br>TRUE only if **both** conditions are TRUE
    -   `|` (OR):<br>TRUE if **at least one** condition is TRUE
    -   `!` (NOT):<br>Inverts the logical (TRUE $\leftrightarrow$ FALSE)

:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/venn.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Boolean Algebra in Action

```{r}
#| echo: true
# Find cars that have great highway mileage AND are compact
mpg |> 
  filter(hwy > 30 & class == "compact") |> 
  select(manufacturer, model, year, trans, hwy, class)
```

# Logical Summaries

## Summarizing Logicals {.fs80}

-   Logical vectors are essentially numbers under the hood
    -   `TRUE` becomes `1` and `FALSE` becomes `0`
-   This allows us to easily use numeric summary functions on them:
    -   `sum()` tells us the **count** of `TRUE` values
    -   `mean()` tells us the **proportion** of `TRUE` values

```{r}
mpg |> 
  summarize(
    n_efficient = sum(hwy > 30),
    p_efficient = mean(hwy > 30)
  )
```

## any() and all() {.fs80}

-   Does a condition exist at all? Does it apply universally?
-   `any()` returns `TRUE` if **at least one** value is `TRUE`
-   `all()` returns `TRUE` if **every** value is `TRUE`

```{r}
mpg |> 
  summarize(
    .by = drv,
    any_efficient = any(hwy > 30),
    all_efficient = all(hwy > 30)
  )
```

# If-Else

## What is Control Flow? {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Control Flow** allows for conditional outcomes, i.e., the output changes depending on the input
-   Think of it like a fork in the road. You ask a question (a condition), and depending on the answer, you take a different path.
-   The tidyverse provides `if_else()` for two-way forks and `case_when()` for many
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/branch.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## The if_else() Function

:::{.fs80}
The `if_else()` function takes three main arguments:

1.  `condition`: A logical test (e.g., `x > 10`)
2.  `true`: The output if the condition is `TRUE`
3.  `false`: The output if the condition is `FALSE`
:::

```{r}
x <- 6
if_else(
  condition = x > 10,
  true = "large",
  false = "small"
)
```

## Vectorized if_else()

```{r}
#| echo: true
# if_else() is vectorized, meaning it works on an entire series of data at once
x <- c(6, 16, 3, 20)

if_else(x > 10, "large", "small")
```

## Creating our Site Data

```{r}
#| echo: true
# Let's create a dataset of site temperatures
sitetemps <- 
  tibble(
    site = c(2, 1, 2, 3, 1),
    temp = c(-2.1, 38.6, 4.6, -0.2, 37.6)
  ) |> 
  print()
```


## Using in a pipeline

```{r}
#| echo: true
# Let's assign a temperature metric based on the site number
sitetemps |> 
  mutate(
    metric = if_else(
      condition = site == 1,
      true = "F",
      false = "C"
    )
  )
```

# Case-When

## Handling Complex Logic {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   `if_else()` is great for binary choices
-   But what if we have three, four, or ten different possibilities?
-   Nesting `if_else()` statements inside each other gets messy quickly.
-   `case_when()` is designed specifically for complex, multi-condition logic.
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/branches.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## The case_when() Function 

:::{.fs80}
-   `case_when()` evaluates a sequence of formulas
-   The left side of the `~` is the condition
-   The right side of the `~` is the output
-   It evaluates from top to bottom.
-   The first `TRUE` condition wins!
:::

```{r}
x <- c(6, 16, 3)
case_when(
  x > 10 ~ "large",
  x > 5  ~ "medium",
  x <= 5 ~ "small"
)
```

## The .default Argument

```{r}
# We can use `.default` to catch anything that doesn't match our specific rules
x <- c(6, 16, 3)

case_when(
  x > 10 ~ "large",
  x > 5  ~ "medium",
  .default = "small"
)
```

## Pitfall: Ordering Matters!

```{r}
x <- c(15)

# Watch what happens if we put the broader condition first:
case_when(
  x > 5  ~ "medium", 
  x > 10 ~ "large",   # 15 is > 10, but it already triggered "medium"!
  .default = "small"
)
```

## Using in a pipeline

```{r}
# Assigning nations based on multiple site codes
sitetemps |> 
  mutate(
    nation = case_when(
      site == 1 ~ "US",
      site == 2 ~ "DE",
      site == 3 ~ "FR",
      .default = "Unknown"
    )
  )
```

## Complex Transformations

```{r}
sitetemps |> 
  mutate(
    metric = if_else(site == 1, "F", "C"),
    temp_c = if_else(metric == "F", (temp - 32) * (5 / 9), temp),
    state = case_when(
      temp_c < 0 ~ "solid",
      temp_c < 100 ~ "liquid",
      temp_c >= 100 ~ "gas"
    )
  )
```

## Summary {.fs80}

-   **Comparisons:** Evaluate relationships using `>`, `<`, `==`, and `!=`
-   **Boolean Algebra:** Combine conditions using `&`, `|`, and `!`
-   **Summaries:** Use `sum()` to count TRUEs, `mean()` for proportions, and `any()`/`all()` for group-level checks
-   **Control Flow:** Handle two-way branches seamlessly with `if_else()`
-   **Complex Cases:** Manage multiple condition mapping safely with `case_when()`
