---
lecture: "14b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(patchwork)

options(
  tibble.print_max = 5,
  tibble.print_min = 5
)

set_theme(theme_gray(base_size = 24))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Package repetitive code into custom functions using the DRY principle

- Control function behavior using arguments, defaults, and branching logic

- Apply custom vector functions inside tidyverse data pipelines
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Why Write Functions?

## The DRY Principle {.fs80}

::: {.columns .pv4}

::: {.column width="60%"}
-   As you code, you may find yourself copying and pasting code over and over.
-   This violates the **DRY Principle**:<br>
    *"Don't Repeat Yourself!"*
-   Repetition increases the chance of errors and makes your code harder to read.
-   [Functions]{.b .blue} allow us to package code into reusable (and sharable) tools.
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/fn_key.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Repetitive Code Example

```{r}
#| echo: true
library(tidyverse)
data("mpg", package = "ggplot2")
data("starwars", package = "dplyr")

# We want quick histograms to explore our data
p1 <- ggplot(mpg, aes(x = hwy)) + geom_histogram()
p2 <- ggplot(mpg, aes(x = cty)) + geom_histogram()
p3 <- ggplot(starwars, aes(x = height)) + geom_histogram()
```

This is a lot of typing just to look at three distributions!

:::{.fragment}
Could we make a function to do this more efficiently?
:::

# Anatomy of a Function

## How Functions Work 

:::{.fs80}
A function declaration always requires four core parts:

1.  **Name:** How we will call the function later.
2.  **Arguments:** The inputs the function expects.
3.  **Body:** The actual code that does the work.<br>*(This can be ANY valid R code, e.g., wrangling, visualizing, modeling)*
4.  **Return:** The final output given back to the user.
:::

``` r
function_name <- function(arguments) { 
  # function_body
  # ...
  # return_value
}
```

## Our First Function

::: {.fs80}
Let's build a function to automate our histogram plotting.<br>
It will take a single vector of numbers and return a plot.
:::

```{r}
gghist <- function(vector) {
  
  # 1. Put the vector into a dataframe
  df <- tibble(x = vector)
  
  # 2. Build the plot
  p <- ggplot(df, aes(x = x)) + geom_histogram()
  
  # 3. Return the plot
  p
}
```

## Using Our Function

Now we can generate our plots with less typing!

:::{.columns}
:::{.column}
```{r}
gghist(mpg$hwy)
```
:::
:::{.column}
```{r}
gghist(starwars$height)
```
:::
:::

# Arguments & Defaults

## Adding More Arguments

:::{.fs80}
What if we want to change the fill color of our histogram or the text of our<br> x-axis label? We can simply add more arguments and use them in the body.
:::

```{r}
gghist_color <- function(vector, fcolor, xlabel) {
  
  df <- tibble(x = vector)
  
  p <- ggplot(df, aes(x = x)) + 
    geom_histogram(fill = fcolor) +
    labs(x = xlabel)
  
  p
}
```

## Using the Arguments

```{r}
#| message: false

gghist_color(mpg$hwy, fcolor = "firebrick", xlabel = "Highway MPG")
```


## The Missing Argument Error {.fs80}

If we use our new function but forget to provide the new arguments, R will throw an error because it doesn't know what color/label to use.

```{r}
#| echo: true
#| error: true
gghist_color(mpg$hwy)
```

We can fix this by setting **default argument values**.

## Setting Defaults

::: {.fs80}
We define defaults in the `function()` declaration. The default will only be used if the argument is missing. The user can still overwrite the defaults!
:::

```{r}
gghist_safe <- function(vector, fcolor = "black", xlabel = "Variable") {
  
  df <- tibble(x = vector)
  
  p <- ggplot(df, aes(x = x)) + 
    geom_histogram(fill = fcolor) +
    labs(x = xlabel)
  
  p
}
```

## Using the Defaults

:::{.columns}
:::{.column}
```{r}
# Provide neither arg
gghist_safe(mpg$hwy)
```
:::
:::{.column}
```{r}
# Provide one only arg
gghist_safe(mpg$hwy, fcolor = "blue")
```

:::
:::

# Validation and<br> Control Flow

## Designing for Safety

::: {.fs80}
When sharing functions, the user may not provide the correct data type.<br>
Let's make a simple function to "center" a variable by subtracting its mean.
:::

```{r}
center_var <- function(x) {
  x - mean(x, na.rm = TRUE)
}
center_var(c(8, 10, 12))
```

```{r}
#| error: true

center_var(c("apple", "banana", "cherry"))
```

::: {.fs80}
Our function gives an error message, but it is not a very helpful one...

Luckily, we can modify our function to do better!
:::

## Input Validation

:::{.fs80}
We can use an `if()` statement at the very beginning of our function to check the input. If the input is wrong, we use `stop()` to print an error message.
:::

```{r}
#| error: true

center_var_safe <- function(x) {
  
  if (!is.numeric(x)) {
    stop("The input vector 'x' must be numeric.")
  }
  
  x - mean(x, na.rm = TRUE)
}

center_var_safe(c("apple", "banana", "cherry"))
```

## Adding Logical Controls 

:::{.fs80}
We can add logical arguments (`TRUE` or `FALSE`) to give users control over how the function operates. Let's add a `robust` argument to our centering function.
:::

``` r
center_var_adv <- function(x, robust = FALSE) {
  
  if (!is.numeric(x)) { stop("Input 'x' must be numeric.") }
  if (!is.logical(robust)) { stop("Input 'robust' must be logical.")}
  
  # But how do we use the robust argument?
  # - we want to subtract the mean if robust is FALSE (default)
  # - we want to subtract the median if robust is TRUE
}
```

## Branching Logic

:::{.fs80}
We can use an `if` and `else` block to create two branches in our function.<br>
The code will do different things based on the `robust` argument's value.
:::

```{r}
center_var_adv <- function(x, robust = FALSE) {
  
  if (!is.numeric(x)) { stop("Input 'x' must be numeric.") }
  if (!is.logical(robust)) { stop("Input 'robust' must be logical.")}
  
  if (robust == TRUE) {
    # The robust branch uses the median
    centered <- x - median(x, na.rm = TRUE)
  } else {
    # The standard branch uses the mean
    centered <- x - mean(x, na.rm = TRUE)
  }
  
  centered
}
```

## Using the Branching Function

::: {.fs80}
Because the mean is highly sensitive to outliers, our `robust` argument provides a very helpful alternative path!
:::

```{r}
my_data <- c(1, 2, 3, 4, 100) # Notice the outlier
```

```{r}
# Standard centering is greatly influenced by the outlier
center_var_adv(my_data, robust = FALSE)

# Robust centering is not influenced by the outlier
center_var_adv(my_data, robust = TRUE)
```

# Functions in Pipelines

## The Payoff in mutate()

::: {.fs80}
Because our `center_var_adv()` function takes a vector and returns a vector of the same length, it drops perfectly into our tidyverse pipelines.
:::

```{r}
mpg |> 
  select(manufacturer, model, hwy) |> 
  mutate(
    hwy_cent = center_var_adv(hwy),
    hwy_rob = center_var_adv(hwy, robust = TRUE)
  )
```

## Prepping for Iteration

::: {.fs80}
You can build any mathematical transformation you need and apply it the same way.
For example, a common task is standardization (z-scoring):
:::

```{r}
standardize <- function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}
mpg |> 
  select(manufacturer, model, cty) |> 
  mutate(cty_z = standardize(cty))
```

## Summary {.fs80}

-   **Limitless Potential:** Functions can package *any* valid R code or workflow. 
-   **Reduce Repetition:** They apply the DRY principle.
-   **Structure:** They require a name, arguments, a body, and a return value.
-   **Arguments & Logic:** You can set default values and use `if`/`else` to branch function behavior.
-   **Validation:** Use `if()` and `stop()` to ensure the correct data is provided.
-   **Functions in Pipelines:** Custom functions that return vectors drop seamlessly into `mutate()` and prepare us for iteration.
