---
lecture: "14c"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(patchwork)

options(
  tibble.print_max = 5,
  tibble.print_min = 5
)

set_theme(theme_gray(base_size = 20))

if(!dir.exists("data")) dir.create("data")
set.seed(2023)
write_csv(tibble(day = 1:30, mood = round(rnorm(30, 5, 1))), "data/pid_001.csv")
write_csv(tibble(day = 1:30, mood = round(rnorm(30, 4, 1))), "data/pid_002.csv")
write_csv(tibble(day = 1:30, mood = round(rnorm(30, 4, 2))), "data/pid_003.csv")
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Iterate functions across multiple columns using `across()`

- Use list objects to store complex, mixed-type data

- Efficiently apply functions to vectors using `map()`
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Iterating Over Columns

## The Repetition Problem

::: {.fsmaller}
We often need to apply the exact same function to many different columns. Doing this manually requires a lot of repetitive, error-prone typing.
:::

```{r}
library(tidyverse)
data("mpg", package = "ggplot2")

# Calculating the mean for multiple columns manually
mpg |> 
  summarize(
    cty_mean = mean(cty),
    hwy_mean = mean(hwy),
    displ_mean = mean(displ)
  )
```

## Introducing across()

::: {.fsmaller}
The `across()` function solves this by allowing us to apply the same transformation to multiple columns simultaneously. It takes two primary arguments: `.cols` (which columns) and `.fns` (what function).
:::

```{r}
# This does the exact same thing, but much more efficiently
mpg |> 
  summarize(
    across(.cols = c(cty, hwy, displ), .fns = mean)
  )
```

## Selecting Columns Smartly

::: {.fsmaller}
We can use tidy-select helpers (like `where()`, `starts_with()`, or `contains()`) inside of `across()` to target specific types of columns.
:::

```{r}
# Apply the mean function to all numeric columns at once
mpg |> 
  summarize(
    across(.cols = where(is.numeric), .fns = mean)
  )
```

## Custom Functions Work Too

::: {.fsmaller}
Importantly, `across()` works perfectly with custom functions! It also drops right into `mutate()` just as easily as `summarize()`.
:::

```{r}
# Our standardization function from Lecture 14b
standardize <- function(x) { (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE) }

mpg |> 
  mutate(across(.cols = c(cty, hwy), .fns = standardize)) |> 
  select(manufacturer, cty, hwy)
```

# Anonymous Functions

## The Need for Quick Functions

::: {.fsmaller}
Sometimes we need to apply a simple operation across many columns, but it feels overkill to create and save a custom, named function for it.
:::

```{r}
# Do we really need to save this just to use it once?
multiply_by_ten <- function(x) { x * 10 }

mpg |> 
  mutate(across(c(cty, hwy), multiply_by_ten)) |> 
  select(manufacturer, cty, hwy)
```

## The \(x) Syntax

::: {.fsmaller}
R allows us to write **anonymous functions** (functions without a name) directly inside our pipelines using the `\(x)` syntax. This simply tells R: *"I am about to write a temporary function that takes a single argument that we can call x."*
:::

```{r}
# This code does the exact same thing, but inline!
mpg |> 
  mutate(across(c(cty, hwy), \(x) x * 10)) |> 
  select(manufacturer, cty, hwy)
```

## Passing Arguments

::: {.fsmaller}
One of the biggest advantage of anonymous functions is how easily they let us pass additional arguments to the function we are iterating (e.g., na.rm).
:::

:::{.columns}
:::{.column}
```{r}
# How to pass na.rm = TRUE to mean()?
starwars |> 
  summarize(
    across(
      .cols = c(mass, height), 
      .fns = mean
    )
  )
```
:::
:::{.column .fragment}
```{r}
# With an anonymous function!
starwars |> 
  summarize(
    across(
      .cols = c(mass, height), 
      .fns = \(x) mean(x, na.rm = TRUE)
    )
  )
```
:::
:::

# Understanding Lists

## The Limits of Standard Vectors

::: {.fsmaller}
- We know how to store data in standard, atomic vectors using `c()`.
- **The limitation:** Every item in a vector must be the same data type.
    + If you mix multiple types, `c()` converts them to be the same.
- **The solution:** A [list]{.b .blue} is a special "super vector" that can hold *anything*.

:::

:::{.columns}
:::{.column}
```{r}
# logicals < *numbers* < characters
c(0, 1, 0, TRUE, FALSE)
# logicals < numbers < *characters*
c(42.1, "Hello World", TRUE)
```
:::
:::{.column .fragment}
```{r}
list(42.1, "Hello World", TRUE)
```
:::
:::

## Lists Can Store Anything

```{r}
wow <- list(
  tibble(x = 1:2, y = c("a", "b")),
  ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
)
```

:::{.columns}
:::{.column}
```{r}
wow[[1]]
```

:::
:::{.column}
```{r}
wow[[2]]
```

:::
:::


## Naming List Elements

:::{.columns}
:::{.column}
```{r}
# Set names during creation
my_list <- 
  list(
    myNumbers = c(1, 2, 3),
    myLetters = c("x", "y")
  ) |> 
  print()
```
:::
:::{.column}
```{r}
# Set names after creation
my_list <- 
  list(
    c(1, 2, 3),
    c("x", "y")
  ) |> 
  set_names(
    c("myNumbers", "myLetters")
  ) |> 
  print()
```
:::
:::

# Iterating Over Vectors

## Two Ways to Iterate {.fsmaller}

- We use `across()` to iterate over columns, but what about vectors?
- **The Traditional Approach: Loops**
    + Many programmers use *loops* to manually micro-manage iteration (e.g., creating empty containers, filling them, and tracking exact positions).
- **The Tidyverse Approach: \{purrr\}**
    + We will use [functional programming]{.b .blue} to "delegate" the boring work 
    + **Simplicity:** Provide just the vector and the function.
    + **Automation:** Automatically tracks positions and stores results.
    + **Readability:** Integrates cleanly into pipelines.

## How map() Works 

:::{.fsmaller}
- The core function `map()` takes a vector (`.x`) and a function (`.f`)
- **Toy Example:** Generating samples of different sizes
    + `runif(n)` generates `n` random numbers between 0 and 1

:::

```{r}
# Generate a sample of random numbers for each sample size
my_samples <- map(.x = c(2, 4, 3), .f = runif)
my_samples
```

## Type-Specific Maps 
:::{.fsmaller}
- If the output is all one type, we can simplify it with a **type-specific map**
    + e.g., `map_dbl()` returns a numeric vector
    + e.g., `map_chr()` returns a character vector
- **Example:** Finding the maximum value in each of our random samples

:::

:::{.columns}
:::{.column}
```{r}
# Keep output as a list
map(.x = my_samples, .f = max)
```
:::
:::{.column}
```{r}
# Simplify to a numeric vector
map_dbl(.x = my_samples, .f = max)
```
:::
:::

## Step 1: Finding Our Files 
:::{.fsmaller}
- The real power of `map()` is working with complex data structures like files.
- **The Scenario:** You have a folder full of CSV files to combine.
- First, we need to locate them using `list.files()`
    + `path`: Which folder to look in
    + `pattern`: What regex pattern to look for (ends in ".csv")
    + `full.names`: Keep this `TRUE` so R knows exactly where the file is

:::

```{r}
# This outputs a simple character vector of file paths
my_files <- list.files(path = "data", pattern = "\\.csv$", full.names = TRUE)
my_files
```

## Step 2: Reading Files into a List 
:::{.fsmaller}
- Now we have a character vector of file paths. 
- We can use `map()` to apply the `read_csv()` function to every path.
- Since `map()` always returns a list, we will get a list where each item is a complete data frame.
:::

```{r}
#| message: false

list_of_dfs <- 
  # read each file as a tibble into a list object
  map(.x = my_files, .f = read_csv) |> 
  # set the name of each element to be its file path
  set_names(my_files)
```

## Previewing the List

```{r}
list_of_dfs
```


## Step 3: Binding the Data 

:::{.fsmaller}
- We have our data, but a list of data frames is difficult to analyze
- We want one combined data frame
- `list_rbind()` will stacks them row-by-row
    + The `names_to` argument creates a helpful ID column so we know where each row came from.
    
:::

```{r}
alldata <- list_rbind(list_of_dfs, names_to = "file")
```

## Previewing the Combined Data

```{r}
# One 90-row tibble instead of three 30-row tibbles
alldata
```

## Using the Combined Data

```{r}
#| fig-width: 12
#| fig-height: 4

alldata |> 
  ggplot(aes(x = day, y = mood, color = file, group = file)) +
  geom_line() + geom_point() +
  theme(legend.position = "top")
```


## Iterating for Side Effects

```{r}
# Create function to read in data, create a plot, and save it
create_plot <- function(file) {
  # Read in the data from current file
  current_data <- read_csv(file)
  # Create plot of the current data
  current_plot <- ggplot(current_data, aes(x = day, y = mood)) +
    geom_line() + geom_point() + labs(title = file)
  # Save the plot to a png image file
  ggsave(filename = str_glue("{file}.png"), plot = current_plot,
    width = 7, height = 5, units = "in")
  # Return the plot object
  current_plot
}
```

```{r}
#| message: false

# Apply this function to each of my_files
p <- map(.x = my_files, .f = create_plot)
```

## Verifying it worked

```{r}
#| fig-width: 12
#| fig-height: 4

library(patchwork)
p[[1]] + p[[2]] + p[[3]]
```

```{r}
list.files(path = "data", pattern = "\\.png$", full.names = TRUE)
```



