---
title: "Week 14: Program"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
library(tidyverse)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [13a] Conditionals

### Topics
- Comparisons and Boolean Logic
- Logical and Numeric Summaries
- If-Else and Case-When

### Readings
- R4DS (2E) Chapter 12: [Logical vectors](https://r4ds.hadley.nz/logicals.html){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[14a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice
### Practice

#### 1) Categorizing City Mileage

Using the `mpg` dataset (from \{ggplot2\}), let's use `if_else()` to create a new categorical column based on city fuel efficiency.

- Use `mutate()` and `if_else()` to create a new column called `efficiency`.
- If `cty` is greater than or equal to 20, label it "High". Otherwise, label it "Low".
- Select the `manufacturer`, `model`, `cty`, and `efficiency` columns to view your results.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("mpg", package = "ggplot2")

mpg |> 
  mutate(
    efficiency = if_else(
      condition = cty >= 20, 
      true = "High", 
      false = "Low"
    )
  ) |> 
  select(manufacturer, model, cty, efficiency)
```

{{< dstop >}}

#### 2) Sleep Summaries

Let's use the `msleep` dataset (from \{ggplot2\}) to practice summarizing logical conditions across groups.

- Use `summarize()` to calculate two new metrics.
- First, calculate the proportion of animals that sleep more than 12 hours (`sleep_total > 12`) and name it `prop_heavy_sleepers`.
- Second, check if *any* animal in the group has a body weight greater than 500 (`bodywt > 500`) and name it `any_giants`.
- Group the summary by the `vore` column using the `.by` argument.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("msleep", package = "ggplot2")

msleep |> 
  summarize(
    prop_heavy_sleepers = mean(sleep_total > 12),
    any_giants = any(bodywt > 500),
    .by = vore
  )
```

{{< dstop >}}

#### 3) Sizing Up Starwars Characters

The `starwars` dataset (from \{dplyr\}) contains physical attributes of characters. Let's practice handling multiple conditions using `case_when()`.

- Use `mutate()` and `case_when()` to create a new column called `size_category` based on the `mass` column.
- If `mass` is less than 50, label it "Light".
- If `mass` is less than 100, label it "Medium".
- If `mass` is 100 or greater, label it "Heavy".
- Use the `.default` argument to label any remaining cases (like missing values) as "Unknown".
- Select the `name`, `mass`, and `size_category` columns to verify your logic worked correctly. *Note: Remember that `case_when()` evaluates from top to bottom!*

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("starwars", package = "dplyr")

starwars |> 
  mutate(
    size_category = case_when(
      mass < 50 ~ "Light",
      mass < 100 ~ "Medium",
      mass >= 100 ~ "Heavy",
      .default = "Unknown"
    )
  ) |> 
  select(name, mass, size_category)
```

{{< dstop >}}

## [14b] Functions

### Topics
- A
- B
- C

### Readings
- R4DS (2E) Chapter 25: [Functions](https://r4ds.hadley.nz/functions.html){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[14b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice



## [14c] Iteration

### Topics
- A
- B
- C

### Readings
- R4DS (2E) Chapter 26: [Iteration](https://r4ds.hadley.nz/iteration.html){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[14c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice



<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../12/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 12</a>

  <a href="../14/index.html" class="btn btn-primary" role="button">Week 14&nbsp;&nbsp;&raquo;</a>
</div>
```
