---
title: "Week 14: Program"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
library(tidyverse)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [14a] Conditionals

### Topics
- Evaluate comparisons and combine multiple conditions using Boolean logic
- Summarize logical vectors to count occurrences and calculate proportions
- Control data transformations conditionally using `if_else()` and `case_when()`

### Readings
- R4DS (2E) Chapter 12: [Logical vectors](https://r4ds.hadley.nz/logicals.html){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[14a] Slideshow">
</iframe>
```

::: {.fs80}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Categorizing City Mileage

Using the `mpg` dataset (from \{ggplot2\}), let's use `if_else()` to create a new categorical column based on city fuel efficiency.

- Use `mutate()` and `if_else()` to create a new column called `efficiency`.
- If `cty` is greater than or equal to 20, label it "High". Otherwise, label it "Low".
- Select the `manufacturer`, `model`, `cty`, and `efficiency` columns to view your results.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("mpg", package = "ggplot2")

mpg |> 
  mutate(
    efficiency = if_else(
      condition = cty >= 20, 
      true = "High", 
      false = "Low"
    )
  ) |> 
  select(manufacturer, model, cty, efficiency)
```

{{< dstop >}}

#### 2) Sleep Summaries

Let's use the `msleep` dataset (from \{ggplot2\}) to practice summarizing logical conditions across groups.

- Use `summarize()` to calculate two new metrics.
- First, calculate the proportion of animals that sleep more than 12 hours (`sleep_total > 12`) and name it `prop_heavy_sleepers`.
- Second, check if *any* animal in the group has a body weight greater than 500 (`bodywt > 500`) and name it `any_giants`.
- Group the summary by the `vore` column using the `.by` argument.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("msleep", package = "ggplot2")

msleep |> 
  summarize(
    prop_heavy_sleepers = mean(sleep_total > 12),
    any_giants = any(bodywt > 500),
    .by = vore
  )
```

{{< dstop >}}

#### 3) Sizing Up Starwars Characters

The `starwars` dataset (from \{dplyr\}) contains physical attributes of characters. Let's practice handling multiple conditions using `case_when()`.

- Use `mutate()` and `case_when()` to create a new column called `size_category` based on the `mass` column.
- If `mass` is less than 50, label it "Light".
- If `mass` is less than 100, label it "Medium".
- If `mass` is 100 or greater, label it "Heavy".
- Use the `.default` argument to label any remaining cases (like missing values) as "Unknown".
- Select the `name`, `mass`, and `size_category` columns to verify your logic worked correctly. *Note: Remember that `case_when()` evaluates from top to bottom!*

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("starwars", package = "dplyr")

starwars |> 
  mutate(
    size_category = case_when(
      mass < 50 ~ "Light",
      mass < 100 ~ "Medium",
      mass >= 100 ~ "Heavy",
      .default = "Unknown"
    )
  ) |> 
  select(name, mass, size_category)
```

{{< dstop >}}

## [14b] Functions

### Topics
- Package repetitive code into custom functions using the DRY principle
- Control function behavior using arguments, defaults, and branching logic
- Apply custom vector functions inside tidyverse data pipelines

### Readings
- R4DS (2E) Chapter 25: [Functions](https://r4ds.hadley.nz/functions.html){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[14b] Slideshow">
</iframe>
```

::: {.fs80}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Building an Outlier Flag

Let's build a function to identify unusually high values in a dataset. We will start with the basic math.

- Create a custom function named `flag_outlier()` with a single argument `x`.
- Inside the function, calculate a cutoff value: the `mean()` of `x` plus 2 times the `sd()` of `x`. (Be sure to use `na.rm = TRUE` for both!).
- Return a logical value indicating whether `x` is greater than your cutoff.
- Test your function on the `Ozone` column of the `airquality` dataset (from the \{datasets\} package).

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("airquality", package = "datasets")

flag_outlier <- function(x) {
  cutoff <- mean(x, na.rm = TRUE) + (2 * sd(x, na.rm = TRUE))
  x > cutoff
}

flag_outlier(airquality$Ozone)
```

{{< dstop >}}

#### 2) Adding Customization

Let's make the function more flexible by allowing the user to decide exactly how extreme an outlier needs to be before it gets flagged.

- Redefine your function and add a second argument called `multiplier`. Set its default value to `2`.
- Update your cutoff calculation. Instead of hardcoding a number, multiply the standard deviation by your new `multiplier` argument.
- Test your upgraded function by passing `airquality$Ozone` to it and setting the `multiplier` to `3`.
- Pipe the result of that test directly into the `any()` function to quickly check if there are *any* extreme outliers in the vector.

{{< dstart summary="Answer key" >}}

```{r}
flag_outlier <- function(x, multiplier = 2) {
  
  cutoff <- mean(x, na.rm = TRUE) + (multiplier * sd(x, na.rm = TRUE))
  
  x > cutoff
}

flag_outlier(airquality$Ozone, multiplier = 3) |> any()
```

{{< dstop >}}

#### 3) Functions in Pipelines

Because your function takes a vector and returns a logical vector of the exact same length, it drops perfectly into your data wrangling workflows.

- Open a tidyverse pipeline using the `airquality` dataset.
- Immediately pipe it to the `as_tibble()` function to make it a tibble.
- Use `mutate()` to create a new column called `is_high_ozone` using your `flag_outlier()` function.
- Use `select()` to view only the `Month`, `Day`, `Ozone`, and `is_high_ozone` columns. 

{{< dstart summary="Answer key" >}}

```{r}
airquality |> 
  as_tibble() |> 
  mutate(is_high_ozone = flag_outlier(Ozone)) |> 
  select(Month, Day, Ozone, is_high_ozone)
```

{{< dstop >}}

## [14c] Iteration

### Topics
- Iterate functions across multiple columns using `across()`
- Use list objects to store complex, mixed-type data
- Efficiently apply functions to vectors using `map()`

### Readings
- R4DS (2E) Chapter 26: [Iteration](https://r4ds.hadley.nz/iteration.html){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[14c] Slideshow">
</iframe>
```

::: {.fs80}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Summarizing Multiple Columns

We can use iteration to apply the same summary function to many columns at once without repetitive typing. Let's use the `iris` dataset (from \{datasets\}) to find the median of all numeric columns for each species.

- Use the appropriate summarizing and iteration functions to calculate the median for all numeric columns automatically. 
- Group your summary by the `Species` column.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)
data("iris", package = "datasets")

iris |> 
  summarize(
    .by = Species,
    across(.cols = where(is.numeric), .fns = median)
  )
```

{{< dstop >}}

#### 2) Inline Operations with Anonymous Functions

Anonymous functions let us apply custom mathematical transformations without having to formally save and name a new function. Let's use the `penguins` dataset (from \{datasets\})to convert several measurements from millimeters to centimeters.

- Use the appropriate mutation and iteration functions to target the `bill_len`, `bill_dep`, and `flipper_len` columns.
- Apply an anonymous function to these columns that divides their values by 10.
- Select your modified columns along with the `species` column to view the results.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("penguins", package = "datasets")

penguins |> 
  as_tibble() |> 
  mutate(
    across(
      .cols = c(bill_len, bill_dep, flipper_len),
      .fns = \(x) x / 10
    )
  ) |> 
  select(species, bill_len, bill_dep, flipper_len)
```

{{< dstop >}}

#### 3) Iterating over Lists of Complex Objects

While `across()` is the best tool for iterating over columns inside a single dataset, `map()` shines when working with lists of complex objects like multiple data frames. Let's practice by doing a quick sanity check on the size of several built-in datasets at once.

- Create a list named `my_datasets` containing the `mtcars`, `iris`, and `ToothGrowth` datasets (all from \{datasets\}).
- Use the `set_names()` function to name the elements of your list "mtcars", "iris", and "ToothGrowth" so your final output is easily readable.
- Use `map_int()` to iterate over your list and return an integer vector.
- Set the `.f` argument to the `nrow` function to extract the number of rows for each dataset.

{{< dstart summary="Answer key" >}}

```{r}
library(tidyverse)
data("mtcars", "iris", "ToothGrowth", package = "datasets")

my_datasets <- 
  list(mtcars, iris, ToothGrowth) |> 
  set_names(c("mtcars", "iris", "ToothGrowth"))

map_int(.x = my_datasets, .f = nrow)
```

{{< dstop >}}

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../13/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 13</a>
</div>
```
