---
lecture: "06a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

set_theme(theme_gray(base_size = 30))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Style your code to improve organization and clarity

- Manage the behavior of individual code chunks in Quarto
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Code Style

## Code Style {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- Style describes *optional* changes

  - e.g., how to name files and objects
  
  - e.g., where to add spaces and line breaks

::: {.fragment .mt1}
- There is no "right" or "wrong" with style
:::

::: {.fragment .mt1}
- But using a consistent code style has benefits

  - Code becomes more readable and predictable
  
  - Collaboration becomes easier and smoother
:::

::: {.fragment .mt1}
- We will learn about the [tidyverse style guide](https://style.tidyverse.org)
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/stylist.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## File Names {.smaller}

- File names should be [meaningful]{.b .blue} and [clear]{.b .blue}

  -   [`models.qmd`]{.danger} ➔ `dissertation_study2_models.qmd`

::: {.fragment .mt1}
- [Avoid special characters]{.b .blue} in names (arguably this includes spaces)

  -   [`M@$teR$ Th3s1s.r`]{.danger} ➔ `masters_thesis.R`
  
:::

::: {.fragment .mt1}
- If your files need to be run in order, [prefix with numbers]{.b .blue}

  -   `1_import.qmd`, `2_model.qmd`, `3_visualize.qmd`
  
:::

::: {.fragment .mt1}
- Use [zero-padding]{.b .blue} as necessary (to sort them properly)

  -   `01_download.qmd`, `02_tidy.qmd`, ..., `10_visualize.qmd`
  
:::

## Sectioning and Comments {.smaller}

- Use [sections and subsections]{.b .blue} to give the file internal structure

- Load [all packages together]{.b .blue} at the top of the document

- Use [comments to explain]{.b .blue} the "why" (not the "what" or "how")

## Object Naming {.smaller}

- Object names should use a **single** and [consistent style]{.b .blue}
  - [`countydata` and `plot_county`]{.danger}
  - **Snake Case:** `county_data` and `county_plot`
  - **Camel Case:** `countyData` and `countyPlot`

::: {.fragment .mt1}
- Use brief, descriptive [noun phrases]{.b .blue} for object names
  - [`x`]{.danger} ➔ `heart_rate`
  - [`df`]{.danger} ➔ `aim1_data`
  
:::

::: {.fragment .mt1}
- [Avoid reusing names]{.b .blue} of base R objects and functions
  - e.g., [`F`, `T`, `c`, `mean`, `sum`, `pi`, `data`]{.danger}
  
:::


## Spacing {.smaller}

- Put spaces [around most operators]{.b .blue} (arithmetic, relational, and logical)
  - [`1/2+3*4-5`]{.danger} ➔  `1 / 2 + 3 * 4 - 5`
  - [`a>3&b<=0`]{.danger} ➔ `a > 3 & b <= 0`

::: {.fragment .mt1}
- But don't put spaces [around "high priority" operators]{.b .green}
  - [`5 ^ 2`]{.danger} ➔ `5^2`
  - [`( 1 + 2 ) * 3`]{.danger} ➔ `(1 + 2) * 3`
  - [`sum ( sales )`]{.danger} ➔ `sum(sales)`
  - [`df $ variable`]{.danger} ➔ `df$variable`
  
:::

::: {.fragment .mt1}
- Always add a space [after a comma]{.b .blue} but not before a comma
  - [`c(1 ,2,3 ,4, 5)`]{.danger} ➔ `c(1, 2, 3, 4, 5)`
  
:::


## Argument Specification {.smaller}

- Function arguments usually either provide [data]{.b .green} or customize [details]{.b .blue}
  - e.g., the `round()` function has two arguments:
  - `x` contains the number(s) to be rounded (a [data]{.b .green} argument)
  - `digits` contains the number of digits to round to (a [details]{.b .blue} argument)

::: {.fragment .mt1}
- In a function call, [omit]{.b .green} (i.e., remove) the names of [data]{.b .green} arguments
  - e.g., omit `x = `
  
:::

::: {.fragment .mt1}
- In a function call, [include]{.b .blue} the full names of [detail]{.b .blue} arguments
  - e.g., include `details = `
  
:::

::: {.fragment .mt1}
- [`round(x = 2 / 3, digits = 1)`]{.danger} ➔ `round(2 / 3, digits = 1)`

:::


## Line Length and Indents

:::{.fsmaller}
- Strive to limit your code to just **80 characters per line**
  - *Tools > Global Options > Code > Display > Show Margin (Column=80)*

::: {.fragment .mt1}
- If a line is longer than 80 characters, [break it up]{.b .blue} and [align it]{.b .green}
  - Put each argument on a separate, indented line
  
:::
:::

::: {.fragment .mt1}
```{r}
#| eval: false
#| echo: true

do_something_very_complicated(
  that = "requires",
  many = arguments,
  some = "of which may be very long"
)
```
:::


## Pipes and Pipelines

:::{.fsmaller}
- Line break and [indent after assignment]{.b .blue}

- Follow each pipe with a [line break]{.b .blue}

- [Separate long lines]{.b .blue} of arguments with line breaks

- [Indent further]{.b .blue} for arguments
:::

```{r}
#| eval: false
#| echo: true

x <- 
  df |> 
  step_one(arg = TRUE) |> 
  step_two(
    arg1 = 1,
    arg2 = 2
  ) |> 
  step_three()
```


# Chunk Options

## Chunk Options {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- The YAML header customizes our document
  - e.g., setting the `author:` or `format:` fields

::: {.fragment .mt1}
- YAML can also customize each code chunk
  - We add a special comment to the top of it
  - The syntax we use is called the "hash pipe"<br />
  `#| field: value`
  
:::

::: {.fragment .mt1}
- We can also give each chunk a unique name
  - This is helpful when troubleshooting
  - It also helps when caching results
  
:::

:::


::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/doc_cog.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Suppress Messages

Sometimes a chunk will issue messages

```{r}
#| message: true
library(mgcv)
```

We can hide them in Quarto's rendered output

````{verbatim, lang = "qmd"}
```{r}
#| message: false
library(mgcv)
```
````

## Suppress Warnings

Sometimes a chunk will issue warnings

```{r}
#| warning: true
as.numeric(c("1", "2", "NA", "4"))
```

We can hide them in Quarto's rendered output

````{verbatim, lang = "qmd"}
```{r}
#| warning: false
as.numeric(c("1", "2", "NA", "4"))
```
````

```{r}
#| warning: false
as.numeric(c("1", "2", "NA", "4"))
```

## Include Errors

Errors will usually prevent Quarto from rendering at all

But sometimes (e.g., for teaching) we do want to render them

````{verbatim, lang = "qmd"}
```{r}
#| error: true
x <- 1 2 3 4 5
```
````

```{r}
#| error: true
x <- 1 2 3 4 5
```

