---
lecture: "06a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

set_theme(theme_gray(base_size = 30))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Style your code to improve organization and clarity

- Manage the behavior of individual code chunks in Quarto
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Code Style

## Code Style {.fs70}

::: {.columns .pv4}

::: {.column width="60%"}
- Style describes *optional* changes

  - e.g., how to name files and objects
  
  - e.g., where to add spaces and line breaks

::: {.fragment .mt1}
- There is no "right" or "wrong" with style
:::

::: {.fragment .mt1}
- But using a consistent code style has benefits

  - Code becomes more readable and predictable
  
  - Collaboration becomes easier and smoother
:::

::: {.fragment .mt1}
- We will learn about the [tidyverse style guide](https://style.tidyverse.org)
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/stylist.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## File Names {.fs70}

- File names should be [meaningful]{.b .blue} and [clear]{.b .blue}

  -   [`models.qmd`]{.danger} ➔ `dissertation_study2_models.qmd`

::: {.fragment .mt1}
- [Avoid special characters]{.b .blue} in names (arguably this includes spaces)

  -   [`M@$teR$ Th3s1s.r`]{.danger} ➔ `masters_thesis.R`
  
:::

::: {.fragment .mt1}
- If your files need to be run in order, [prefix with numbers]{.b .blue}

  -   `1_import.qmd`, `2_model.qmd`, `3_visualize.qmd`
  
:::

::: {.fragment .mt1}
- Use [zero-padding]{.b .blue} as necessary (to sort them properly)

  -   `01_download.qmd`, `02_tidy.qmd`, ..., `10_visualize.qmd`
  
:::

## Sectioning and Comments {.fs70}

- Use [sections and subsections]{.b .blue} to give the file internal structure

- Load [all packages together]{.b .blue} at the top of the document

- Use [comments to explain]{.b .blue} the "why" (not the "what" or "how")

## Object Naming {.fs70}

- Object names should use a **single** and [consistent style]{.b .blue}
  - [`countydata` and `plot_county`]{.danger}
  - **Snake Case:** `county_data` and `county_plot`
  - **Camel Case:** `countyData` and `countyPlot`

::: {.fragment .mt1}
- Use brief, descriptive [noun phrases]{.b .blue} for object names
  - [`x`]{.danger} ➔ `heart_rate`
  - [`df`]{.danger} ➔ `aim1_data`
  
:::

::: {.fragment .mt1}
- [Avoid reusing names]{.b .blue} of base R objects and functions
  - e.g., [`F`, `T`, `c`, `mean`, `sum`, `pi`, `data`]{.danger}
  
:::


## Spacing {.fs70}

- Put spaces [around most operators]{.b .blue} (arithmetic, relational, and logical)
  - [`1/2+3*4-5`]{.danger} ➔  `1 / 2 + 3 * 4 - 5`
  - [`a>3&b<=0`]{.danger} ➔ `a > 3 & b <= 0`

::: {.fragment .mt1}
- But don't put spaces [around "high priority" operators]{.b .green}
  - [`5 ^ 2`]{.danger} ➔ `5^2`
  - [`( 1 + 2 ) * 3`]{.danger} ➔ `(1 + 2) * 3`
  - [`sum ( sales )`]{.danger} ➔ `sum(sales)`
  - [`df $ variable`]{.danger} ➔ `df$variable`
  
:::

::: {.fragment .mt1}
- Always add a space [after a comma]{.b .blue} but not before a comma
  - [`c(1 ,2,3 ,4, 5)`]{.danger} ➔ `c(1, 2, 3, 4, 5)`
  
:::


## Argument Specification {.fs70}

- Function arguments usually either provide [data]{.b .green} or customize [details]{.b .blue}
  - e.g., the `round()` function has two arguments:
  - `x` contains the number(s) to be rounded (a [data]{.b .green} argument)
  - `digits` contains the number of digits to round to (a [details]{.b .blue} argument)

::: {.fragment .mt1}
- In a function call, [omit]{.b .green} (i.e., remove) the names of [data]{.b .green} arguments
  - e.g., omit `x = `
  
:::

::: {.fragment .mt1}
- In a function call, [include]{.b .blue} the full names of [detail]{.b .blue} arguments
  - e.g., include `details = `
  
:::

::: {.fragment .mt1}
- [`round(x = 2 / 3, digits = 1)`]{.danger} ➔ `round(2 / 3, digits = 1)`

:::


## Line Length and Indents

:::{.fs80}
- Strive to limit your code to just **80 characters per line**
  - *Tools > Global Options > Code > Display > Show Margin (Column=80)*

::: {.fragment .mt1}
- If a line is longer than 80 characters, [break it up]{.b .blue} and [align it]{.b .green}
  - Put each argument on a separate, indented line
  
:::
:::

::: {.fragment .mt1}
```{r}
#| eval: false
#| echo: true

do_something_very_complicated(
  that = "requires",
  many = arguments,
  some = "of which may be very long"
)
```
:::


## Pipes and Pipelines

:::{.fs80}
- Line break and [indent after assignment]{.b .blue}

- Follow each pipe with a [line break]{.b .blue}

- [Separate long lines]{.b .blue} of arguments with line breaks

- [Indent further]{.b .blue} for arguments
:::

```{r}
#| eval: false
#| echo: true

x <- 
  df |> 
  step_one(arg = TRUE) |> 
  step_two(
    arg1 = 1,
    arg2 = 2
  ) |> 
  step_three()
```


# Chunk Options

## Chunk Options {.fs70}

::: {.columns .pv4}

::: {.column width="60%"}
- The YAML header customizes our document
  - e.g., setting the `author:` or `format:` fields

::: {.fragment .mt1}
- YAML can also customize each code chunk
  - We add a special comment to the top of it
  - The syntax we use is called the "hash pipe"<br />
  `#| field: value`
  
:::

::: {.fragment .mt1}
- We can also give each chunk a unique name
  - This is helpful when troubleshooting
  - It also helps when caching results
  
:::

:::


::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/doc_cog.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Hash Pipe Rules

:::{.fs80}
- Hash pipes must be the [very first lines]{.b .blue} inside the chunk.
- You can [combine options]{.b .green} by putting each on a new line.

:::

::: {.fragment .mt1}
````{verbatim, lang = "qmd"}
```{r}
#| label: summarize-data
#| echo: false

# Regular R code safely goes below the options!
summary(mtcars)
```
````
:::

::: {.fragment .callout-important title="Danger"}
Do not include empty lines or R comments above the hash pipes.<br>Quarto will stop reading options as soon as it hits normal code or empty space.
:::

## Message

Sometimes a chunk will issue messages, which we may want to hide in our rendered Quarto output file.

```{r}
#| message: true
library(mgcv)
```

This is often a side-effect of loading packages. Some packages like tidyverse give easy ways to suppress this, but others don't.

## Message

We can add a hashpipe and set *message* to false.

````{verbatim, lang = "qmd"}
```{r}
#| message: false
library(mgcv)
```
````

Produces: 

```{r}
#| message: false
library(mgcv)
```

## Warning

Warnings are somewhere between messages and errors.

```{r}
#| warning: true
as.numeric(c("1", "2", "NA", "4"))
```

:::{.fragment}
````{verbatim, lang = "qmd"}
```{r}
#| warning: false
as.numeric(c("1", "2", "NA", "4"))
```
````

Produces: 

```{r}
#| warning: false
as.numeric(c("1", "2", "NA", "4"))
```
:::

## Error

Errors will usually prevent Quarto from rendering at all. But sometimes (e.g., for teaching) we *do* want to render them.

````{verbatim, lang = "qmd"}
```{r}
#| error: true
x <- 1 2 3 4 5
```
````

Produces: 

```{r}
#| error: true
x <- 1 2 3 4 5
```

## Eval

Sometimes we want to [show code without running it]{.b .blue}. This is helpful for showing examples that take too long to run or for demonstrating syntax.

````{verbatim, lang = "qmd"}
```{r}
#| eval: false
remove.packages("tidyverse")
```
````

Produces:

```{r}
#| eval: false
remove.packages("tidyverse") # Doesn't actually run
```

## Echo

Sometimes we want to [run the code without showing it]{.b .blue} (i.e., without *echoing* it). This is very common in reports where the audience only cares about the final results, plot, or table.

````{verbatim, lang = "qmd"}
```{r}
#| echo: false
summary(mtcars$mpg)
```
````

Produces:

```{r}
#| echo: false
summary(mtcars$mpg)
```

## Code Fold

An alternative to *echo* is to [let the user fold/unfold the code]{.b .blue}.

````{verbatim, lang = "qmd"}
```{r}
#| code-fold: true
summary(mtcars$mpg)
```
````

Produces:

```{r}
#| code-fold: true
summary(mtcars$mpg)
```

## Collapse

:::{.fs80}
By default, Quarto puts your code and its output into [separate blocks]{.b .blue}.

We can use the *collapse* option to [merge them]{.b .green} into a single, compact block.
:::

````{verbatim, lang = "qmd"}
```{r}
#| collapse: true
summary(mtcars$mpg)
```
````

Produces:

```{r}
#| collapse: true
summary(mtcars$mpg)
```


## Label {.fs80}

We can give each code chunk a unique name using the *label* option.

- Makes troubleshooting easier by identifying which chunk failed
- Required for cross-referencing figures and tables in your text
- Labels must be strictly unique across the entire document 

````{verbatim, lang = "qmd"}
```{r}
#| label: import-data
aim1_data <- read_csv("aim1_data.csv")
```
````

::: {.callout-tip title="The Setup Label"}
The "setup" label is special! RStudio will automatically run the chunk with this label before running any other chunk in interactive mode. Load your packages and data in this chunk.
:::

## Figure Options

We can use hash pipes to control how generated plots are displayed in our rendered document. 
Helpful options include *fig-width*, *fig-height*, *fig-align*, and *fig-cap*.

````{verbatim, lang = "qmd"}
```{r}
#| label: mpg-plot
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: left
#| fig-cap: "Miles per gallon by weight."

mtcars |> ggplot(aes(x = wt, y = mpg)) + geom_point()
```
````

## Figure Options

```{r}
#| label: mpg-plot
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: left
#| fig-cap: "Miles per gallon by weight and horsepower."

mtcars |> ggplot(aes(x = wt, y = mpg)) + geom_point()
```
