---
lecture: "06c"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} |
{{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

set_theme(theme_gray(base_size = 20))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Customize the axis and plot labels in a figure
- Customize the appearance of scales in a figure
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Labels

## Labels {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- By default, scales are named after data variables

::: {.mt1}
- Changing names may [enhance communication]{.b .blue}
  - Provide accessible names/definitions
  - Include units and other methods info.
  - Appear prettier and more professional
  
:::

::: {.fragment .mt1}
- We can also add [plot-level information]{.b .blue}
  - Titles and subtitles for take-aways
  - Captions for asides or data sources
  
:::

::: {.fragment .mt1}
- This is all done by adding on `labs()`
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/sale_tag.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Step 1: The Base Plot

```{r}
#| output-location: slide

# Create a base plot with default labels and scales
p1 <- 
  ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(size = 3, aes(color = drv, shape = drv)) + 
  geom_smooth(method = "lm", color = "black")
p1
```

## Step 2: Axis Titles

```{r}
#| output-location: slide

# Adjusting the positional (x and y) axis titles
p2 <- 
  p1 +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)"
  )
p2
```

## Step 3: Legend Titles

```{r}
#| output-location: slide

# Adjusting the color and shape legend titles
p3 <- 
  p2 +
  labs(
    color = "Drivetrain",
    shape = "Drivetrain"
  )
p3
```

## Step 4: Title, Subtitle, and Caption

```{r}
#| output-location: slide

# Adding context for key takeaways and data sourcing
p4 <- 
  p3 +
  labs(
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Lightweight sports cars with rear-wheel drive break this rule",
    caption = "Data from fueleconomy.gov"
  )
p4
```

# Scales

## Scales {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- Scales are one of the basic gg elements

::: {.mt1}
- So far, we have let R guess on the scales
  - Type (e.g., continuous or discrete)
  - Limits (minimum and maximum values)
  - Breaks (values to emphasize)
:::

::: {.fragment .mt1}
- But we can completely [customize the scales]{.b .blue}!
  - Changes may [enhance communication]{.b .green}
  - We will add on various `scale_` functions
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/sliders.json" trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::


## Step 5: Y-Axis Scale

```{r}
#| output-location: slide

# Configuring the limits and breaks of the continuous y axis
p5 <- 
  p4 +
  scale_y_continuous(
    limits = c(0, 50),
    breaks = seq(0, 50, by = 10)
  )
p5
```

## Step 6: X-Axis Scale

```{r}
#| output-location: slide

# Configuring the limits and breaks of the continuous x axis
p6 <- 
  p5 +
  scale_x_continuous(
    limits = c(1, 7),
    breaks = seq(1, 7)
  )
p6
```

## Step 7: Color Scale

```{r}
#| output-location: slide

# Manually specifying the color values and labels
p7 <- 
  p6 +
  scale_color_manual(
    breaks = c("r", "f", "4"),
    labels = c(
      "4" = "Four-Wheel",
      "f" = "Front-Wheel",
      "r" = "Rear-Wheel"
    ),
    values = c(
      "4" = "salmon",
      "f" = "cornflowerblue",
      "r" = "seagreen3"
    )
  )
p7
```

## Step 8: Shape Scale

```{r}
#| output-location: slide

# Matching the shape scale to the color scale for accessibility
p8 <- 
  p7 +
  scale_shape_discrete(
    breaks = c("r", "f", "4"),
    labels = c(
      "4" = "Four-Wheel",
      "f" = "Front-Wheel",
      "r" = "Rear-Wheel"
    )
  )
p8
```

## Putting It All Together

```{r}
#| output-location: slide

ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_jitter(size = 3, aes(color = drv, shape = drv)) + 
  geom_smooth(method = "lm", color = "black") +
  labs(
    x = "Engine displacement (L)",
    y = "Highway fuel economy (mpg)",
    color = "Drivetrain",
    shape = "Drivetrain",
    title = "Fuel efficiency generally decreases with engine size",
    subtitle = "Rear-wheel drive vehicles are often an exception due to sports cars",
    caption = "Data from fueleconomy.gov"
  ) +
  scale_y_continuous(
    limits = c(0, 50),
    breaks = seq(0, 50, by = 10)
  ) +
  scale_x_continuous(
    limits = c(1, 7),
    breaks = seq(1, 7, by = 1)
  ) +
  scale_color_manual(
    breaks = c("r", "f", "4"),
    labels = c(
      "4" = "Four-Wheel",
      "f" = "Front-Wheel",
      "r" = "Rear-Wheel"
    ),
    values = c(
      "4" = "salmon",
      "f" = "cornflowerblue",
      "r" = "seagreen3"
    )
  ) +
  scale_shape_discrete(
    breaks = c("r", "f", "4"),
    labels = c(
      "4" = "Four-Wheel",
      "f" = "Front-Wheel",
      "r" = "Rear-Wheel"
    )
  )
```


