---
title: "Week 06: Communicate"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [06a] Code Style / Code Chunks

### Topics
- Style your code to improve organization and clarity
- Manage the behavior of individual code chunks in Quarto

### Readings
- R4DS (2E) Chapter 4: [Code Style](https://r4ds.hadley.nz/workflow-style.html){target="_blank"}
- R4DS (2E) Section 28.5: [Code Chunks](https://r4ds.hadley.nz/quarto.html#code-chunks){target="_blank"}
- R4DS (2E) Section 28.6: [Figures](https://r4ds.hadley.nz/quarto.html#sec-figures){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[06a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Code Style

The following code works, but it violates several principles of the tidyverse style guide. Using consistent code style makes collaboration easier and smoother. Rewrite this code to improve its style. 

* Use a consistent snake case style for the object name.
* Put spaces around the relational operators, arithmetic operators, and after commas.
* Follow each pipe with a line break and indent appropriately.

```{r}
#| eval: false
Clean_Sleep_Data<-msleep|>filter(order=="Carnivora",sleep_total>10)|>mutate(sleep_ratio=sleep_rem/sleep_total)|>select(name,sleep_ratio)|>arrange(desc(sleep_ratio))
```

{{< dstart summary="Answer key" >}}

```{r}
#| eval: false
clean_sleep_data <- 
  msleep |> 
  filter(order == "Carnivora", sleep_total > 10) |> 
  mutate(sleep_ratio = sleep_rem / sleep_total) |> 
  select(name, sleep_ratio) |> 
  arrange(desc(sleep_ratio))
```

{{< dstop >}}

#### 2) Chunk Options

Create a new code chunk that generates a simple plot (e.g., `plot(1:10)`). Use hash pipes (`#|`) to configure the chunk options. You can combine multiple options by putting each on a new line.

* Give the chunk a unique label called `simple-plot`.
* Run the code without showing it in the final document.
* Prevent any messages or warnings from rendering.
* Add a figure caption that reads "A simple plot of numbers.".

{{< dstart summary="Answer key" >}}

```markdown
#| label: simple-plot
#| echo: false
#| message: false
#| warning: false
#| fig-cap: "A simple plot of numbers."

plot(1:10)
```

{{< dstop >}}

## [06b] Table Anatomy / Table Styling

### Topics
- Understand the structural anatomy of a `gt` table
- Format and style table data for clear communication

### Readings
- gt Vignette: [Introduction to Creating gt Tables](https://gt.rstudio.com/articles/gt.html){target="_blank"}
- gt Vignette: [Creating Summary Lines](https://gt.rstudio.com/articles/creating-summary-lines.html){target="_blank"}
- gt Vignette: [Case Study--gtcars](https://gt.rstudio.com/articles/case-study-gtcars.html){target="_blank"}


### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[06b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Table Structure

Using the `msleep` dataset (from the {ggplot2} package), create a base display table using the `gt` package.

* First, filter the data to only include animals where the `vore` is "carni", and select the `name`, `sleep_total`, and `sleep_rem` columns.
* Pass this data to `gt()` to create the base table.
* Use `tab_header()` to add a title ("Carnivore Sleep Patterns") and a subtitle ("Total and REM sleep").
* Use `cols_label()` to rename the columns to "Animal", "Total Sleep", and "REM Sleep".

{{< dstart summary="Answer key" >}}

```{r}
#| message: false
library(tidyverse)
library(gt)

msleep |> 
  filter(vore == "carni") |> 
  select(name, sleep_total, sleep_rem) |> 
  gt() |> 
  tab_header(
    title = md("**Carnivore Sleep Patterns**"),
    subtitle = "Total and REM sleep"
  ) |> 
  cols_label(
    name = "Animal",
    sleep_total = "Total Sleep",
    sleep_rem = "REM Sleep"
  )
```

{{< dstop >}}

#### 2) Spanners and Styling

Modify your table from question 1 to add grouping and formatting options.

* Add a spanner using `tab_spanner()` with the label "Hours" over the `sleep_total` and `sleep_rem` columns.
* Center the alignment of the data columns using `cols_align()`.
* Format the sleep values to only show one decimal place using `fmt_number()`.

{{< dstart summary="Answer key" >}}

```{r}
msleep |> 
  filter(vore == "carni") |> 
  select(name, sleep_total, sleep_rem) |> 
  gt() |> 
  tab_header(
    title = md("**Carnivore Sleep Patterns**"),
    subtitle = "Total and REM sleep"
  ) |> 
  cols_label(
    name = "Animal",
    sleep_total = "Total Sleep",
    sleep_rem = "REM Sleep"
  ) |> 
  tab_spanner(
    label = "Hours",
    columns = c(sleep_total, sleep_rem)
  ) |> 
  cols_align(
    align = "center",
    columns = c(sleep_total, sleep_rem)
  ) |> 
  fmt_number(
    columns = c(sleep_total, sleep_rem),
    decimals = 1
  )
```

{{< dstop >}}

## [06c] Plot Labels / Plot Scales

### Topics
- Customize the axis and plot labels in a figure
- Customize the appearance of scales in a figure

### Readings
- R4DS (2E) Section 11.2: [Labels](https://r4ds.hadley.nz/communication.html#labels){target="_blank"}
- R4DS (2E) Section 11.4: [Scales](https://r4ds.hadley.nz/communication.html#scales){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[06c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Customizing Labels

Create a scatterplot using the `iris` dataset (from the {datasets} package).

* Map **Sepal Length (`Sepal.Length`)** to the x-axis and **Sepal Width (`Sepal.Width`)** to the y-axis.
* Map **Species (`Species`)** to the color aesthetic.
* Use `labs()` to add a clear title, x-axis label, y-axis label, and legend title.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false
library(tidyverse)

iris |> 
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  labs(
    title = "Sepal dimensions vary by iris species",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)",
    color = "Iris Species"
  )
```

{{< dstop >}}

#### 2) Customizing Scales

Modify your plot from question 1 to customize its scales.

* Use `scale_y_continuous()` to set the limits from 2 to 5, and add breaks every 0.5 units.
* Use `scale_color_manual()` to assign specific colors ("darkorange", "purple", and "cyan4") to the three iris species.

{{< dstart summary="Answer key" >}}

```{r}
iris |> 
  ggplot(aes(x = Sepal.Length, y = Sepal.Width, color = Species)) +
  geom_point(size = 3) +
  labs(
    title = "Sepal dimensions vary by iris species",
    x = "Sepal Length (cm)",
    y = "Sepal Width (cm)",
    color = "Iris Species"
  ) +
  scale_y_continuous(
    limits = c(2, 5),
    breaks = seq(2, 5, by = 0.5)
  ) +
  scale_color_manual(
    values = c(
      "setosa" = "darkorange",
      "versicolor" = "purple",
      "virginica" = "cyan4"
    )
  )
```

{{< dstop >}}

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../05/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 05</a>

  <a href="../07/index.html" class="btn btn-primary" role="button">Week 07&nbsp;&nbsp;&raquo;</a>
</div>
```
