---
lecture: "11a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 24))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Calculate summary statistics to describe an entire dataset

- Extract extreme rows using the slice family of functions

- Apply these operations to distinct subgroups within your data
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Basic Summarize

## Summarize {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Question:** *What does the data look like overall?*

-   Although we store data about many observations, we often want to [summarize across]{.b .blue} them.
    -   This is like [folding]{.b .green} the tibble down to [one row]{.b .green}.

::: {.fragment .mt1}
-   We have seen functions that summarize vectors:
    -   `length()`, `sum()`, `min()`, `max()`
    -   `mean()`, `median()`, `sd()`, `var()`
:::

::: {.fragment .mt1}
-   `summarize()` lets us use them on tibbles.
    -   It works very similarly to `mutate()`.
    -   It always creates a new tibble as output.
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/fold.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Preparing the Data

::: {.fsmaller}
First, let's create a small example dataset of sales to practice with.
:::

```{r}
library(tidyverse)
sales <- tibble(
  customer = c(1, 2, 3, 1, 3),
  store = c("A", "A", "A", "B", "B"),
  quantity = c(25, 20, 16, 10, 5),
  revenue = c(685, 590, 392, 185, 123)
)
sales
```

## Overall Averages

::: {.fsmaller}
**Question:** *What is the average transaction size and value?*

We can use `summarize()` to calculate the mean quantity and revenue across all observations in the dataset.
:::

```{r}
sales |> 
  summarize(
    avg_quantity = mean(quantity),
    avg_revenue = mean(revenue)
  )
```

## The summary() Pitfall

::: {.fsmaller}
[Warning:]{.b .dark-red} Be careful not to accidentally use base R's `summary()` function when you mean to use `dplyr::summarize()`. It will ignore your arguments!
:::

```{r}
sales |> 
  summary(
    avg_quantity = mean(quantity),
    avg_revenue = mean(revenue)
  )
```

## Counting Functions

::: {.fsmaller}
**Question:** *How much data do we actually have?*

We can use special functions like `n()` to count total rows, and `n_distinct()` to count unique values.
:::

```{r}
sales |> 
  summarize(
    n_sales = n(),
    n_customers = n_distinct(customer),
    n_stores = n_distinct(store)
  )
```

# Grouped Summarize

## Grouped Summarize {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   **Question:** *How do different categories compare to one another?*

-   We can summarize a tibble [by group]{.b .blue}.
    -   This is like [folding]{.b .green} the tibble multiple times.
    -   We will fold down to [one row per group]{.b .green}.

::: {.fragment .mt1}
-   We just add the `.by` argument to `summarize()`.
    -   We can group by one or many variables.
    -   With many, we group by their combinations.
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/origami.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Summarizing by One Group

::: {.fsmaller}
**Question:** *Which store is performing better?*

Let's see how the sales perform on a per-store basis by adding `.by`.
:::

```{r}
sales |> 
  summarize(
    .by = store,
    customers = n_distinct(customer),
    total_quantity = sum(quantity),
    total_revenue = sum(revenue),
    avg_quantity = mean(quantity),
    avg_revenue = mean(revenue)
  )
```

## Summarizing by Multiple Groups

::: {.fsmaller}
**Question:** *How much did each customer purchase at each store?*

Now let's group by customer and store combinations.
:::

```{r}
sales |> 
  summarize(
    .by = c(customer, store),
    n_visits = n(),
    total_quantity = sum(quantity),
    total_revenue = sum(revenue)
  )
```

## Real-World Data

::: {.fsmaller}
Let's apply these concepts to a larger, more realistic dataset.
:::

```{r}
library(nycflights13)

delays <- 
  flights |> 
  select(month, day, carrier, flight, dep_delay)

glimpse(delays)
```

## Average Delay by Carrier

::: {.fsmaller}
**Question:** *What is the average departure delay for each airline carrier?*
:::

```{r}
delays |> 
  summarize(
    .by = carrier,
    m_delay = mean(dep_delay, na.rm = TRUE)
  )
```

# Slice 

## The Slice Functions {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
**Question:** *Who are the extreme winners and losers?*

We often want to extract specific rows.

::: {.fragment .mt1}
- The `slice` functions do exactly this:
    - `slice_min()` extracts rows based on their lowest values.
    - `slice_max()` extracts rows based on their highest values.

:::

::: {.fragment .mt1}
- We just tell it which column to look at using `order_by` and how many rows to return using `n`.
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/scissors.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Best Average Performance

::: {.fsmaller}
**Question:** *Which 3 carriers have the lowest average departure delay?*

We can calculate the summary and then slice the results to get our answer.
:::

```{r}
delays |> 
  summarize(
    .by = carrier, 
    m_delay = mean(dep_delay, na.rm = TRUE)
  ) |> 
  slice_min(n = 3, order_by = m_delay)
```

## Peak Volume

::: {.fsmaller}
**Question:** *Which 2 dates had the most NYC flights (in 2013)?*
:::

```{r}
delays |> 
  summarize(
    .by = c(month, day), 
    n_flights = n()
  ) |> 
  slice_max(n = 2, order_by = n_flights)
```

# Grouped Slice

## Grouping inside Slice {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- **Question:** *What is the extreme value WITHIN each category?*

- We do not always have to summarize before we slice.
- We can slice the raw data directly, and we can apply the `by` argument to `slice_max()` and `slice_min()`.
    - **Note:** Notice we use `by` here, not `.by`!

::: {.fragment .mt1}
- Instead of returning the top 1 row overall, this returns the top 1 row **for each group**.
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/layers.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## The Worst Flights per Carrier

::: {.fsmaller}
**Question:** *What was the single most delayed flight for each airline?*
:::

```{r}
delays |> 
  slice_max(
    n = 1, 
    order_by = dep_delay, 
    by = carrier
  )
```

# Grouped Mutate

## Grouped Mutate {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
- **Question:** *How does a row compare to its peers?*

- We have already learned that `mutate()` creates new columns or alters existing ones.
- What happens if we combine `mutate()` with the `.by` argument?

::: {.fragment .mt1}
- The new column is calculated **within each group** but keeps the original structure of the dataset.
- This is useful for calculating deviations from group means or standardizing scores within categories.
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/gears.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Comparing to Group Means

::: {.fsmaller}
**Question:** *Was this specific flight worse than the carrier's usual performance?*
:::

```{r}
delays |> 
  mutate(
    .by = carrier,
    carrier_avg = mean(dep_delay, na.rm = TRUE),
    relative_delay = dep_delay - carrier_avg
  )
```
