---
lecture: "11b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

options(
  tibble.print_max = 6,
  tibble.print_min = 6
)

set_theme(theme_gray(base_size = 30))

set.seed(1234)
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Identify when "wide" or "long" format is needed

- Reshape wide data into long data using `pivot_longer()`

- Reshape long data into wide data using `pivot_wider()`

:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Pivot Longer

## Pivot Longer {.fs70}

::: {.columns .pv4}

::: {.column width="60%"}
- Tidy data can have many different shapes.
    + [Long format]{.b .blue} is useful for group summaries and most plots.
    + Wide format is useful for correlations and calculating differences.
  
::: {.fragment .mt1}
- `pivot_longer()` adds rows from columns.
    + `cols`: select the columns we want to pivot.
    + `names_to`: create a *new* column where the old names will go (provide as a "string").
    + `values_to`: create a *new* column where the old values will go (provide as a "string").
  
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/arrows_down.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Preparing the Data

::: {.fs80}
First, let's simulate a dataset in wide format.
:::

```{r}
gradebook <- tibble(
  student = 1:10,
  test1 = round(rnorm(n = 10, mean = 77, sd = 9)),
  test2 = round(rnorm(n = 10, mean = 75, sd = 8)),
  test3 = round(rnorm(n = 10, mean = 80, sd = 7))
)
gradebook
```

## The Problem with Wide Data

::: {.fs80}
**Question:** *How do we plot the trend of grades across tests for each student?*
:::

```{r}
glimpse(gradebook)
```

```{r}
#| eval: false
#| error: true

# We cannot map test and grade to x and y because they aren't columns!
ggplot(gradebook, aes(x = ???, y = ???, group = student)) + ...
```

:::{.fragment}
```{r}
#| eval: false
#| error: true

# What we need are test and grade columns
ggplot(gradebook, aes(x = test, y = grade, group = student)) + ...
```
:::

## Basic pivot_longer()

```{r}
gradebook2 <- gradebook |> 
  pivot_longer(
    cols = c(test1, test2, test3), 
    names_to = "test", 
    values_to = "grade"
  ) 
gradebook2
```

## Plotting Long Data

::: {.fs80}
Now that we have the data in long format, plotting is easy!
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

ggplot(
  gradebook2, 
  aes(
    x = test, 
    y = grade, 
    group = student
  )
) + 
  geom_line() + 
  geom_point()
```

## Convenience Options

```{r}
gradebook |> 
  pivot_longer(
    cols = starts_with("test"), # tidy select cols
    names_to = "test", 
    values_to = "grade",
    names_prefix = "test", # remove prefix from names
    names_transform = parse_number # turn names into numbers
  )
```

## Numeric Column Names {.fs90}

:::{.fs80}
Datasets with numeric columns names can cause issues...
:::

```{r}
data("table4a", package = "tidyr")
table4a
```
::: {.columns}
::: {.column .fragment}
```{r}
#| error: true

table4a |> select(1999, 2000)
```
:::
::: {.column .fragment}
```{r}
table4a |> select(`1999`, `2000`)
```
:::
:::

## Pivoting Such Columns

```{r}
table4a |> 
  pivot_longer(
    cols = c(`1999`, `2000`),
    names_to = "year",
    values_to = "cases"
  )
```

# Pivot Wider

## Pivot Wider {.fs70}

::: {.columns .pv4}

::: {.column width="60%"}
- Tidy data can have many different shapes.
    + Long format is useful for group summaries and most plots.
    + [Wide format]{.b .blue} is useful for correlations and calculating differences.
  
::: {.fragment .mt1}
- `pivot_wider()` adds columns from rows.
    + `names_from`: select the columns containing the new column names.
    + `values_from`: select the columns containing the data values.
  
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/arrows_right.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Preparing the Data

::: {.fs80}
First, let's simulate a dataset in long format.
:::

```{r}
diary_data <- tibble(
    participant = rep(1:3, each = 7, times = 2),
    day = rep(1:7, each = 2, times = 3),
    scale = rep(c("mood", "sleep"), times = 3*7),
    score = round(rnorm(n = 3*7*2), digits = 1)
  )
diary_data
```

## The Problem with Long Data

::: {.fs80}
**Question:** *Does daily mood correlate with daily sleep?*
:::

```{r}
glimpse(diary_data)
```

```{r}
#| eval: false
#| error: true

# We cannot map mood and sleep to x and y because they aren't columns!
ggplot(diary_data, aes(x = ???, y = ???)) + ...
```

:::{.fragment}
```{r}
#| eval: false
#| error: true

# What we need are mood and sleep columns
ggplot(diary_data, aes(x = mood, y = sleep)) + ...
```
:::

## Basic pivot_wider()

```{r}
diary_data |>
  pivot_wider(
    names_from = scale,
    values_from = score
  )
```

## Plotting Wide Data

::: {.fs80}
Now that we have the data in wide format, plotting is easy!
:::

```{r}
#| output-location: column
#| fig-width: 7
#| fig-height: 5

diary_data |> 
  pivot_wider(
    names_from = scale,
    values_from = score
  ) |> 
  ggplot(aes(x = mood, y = sleep)) + 
  geom_point() +
  geom_smooth()
```

## Numerical Column Names

::: {.fs80}
Sometimes pivoting leaves you with column names that are just numbers.
:::

```{r}
diary_data |> 
  pivot_wider(
    names_from = day,
    values_from = score
  )
```

## Avoiding Such Columns

::: {.fs80}
We can use `names_prefix` to make them clearer and easier to work with.
:::

```{r}
diary_data |> 
  pivot_wider(
    names_from = day,
    values_from = score,
    names_prefix = "day_"
  )
```

## Pivoting Multiple Columns {.fs90}

::: {.fs80}
We can also create new columns by combining categories. Here, we generate new column names based on combinations of the `day` and `scale` columns.
:::

```{r}
diary_data |> 
  pivot_wider(
    names_from = c(day, scale),
    values_from = score,
    names_prefix = "day",
    names_sep = "_"
  )
```

## Why Pivot Multiple Columns?

::: {.fs80}
This extreme wide format collapses every observation for a single participant onto a single row. This makes calculating change scores across time easy!
:::

```{r}
diary_data |> 
  pivot_wider(
    names_from = c(day, scale),
    values_from = score,
    names_prefix = "day",
    names_sep = "_"
  ) |> 
  mutate(mood_change = day7_mood - day1_mood) |> 
  select(participant, day1_mood, day7_mood, mood_change)
```

## Data with Duplicate Identifiers

::: {.fs80}
Let's create a dataset where two participants share the same first name.
:::

```{r}
records <- tibble(
    fname = rep(c("John", "Alice", "Buster", "John", "Zenia"), each = 2),
    variable = rep(c("grad_year", "cum_gpa"), times = 5),
    value = c(1990, 3.01, 1993, 3.32, 1997, 2.09, 1994, 3.28, 1991, 2.64)
  )
records
```

## The Duplicate Row Problem

::: {.fs80}
Pivoting assumes it can uniquely identify every observation. Because we have two "John" observations, it jams them both into a list.
:::

```{r}
records |> 
  pivot_wider(
    names_from = variable, 
    values_from = value
  )
```

## Resolving Duplicate Rows

::: {.fs80}
We fix this by ensuring our rows are uniquely identified before we pivot (e.g., by adding the last name initials to distinguish "John A" from "John Z").
:::

```{r}
records |> 
  mutate(linitial = rep(c("A", "S", "K", "Z", "I"), each = 2)) |> 
  pivot_wider(
    names_from = variable,
    values_from = value
  )
```
