---
title: "Week 11: Reshape"
execute:
  echo: true
  eval: true
---

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

knitr::opts_chunk$set(collapse = TRUE)
```

## [11a] Summarize / Group

### Topics
- Calculate summary statistics to describe an entire dataset
- Extract extreme rows using the slice family of functions
- Apply these operations to distinct subgroups within your data

### Readings
- R4DS (2E) Section 3.5: [Groups](https://r4ds.hadley.nz/data-transform#groups)

### Slides
```{=html}
<iframe class="slide-deck" src="a_Slides.html" title="[11a] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

### Practice

#### 1) Summarizing Data

Using the `mpg` dataset (included with the tidyverse), calculate summary statistics for vehicle fuel efficiency.

* First, use `summarize()` to calculate the overall average city mileage (`cty`) and highway mileage (`hwy`). Name these new columns `avg_cty` and `avg_hwy`.

* Next, modify your code to calculate these same averages, but grouped by vehicle `class` using the `.by` argument.

* Add a third summary column called `n_cars` that counts the number of observations in each class using `n()`.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)

mpg |> 
  summarize(
    avg_cty = mean(cty),
    avg_hwy = mean(hwy),
    n_cars = n(),
    .by = class
  )
```

{{< dstop >}}

#### 2) Extracting Extreme Rows

Using the `msleep` dataset (also included with the tidyverse), find the animals that sleep the most.

* First, select just the `name`, `vore`, and `sleep_total` columns to make the output easier to read.

* Use `slice_max()` to find the 3 animals with the highest total sleep (`sleep_total`) across the entire dataset.

* Now, modify your `slice_max()` code to find the single animal with the highest total sleep *within each feeding type* (`vore`). Remember to use the `by` argument instead of `.by` inside slice functions!

{{< dstart summary="Answer key" >}}

```{r}
# Overall top 3 sleepers
msleep |> 
  select(name, vore, sleep_total) |> 
  slice_max(n = 3, order_by = sleep_total)

# Top sleeper per feeding type
msleep |> 
  select(name, vore, sleep_total) |> 
  slice_max(n = 1, order_by = sleep_total, by = vore)
```

{{< dstop >}}

#### 3) Grouped Transformations

Let's calculate how each animal's sleep compares to others with the exact same diet using a grouped mutate.

* Start with the `msleep` dataset and use `mutate()` with the `.by` argument to group by `vore`.

* Calculate a new column called `mean_sleep` that represents the average `sleep_total` for that specific diet group. *(Note: you will need to add `na.rm = TRUE` inside your mean function since some `vore` values are missing).*

* Calculate a second new column called `diff_sleep` that subtracts the group's `mean_sleep` from the animal's actual `sleep_total`.

* Finally, select the `name`, `vore`, `sleep_total`, `mean_sleep`, and `diff_sleep` columns to view your work.

{{< dstart summary="Answer key" >}}

```{r}
msleep |> 
  mutate(
    .by = vore,
    mean_sleep = mean(sleep_total, na.rm = TRUE),
    diff_sleep = sleep_total - mean_sleep
  ) |> 
  select(name, vore, sleep_total, mean_sleep, diff_sleep)
```

{{< dstop >}}

## [11b] Lengthen / Widen

### Topics
- Identify when "wide" or "long" format is needed
- Reshape wide data into long data using `pivot_longer()`
- Reshape long data into wide data using `pivot_wider()`

### Readings
- R4DS (2E) Section 5.3: [Lengthening data](https://r4ds.hadley.nz/data-tidy.html#sec-pivoting){target="_blank"}
- R4DS (2E) Section 5.4: [Widening data](https://r4ds.hadley.nz/data-tidy.html#widening-data){target="_blank"}

### Slides
```{=html}
<iframe class="slide-deck" src="b_Slides.html" title="[11b] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice

#### 1) Lengthening Data

Using the `billboard` dataset (included with the tidyr package), reshape the song ranking data from wide to long format.

* The dataset records the chart rank of songs over 76 weeks in columns `wk1` through `wk76`.
* Use `pivot_longer()` to collapse these week columns. Use the `starts_with()` helper to select them efficiently.
* Name the new category column `"week"` and the new value column `"rank"`.
* Use the `names_prefix` and `names_transform = parse_number` arguments to strip the "wk" text and convert the week column into true numbers.

{{< dstart summary="Answer key" >}}

```{r}
#| message: false

library(tidyverse)
data("billboard", package = "tidyr")

billboard |> 
  pivot_longer(
    cols = starts_with("wk"),
    names_to = "week",
    values_to = "rank",
    names_prefix = "wk",
    names_transform = parse_number
  )
```

{{< dstop >}}

#### 2) Widening Data

Using the `us_rent_income` dataset (included with the tidyr package), reshape the state data from long to wide format so that rent and income are presented side-by-side.

* First, use `select()` to keep only the `NAME`, `variable`, and `estimate` columns. This prevents issues with duplicate rows caused by other identifiers.
* Use `pivot_wider()` to create separate columns for each category stored in the `variable` column.
* Pull the actual numerical values from the `estimate` column.

{{< dstart summary="Answer key" >}}

```{r}
data("us_rent_income", package = "tidyr")

us_rent_income |> 
  select(NAME, variable, estimate) |> 
  pivot_wider(
    names_from = variable,
    values_from = estimate
  )
```

{{< dstop >}}

#### 3) Pivoting Multiple Columns

Pivoting multiple columns simultaneously is a powerful tool for longitudinal or repeated-measures data. Let's practice this on a simulated weather dataset.

* First, copy the code below to create the `weather` dataset.
* Use `pivot_wider()` to flatten this dataset so each station occupies only a single row.
* Provide both category columns to the `names_from` argument to combine the season and measurement types.
* Separate the combined names using a custom character with the `names_sep` argument.

```{r}
weather <- tibble(
  station = rep(c("Station1", "Station2"), each = 4),
  season = rep(c("summer", "summer", "winter", "winter"), times = 2),
  measure = rep(c("temp", "precip"), times = 4),
  value = c(85, 2.1, 32, 5.5, 90, 1.8, 40, 6.2)
)
```


{{< dstart summary="Answer key" >}}

```{r}
weather |> 
  pivot_wider(
    names_from = c(season, measure),
    values_from = value,
    names_sep = "_"
  )
```

{{< dstop >}}

## [11c] Link / Join Data

### Topics
- A
- B
- C

### Readings
- A
- B
- C

### Slides
```{=html}
<iframe class="slide-deck" src="c_Slides.html" title="[11c] Slideshow">
</iframe>
```

::: {.fsmaller}
\*Note that you can click the three-line (hamburger) icon on the bottom-left of the slides to access a navigation menu. You can click inside the slides region and press the left and right arrow keys on your keyboard to advance and reverse the slides and animations.
:::

### Practice
TBA

<!-- Bottom Nav -->

```{=html}
<div style="display: flex; justify-content: space-between; margin-top: 3rem;">
  <a href="../10/index.html" class="btn btn-outline-primary" role="button">&laquo;&nbsp;&nbsp;Week 10</a>

  <a href="../12/index.html" class="btn btn-primary" role="button">Week 12&nbsp;&nbsp;&raquo;</a>
</div>
```
