---
title: "Week 02 Assignment"
author: "YOUR NAME HERE"
format: html
embed-resources: true
---

## Instructions

1.  **Set up:** Change "YOUR NAME HERE" in the header above to your actual name.
2.  **Answer:** For text questions, replace the `ANSWER` text inside the block quotes with your own response. For coding questions, write your code in the empty chunks provided.
3.  **Render:** When you are finished, click the **Render** button.
4.  **Submit:** Upload the resulting HTML file to Canvas.

> **Warning:** If any of your R code chunks contain errors, the document will **fail to render**. 
>
> * Test your code in the Console first to ensure it runs without error.
> * Remove any failed code attempts from the document before rendering.

## Question 1 (RStudio)

**a) Why are we using R in this course and not some other programming language?**

> ANSWER: (Write your 2-4 sentence answer here)

**b) What are R and RStudio? How are they related and how are they different?**
*Please explain in your own words (i.e., don't just repeat the example from lecture).*

> ANSWER: 

## Question 2 (Console)

**a) Use R as a calculator to answer the following question: what is negative one plus two-thirds?**

```{r}
# Write your code below this line

```

**b) Use R as a calculator to answer the following question: what is the product of 1.234 and five-raised-to-the-sixth-power?**

```{r}
# Write your code below this line

```

## Question 3 (Scripts)

**a) What are the differences between running code in the console, in a script, and in a Quarto document? When might an R user want to use each of these approaches?**

> ANSWER: 

**b) What are code comments and when might they be useful? How are code comments in a script different from markdown in a Quarto document?**

> ANSWER: 

## Question 4 (Projects)

**a) What is the purpose of a Project in RStudio? Why do I recommend creating a project rather than simply saving all your course stuff in a normal computer folder?**

> ANSWER: 

**b) What are some benefits of having a consistent naming convention for the files in your project? What kinds of tasks might be easier if you stick with a convention?**

> ANSWER: 

## Question 5 (Quarto)

**a) What is the difference between a .qmd file and a .html file created by Quarto? When would you send the .qmd file to a collaborator and when would you send the .html file?**

> ANSWER: 

**b) Quarto lets you include many code chunks in the same document. When would you want to put all your code in a single chunk and when would you want to break it up across multiple chunks?**

> ANSWER: 

## Question 6 (Markdown)

**a) Markdown Headers**
*The instructions originally asked you to use headers to divide the document. Since this file already has headers, practice by creating a Level 3 header below this line that says "My Verbatim Code".*

**b) Verbatim Code**
*The verbatim code format in markdown is really useful when we want to talk about code in text without actually running that code. For instance, I can say that division in R is accomplished using the `/` symbol as in `1 / 2`.*

*Practice using verbatim code by re-writing your code from Question 2a in the block quote below (do not use a code chunk).*

> ANSWER: 

---

✅ **Done!**
Please save and render your document to an HTML file. Check it over to make sure it is complete and uploaded to Canvas.
