---
lecture: "11a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)

set_theme(theme_gray(base_size = 24))
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Learn how to summarize whole datasets

- Learn how to summarize data by distinct groups

:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Summarize

## Summarize {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   Although we store data about many observations...
-   ...we often want to [summarize across]{.b .blue} observations
    -   This is like [folding]{.b .green} the tibble down to [one row]{.b .green}

::: {.fragment .mt1}
-   We have seen functions that summarize vectors
    -   `length()`, `sum()`, `min()`, `max()`
    -   `mean()`, `median()`, `sd()`, `var()`
:::

::: {.fragment .mt1}
-   `summarize()` lets us use them on tibbles
    -   It works very similarly to `mutate()`
    -   It always creates a tibble as output
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/fold.json" trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## 1. Preparing the Data

::: {.fsmaller}
First, let's create a small example dataset of sales to practice our summary functions.
:::

```{r}
#| echo: true

library(tidyverse)

sales <- 
  tibble(
    customer = c(1, 2, 3, 1, 3),
    store = c("A", "A", "A", "B", "B"),
    items = c(25, 20, 16, 10, 5),
    spent = c(685, 590, 392, 185, 123)
  )

sales
```

## 2. Basic Summarize

::: {.fsmaller}
We can use `summarize()` to calculate the average items and spent across all observations in the dataset.
:::

```{r}
#| echo: true

sales |> 
  summarize(
    avg_items = mean(items),
    avg_spent = mean(spent)
  )
```

## 3. The `summary()` Pitfall

::: {.fsmaller}
**Warning:** Be careful not to accidentally use base R's `summary()` function when you mean to use `dplyr::summarize()`. The former will not return a tidy tibble.
:::

```{r}
#| echo: true

sales |> 
  summary(
    avg_items = mean(items),
    avg_spent = mean(spent)
  )
```

## 4. Counting Functions

::: {.fsmaller}
We can also use special functions like `n()` to count total rows, and `n_distinct()` to count unique values.
:::

```{r}
#| echo: true

sales |> 
  summarize(
    n_sales = n(),
    n_customers = n_distinct(customer),
    n_stores = n_distinct(store)
  )
```

# Grouped Summarize

## Grouped Summarize {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   We can also summarize a tibble [by group]{.b .blue}
    -   This is like [folding]{.b .green} the tibble multiple times
    -   Specifically, we fold down to [one row per group]{.b .green}

::: {.fragment .mt1}
-   We just add the `.by` argument to `summarize()`
    -   We can group by one or many variables
    -   With many, we group by their combinations
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/origami.json" trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## 1. Summarizing by One Group

::: {.fsmaller}
Let's see how the sales perform on a per-store basis by adding the `.by` argument.
:::

```{r}
#| echo: true

sales |> 
  summarize(
    customers = n_distinct(customer),
    items_sold = sum(items),
    total_sales = sum(spent),
    avg_items = mean(items),
    avg_spent = mean(spent),
    .by = store
  )
```

## 2. Summarizing by Multiple Groups

::: {.fsmaller}
If we pass multiple variables to `.by`, it summarizes every unique combination of those groups.
:::

```{r}
#| echo: true

sales |> 
  summarize(
    n_visits = n(),
    items_bought = sum(items),
    total_spent = sum(spent),
    .by = c(customer, store)
  )
```

## 3. Real-World Data

::: {.fsmaller}
Let's apply these concepts to a larger, more realistic dataset using `nycflights13`.
:::

```{r}
#| echo: true

library(nycflights13)

flights |> 
  select(month, day, carrier, flight, dep_delay) |> 
  head()
```

## 4. Flight Delays by Carrier

::: {.fsmaller}
We can find the carrier with the lowest average departure delays by grouping and sorting. Note that we need `na.rm = TRUE` to ignore missing values.
:::

```{r}
#| echo: true

flights |> 
  summarize(
    m_delay = mean(dep_delay, na.rm = TRUE),
    .by = carrier
  ) |> 
  arrange(m_delay) |> 
  head()
```

## 5. Busiest Days of the Year

::: {.fsmaller}
By grouping by both `month` and `day`, we get 365 rows instead of just 31 (which would happen if we only grouped by `day`).
:::

```{r}
#| echo: true

flights |> 
  summarize(
    n_flights = n(),
    .by = c(month, day)
  ) |> 
  arrange(desc(n_flights)) |> 
  head()
```
