---
lecture: "10b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))

library(tidyverse)
library(easystats)
library(faux)
library(patchwork)

set_theme(theme_gray(base_size = 24))

rsim <- function(r) {
  rnorm_multi(n = 200, vars = 2, mu = 0, sd = 1, r = r)
}
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Run correlation tests for continuous variables

- Run Chi-Square tests for categorical variables
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Moving Beyond Groups {.smaller}

Last lecture, we compared a continuous variable across different categories

::: {.fragment .mt1}
- What if we want to test the association of [two continuous variables]{.b .blue}?
    + *e.g., As a penguin's flipper gets longer, does its body mass also increase?*
    + We can use **Correlation** to test this
    
:::

::: {.fragment .mt1}
- What if we want to test the association of [two categorical variables]{.b .green}?
    + *e.g., Does the species of a penguin depend on which island it lives on?*
    + We can use **Chi-Square** to test this
    
:::

# Correlations

## Interpreting Correlations {.smaller}

- A correlation (parameter = $\rho$, statistic = $r$) quantifies the **direction** and **strength** of the *linear* relationship between two variables (how closely the points form a straight line)
    
- [Hypotheses]{.b .dark-red}:
    + **$H_0$**: $\rho = 0$ (There is no linear relationship in the population)
    + **$H_1$**: $\rho \neq 0$ (There is a linear relationship in the population)

- [Direction]{.b .blue}: do the variables change together or opposing?
    + $r > 0$: the variables change together
    + $r < 0$: the variables change in opposition

- [Strength]{.b .green}: how well does one variable predict the other?
    + $r \to 0$: no linear relationship (but maybe a non-linear one)
    + $r \to \pm 1$: perfect (positive or negative) linear relationship

## Positive linear

```{r}
#| echo: false

p1 <- 
  ggplot(rsim(0.25), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ 0.25", x = "x1", y = "y") + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
p2 <-   
  ggplot(rsim(0.50), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ 0.50", x = "x2", y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
p3 <- 
  ggplot(rsim(0.75), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ 0.75", x = "x3", y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

p1 | p2 | p3
```

## Negative linear

```{r}
#| echo: false

p4 <- 
  ggplot(rsim(-0.25), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ −0.25", x = "x4", y = "y") + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
p5 <-   
  ggplot(rsim(-0.50), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ −0.50", x = "x5", y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
p6 <- 
  ggplot(rsim(-0.75), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ −0.75", x = "x6", y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )

p4 | p5 | p6
```

## No linear / Nonlinear

```{r, echo=FALSE}
set.seed(2021)
p7 <- 
  ggplot(rsim(0), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ 0.00", x = "x7", y = "y") + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
p8 <-   
  ggplot(tibble(X1 = runif(200, -5, 5), 
                X2 = -X1^2 + rnorm(200, 0, 3)), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ 0.00", x = "x8", y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
p9 <-   
  ggplot(tibble(X1 = runif(200, -5, 5), 
                X2 = X1^2 + X1^3 + rnorm(200, 0, 10)), aes(x = X1, y = X2)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon", size = 2) +
  labs(title = "r ≈ 0.87", x = "x9", y = NULL) + 
  theme(
    plot.title = element_text(hjust = 0.5),
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )


p7 | p8 | p9
```


## 1. Preparing the Data

```{r}
#| message: false 

library(tidyverse)
library(easystats)

data("penguins", package = "datasets")

dat1 <- penguins |> select(flipper_len, body_mass)

glimpse(dat1)
```


## 2. Exploring our Sample

```{r}
#| warning: false

ggplot(dat1, aes(x = flipper_len, y = body_mass)) + 
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "salmon")
```

## 3. The Correlation Test
```{r}
res1 <- correlation(dat1) |> print()
```

:::{.fsmaller}
**Interpretation:** The correlation ($r$) is 0.87. The significant result (*p* < .001) provides evidence that flipper length and body mass are positively associated in the population: penguins with more (or less) body mass tend to have longer (or shorter) flippers.
:::

## 4. Multiple Variables

```{r}
dat2 <- 
  penguins |> 
  select(bill_len, bill_dep, flipper_len, body_mass)

res2 <- correlation(dat2, p_adjust = "holm") |> print()
```


## 4. The Correlation Matrix

```{r}
mat2 <- summary(res2) |> print()
```

:::{.fsmaller}
**Interpretation:** The function automatically tests every possible pair of variables and provides adjusted $p$-values for each relationship.
:::

## 5. Visualizing the Matrix

```{r}
plot(mat2)
```

## Interpreting Correlation Strength {.smaller}

Once we know a correlation is significant, how do we evaluate its size?

::: {.fragment .mt1}
- **Cohen's Guidelines (1988):**
    + A common heuristic for interpreting the absolute value of *r*
    
:::

::: {.fragment .mt1}
| Effect Size | Interpretation |
|:---|:---|
| $|r|\approx0.10$ | Small / Weak |
| $|r|\approx0.30$ | Medium / Moderate |
| $|r|\approx0.50$ | Large / Strong |

:::

::: {.fragment .mt1}
- **Context matters!** 
    + An $r\approx0.10$ might be groundbreaking in medicine but useless in physics

:::

# Two Categorical Variables

## Understanding Chi-Square {.smaller}

The **Chi-Square $(\chi^2)$ Test of Independence** looks at categorical relationships

::: {.fragment .mt1}
- **The Core Question:**
    + Does knowing one category tell us something about the other category?
    + *e.g., Are the Species and Island categorical variables associated?*
    
:::

::: {.fragment .mt1}
- **The Hypotheses:**
    + **Null Hypothesis ($H_0$):** The variables are completely independent
        + *e.g., species are randomly distributed across islands*
    + **Alternative Hypothesis ($H_1$):** The variables are associated
        + *e.g., some species are more common on some islands*
    
:::

::: {.fragment .mt1}
- We look at **frequencies** (counts) of each combination of categories
    
:::

## 1. Preparing the Data

```{r}
data("penguins", package = "datasets")

dat3 <- penguins |> select(species, island)

glimpse(dat3)
```

## 2. Exploring our Sample

```{r}
ggplot(dat3, aes(x = island, fill = species)) + 
  geom_bar(position = "fill") + # "fill" makes all bars the same height
  labs(y = "Proportion")
```

## 3. Cross-Tabulation

::: {.fsmaller}
Before we can test these variables, we need to count how many penguins fall into each combination of categories. We use the `xtabs()` function for this

The syntax uses a one-sided formula `~ variable1 + variable2`
:::

::: {.fragment .mt1}
```{r}
tab3 <- dat3 |> table() |> print()
```

::: {.fsmaller}
**Note:** All Gentoo penguins were observed on Biscoe, and all Chinstrap penguins were observed on Dream This suggests a very strong association!
:::
:::

## 4. The Logic of Chi-Square {.smaller}

How does a Chi-Square test actually work?

- It compares what we **observed** in our data to what we would **expect** to see if the two variables were completely independent.
- If species and island were unrelated, penguins would be distributed across the islands purely based on the overall size of each island and the overall population of each species.
- The larger the difference between Observed and Expected, the larger our resulting Chi-Square statistic ($\chi^2$).

```{r}
res3 <- chisq.test(tab3)
```


## 5. Expected Frequencies

::: {.fsmaller}
R can calculate those "Expected" numbers for us, and we can compare them
:::

::: {.columns}
::: {.column}
```{r}
res3$observed
```
:::
::: {.column}
```{r}
res3$expected
```

:::
:::

::: {.fsmaller}
**Note:** These tables are very different! If there were no association, we would expect to find about 45 Gentoo on Dream. But in our sample, we found 0
:::

## 6. The Chi-Square Test

:::{.fsmaller}
Now we look at the formal test results using `model_parameters()`
:::

```{r}
model_parameters(res3)
```

:::{.fsmaller}
**Interpretation:** The highly significant result (*p* < .001) means the distribution of species is dependent on the island. We have strong evidence that this geographic sorting exists in the broader population.
:::

## 7. Effect Size

::: {.fsmaller}
To quantify the *strength* of the association, we can calculate **Cramer's V**
:::

```{r}
effectsize(res3)
```

::: {.fsmaller}
**Interpretation:** Cramer's V ranges from 0 to 1. A value of 0.84 indicates a very strong association between species and island location.
:::

## Interpreting Cramer's V {.smaller}

- **Cohen's Guidelines (Adapted):**
    + Similar to correlation, we can use heuristics to interpret $V$
    + *(Note: Exact thresholds depend on table size, but this is a solid start!)*

::: {.fragment .mt1}
| Effect Size | Interpretation |
|:---|:---|
| $V\approx0.10$ | Small / Weak |
| $V\approx0.30$ | Medium / Moderate |
| $V\approx0.50$ | Large / Strong |
:::

::: {.fragment .mt1}
- **Applying it to our penguins:**
    + We previously found a Cramer's $V$ of 0.66 for Species and Island.
    + According to these guidelines, this is a *strong* association!
    
:::

# Wrap-Up

## Choosing the Right Test {.smaller}

You now have three primary tools for testing basic bivariate relationships<br><br>

| Variable 1 | Variable 2 | Test | R Function |
|:---|:---|:---|:---|
| Categorical | Continuous | **ANOVA** | `aov_ez()` |
| Continuous | Continuous | **Correlation** | `correlation()` |
| Categorical | Categorical | **Chi-Square** | `chisq.test()` |

<br><br>
*Note: "Bivariate" means we are only looking at two variables at a time.<br>In future courses, you will learn how to build models with many variables!*
