---
lecture: "04b"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| label: setup
#| include: false

options(
  tibble.print_min = 8, 
  tibble.print_max = 8
)
library(here)
source(here("_common.R"))
```


## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Transform data by row using `arrange()`, `filter()`, and `distinct()`

- Transform data by column using `mutate()`, `select()`, `rename()`, and `relocate()`
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

## Basic wrangling verbs {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   [tidyverse]{.b .blue} provides tools for wrangling tibbles
    -   These functions are named after *verbs*
    -   So if you name your objects after *nouns*...
    -   ...your code becomes easier to read

::: {.fragment .mt1}
| Noun(noun) ❌        |  Verb(noun)  ✔️  | 
|:----------------------|:----------------|
| `blender(fruit)`      | `blend(fruit)`  |
| `screwdriver(screw)`  | `drive(screw)`  |
| `boxcutter(box)`      | `cut(box)`      |

:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/wrench.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Important: Saving changes {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   `dplyr` functions **do not** modify the original
    -   Running a command returns a *copy* of the result
    -   It does not change the original object!
    -   This is useful in *teaching* but less so in practice

::: {.fragment .mt1}
-   To **save** changes, you must use assignment [`<-`]{.b .blue}
    -   Print only: `arrange(data, var)`
    -   Save/update: `data <- arrange(data, var)`
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/caution.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Setup

```{r}
#| message: false

library(tidyverse)
data("starwars", package = "dplyr")
```

```{r}
glimpse(starwars)
```

# Rows

## Row-focused verbs {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Arrange]{.b .blue}
    -   Reorders the rows according to the values in specified columns

::: {.fragment .mt1}
-   [Filter]{.b .blue}
    -   Retains only the rows that satisfy your specified conditions
:::

::: {.fragment .mt1}
-   [Distinct]{.b .blue}
    -   Removes duplicate rows to retain only unique combinations of values
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/rowing.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

# Arrange

## Arrange alphabetically from A to Z

```{r}
# Sort the rows by the name variable (ascending)
arrange(starwars, name)
```

## Arrange alphabetically from Z to A

```{r}
# Sort the rows by the name variable (descending)
arrange(starwars, desc(name))
```

## Arrange numerically

```{r}
# Sort the rows by height from shortest to tallest
arrange(starwars, height)
```

## Arrange by multiple variables

```{r}
# Sort by multiple variables to break ties
arrange(starwars, hair_color, mass)
```

# Filter

## Filter by exact match

```{r}
# Retain only rows where the species is exactly "Droid"
filter(starwars, species == "Droid")
```

## Filter by numeric condition

```{r}
# Retain only characters with a body mass greater than or equal to 100
filter(starwars, mass >= 100)
```

## Filter by multiple conditions (AND)

```{r}
# Retain "Droid" characters AND those with mass < 50
filter(starwars, species == "Droid", mass < 50)
```

## Filter by multiple conditions (OR)

```{r}
# Retain characters that have sex equal to male OR none
filter(starwars, sex == "male" | sex == "none")
```

# Distinct

## Find distinct values

```{r}
# Find the unique homeworlds listed in the dataset
distinct(starwars, homeworld)
```

## Find distinct combinations

```{r}
# Find unique combinations of species and sex
distinct(starwars, species, sex)
```

# Columns

## Column-focused verbs {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Select]{.b .blue}
    -   Retains specific columns and drops the rest

::: {.fragment .mt1}
-   [Mutate]{.b .blue}
    -   Creates new columns or modifies existing ones
:::

::: {.fragment .mt1}
-   [Rename]{.b .blue}
    -   Changes the names of specific columns
:::

::: {.fragment .mt1}
-   [Relocate]{.b .blue}
    -   Changes the positions of specific columns
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/pillar.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

# Select

## Select specific columns

```{r}
# Retain only the name, height, and mass columns
select(starwars, name, height, mass)
```

## Drop specific columns

```{r}
# Retain all columns EXCEPT films, vehicles, and starships
select(starwars, -films, -vehicles, -starships)
```

## Select all between two columns

```{r}
# Retain all columns between height and gender
select(starwars, height:gender)
```

## Select using helpers

```{r}
# Retain only columns that end with the phrase "color"
select(starwars, ends_with("color"))
```

:::{.fsmaller}
See also: `starts_with()`, `contains()`, `matches()`, and `where()`
:::

# Mutate

## Mutate new columns

```{r}
# Create a new column and add all the way to the right
mutate(starwars, height_m = height / 100)
```

## Mutate existing columns

```{r}
# Modify the height variable "in place"
mutate(starwars, height = height * 10)
```

# Rename and Relocate

## Rename columns

```{r}
# Change the name of 'homeworld' to 'planet' (new = old)
rename(starwars, planet = homeworld)
```

## Relocate columns

```{r}
# Move the 'species' column to the very first position
relocate(starwars, species, .before = name)
```

:::{.fsmaller}
Can also use `.after` and position numbers, e.g., `.after = 1`
:::
