---
lecture: "04c"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

```{r}
#| include: false
#| label: setup

library(here)
source(here("_common.R"))
options(
  tibble.print_min = 5,
  tibble.print_max = 5
)
```

## Setup

```{r}
#| message: false

library(tidyverse)
data("starwars", package = "dplyr")
```

```{r}
glimpse(starwars)
```

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
- Use the pipe (`|>`) to efficiently pass an object to a function

- Build a pipeline by connecting multiple pipes together
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# The Pipe

## The Pipe Operator {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   The pipe passes the object on the **left** to the<br>first argument of the function on the **right**
    -   Think of it as reading **"and then..."**
    -   e.g., take `x` **and then** square root it

::: {.fragment .mt1}
-   **Why use it?**
    1.  **Readability:** Code reads left-to-right (like English), not inside-out (like math functions).
    2.  **Cleanliness:** Avoids creating temporary variables for every single step.
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/arrow.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## RStudio Setup & Shortcuts {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
1. **Configure RStudio**
    -   Go to **Tools** > **Global Options**
    -   Click **Code** (on the left)
    -   Check box: ["Use native pipe operator"]{.b .blue}
    -   Click **Apply** / **OK**
    
2. **Keyboard Shortcut**
    -   Let RStudio insert `|>` for you
    -   Win: {{< kbd Ctrl+Shift+M >}}
    -   Mac: {{< kbd Cmd+Shift+M >}}
    -   Without the setting above, you get `%>%` which is similar but requires **tidyverse** to be loaded
    
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/keyboard.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Piping to a simple function

```{r}
# Traditional way
sqrt(100)
```

::: {.fragment}
```{r}
# Piped way
100 |> sqrt()
```
:::

::: {.fragment}
```{r}
# Saving the output
x <- 100 |> sqrt()
x
```
:::

## Piping with extra arguments

```{r}
# Traditional way
round(3.14159, digits = 2)
```

::: {.fragment}
```{r}
# Piped way
3.14159 |> round(digits = 2)
```
:::

::: {.fragment}
```{r}
#| error: true

# Be careful not to give x twice
x <- 3.14159 
x |> round(x, digits = 2)
```
:::

## Piping to other arguments

```{r}
seq(from = 1, to = 10)
```

:::{.fragment}
```{r}
# The pipe defaults to filling in the *first* argument of the function
1 |> 
  seq(to = 10)
```
:::

::: {.fragment}
```{r}
# But, if needed, we can use the _ placeholder to pipe to a different argument
10 |> 
  seq(from = 1, to = _)
```
:::

## Bridging to dplyr

```{r}
# Traditional way (hard to read)
x <- select(starwars, name, species, homeworld)
```

::: {.fragment}
```{r}
# Piped way ("Take starwars AND THEN select...")
x <- 
  starwars |> 
  select(name, species, homeworld)
x
```
:::

# Pipelines

## Building Pipelines {.smaller}

::: {.columns .pv4}
::: {.column width="60%"}
-   We rarely do just one thing to data
    -   We usually want to filter, *and then* select, *and then* arrange...
-   We can chain multiple pipes together
    -   This is called building a [pipeline]{.b .blue}
    -   The result of each line becomes the input for the next line, creating a kind of "assembly line"

::: {.fragment .mt1}
**Formatting Tip:**

Add a line break after each pipe `|>` and then indent the next line. RStudio will do this automatically.
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/flow.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Traditional Approaches

**Intermediate Assignment**

```{r}
# Error-prone: easy to use wrong variable name or overwrite data
x <- 100
y <- sqrt(x)
z <- log(y, base = 2)
z
```

:::{.fragment}
**Nested Functions**

```{r}
# Difficult to read and add on to, especially as this gets longer
z <- log(sqrt(100), base = 2)
z
```
:::

## Compare: The Pipeline

```{r}
# Clean, linear, and easy to read and add on to
z <- 
  100 |> 
  sqrt() |> 
  log(base = 2)
z
```

:::{.fragment}
```{r}
# We can even add print to the end to save and print
z <- 
  100 |> 
  sqrt() |> 
  log(base = 2) |> 
  print()
```
:::

## Data Pipeline 1

```{r}
# What are the names and genders of all the Droids?
out <- 
  starwars |> 
  filter(species == "Droid") |> 
  select(name, gender) |> 
  print()
```

## Data Pipeline 2

```{r}
# Which characters have a BMI (Body Mass Index) greater than 30?
out <- 
  starwars |> 
  mutate(bmi = mass / ((height / 100)^2)) |> 
  filter(bmi > 30) |> 
  arrange(desc(bmi)) |> 
  select(name, bmi, mass, height) |> 
  print()
```

## Data Pipeline 3

```{r}
# What unique homeworlds are represented by the Human characters?
out <-
  starwars |>
  filter(species == "Human") |>
  distinct(homeworld) |>
  arrange(homeworld) |>
  print()
```

## Data Pipeline 4

```{r}
# Which non-humans are shorter than 100cm? (Rename height for clarity)
out <-
  starwars |>
  filter(species != "Human", height < 100) |>
  rename(height_cm = height) |> 
  select(name, species, height_cm) |>
  print()
```
