---
lecture: "03a"
metadata-files:
  - ../../_slide-settings.yml
format:
  revealjs:
    theme: default
---

::: {.my-title}
# [Foundations of]{.my-subtitle} <br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=30 right=0 width="400"}
:::

## Roadmap

::: {.columns .pv4}

::: {.column width="60%"}
1. Store reusable data in named objects using assignment (`<-`)

2. Transform data and objects using one or more functions
:::

::: {.column .tc .pv4 width="40%"}
{{< lif "../../icons/map.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::

:::

# Assignment

## Assignment {.fs70}

::: {.columns .pv4}
::: {.column width="60%"}
-   It is often useful to store data in [named objects]{.b .green}
    -   This makes the data easier to use and re-use
    -   This makes the code easier to write and read

::: {.fragment .mt1}
-   Which command is easier to follow?
    1.  Dial *7 8 5 8 6 4 0 8 4 1*
    2.  Call *Office Phone*
:::

::: {.fragment .mt1}
-   Named objects are created using [assignment]{.b .blue}
    -   Give a **name** then an **arrow** then the **data**
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/contacts.json" trigger=hover colors=secondary:#2a76dd class=rc >}}

::: {.fragment}
`office <- 7858640841`
:::
:::
:::

## Naming {.fs70}

::: {.columns .pv4}
::: {.column width="60%"}
-   Object names can [only include]{.b .blue}:
    -   Letters: `a-Z`
    -   Numbers: `0-9`
    -   Underscores: `_`
    -   Periods: `.`

::: {.fragment .mt1}
-   Additional Rules:
    -   Must [start with a letter or period]{.b .blue}
    -   Cannot contain spaces or dashes
    -   Cannot contain other symbols
    -   Names are case-sensitive (`age` ≠ `Age`)
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/visa.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::

## Basic Assignment

```{r}
# x gets 2, print x
x <- 2
x

# Make y from x, print y
y <- 3 * x
y
```

## Updating Objects

```{r}
# Using x will not update it
x + 1

# x is still 2
x

# Use <- again to update x
x <- x + 1
x
```

## Names are case-sensitive

```{r}
# We already created lowercase y with assignment
y
```

```{r}
#| error: true

# But we have not yet created uppercase Y
Y
```

## Finding the right balance

```{r}
# rate is allowed but pretty short
rate <- 93
rate

# heart_rate_in_beats_per_minute is allowed but pretty long
heart_rate_in_beats_per_minute <- 93
heart_rate_in_beats_per_minute

# heart_rate_bpm might be a nice compromise
heart_rate_bpm <- 93
heart_rate_bpm
```

## Be careful of special symbols

```{r}
#| error: true

# Don't try to add spaces
second age <- 12
```

```{r}
#| error: true

# Don't try to add special symbols
age@time2 <- 12
```

```{r}
# Focus on letters, numbers, and underscores
age_time2 <- 12
age_time2
```


# Functions

## Functions {.fs70}

::: {.columns .pv4}
::: {.column width="60%"}
-   [Functions]{.b .blue} are like calculator buttons
    -   They let you easily apply a transformation
    -   They can have descriptive names for readability
    -   Is it obvious `x^0.5` is the square root of x?
    -   Is it obvious `sqrt(x)` is the square root of x?

::: {.fragment .mt1}
-   We may want to combine multiple functions
    -   We *could* use intermediate objects...
    -   But "nested" functions take less room
    -   We resolve nested functions inside-out
    -   Using "pipes" is even better \[04c\]
    
:::
:::

::: {.column .tc .pv5 width="40%"}
{{< lif "../../icons/fn_key.json" trigger=hover colors=secondary:#2a76dd class=rc >}}
:::
:::


## Calling Functions

```{r}
# Can I turn this into a "button" to press?
9^0.5

# Take the square root of a number
x <- sqrt(9)
x

# Or take the square root of an object
y <- sqrt(x)
y
```

## Rounding Functions

```{r}
# Round up
ceiling(1 / 3)

# Round down
floor(1 / 3)

# Round to nearest even integer
round(0.5)

# Round to nearest even integer
round(1.5)
```


## Other Useful Functions

```{r}
# Absolute Value
abs(-3)

# Sign
sign(-3)

# Logarithm
log(1)

# Exponential
exp(1)
```

## Basic Trigonometry

```{r}
# Sine
sin(1)

# Cosine
cos(1)

# Tangent
tan(1)
```

:::{.fs80}
Note: there are corresponding `asin()`, `acos()`, and `atan()` functions.
:::

## Combining Functions

```{r}
# Using intermediate objects works but takes many lines
x <- sqrt(9)
y <- cos(x)
z <- round(y)
z

# Alternatively, "nest" them and then resolve inside-out
z <- round(cos(sqrt(9)))
z
```
