---
title: "Week 03 Assignment"
format: 
  html:
    css: ../../styles.css
    embed-resources: true
---

![](../../img/programming_2780E3.svg){width="200"}

## Instructions

*Create a new Quarto document (HTML format), clear out the boilerplate code/text, and use this document to store your answers to each of the six questions below. Be sure to enable `embed-resources` using the header below. If your answer involves writing text, then use markdown to write 2--4 sentences per question; if your answer involves running code, then use one or more code chunks. For questions that ask you to fix code, correct any errors (while achieving the stated goal) and include only the working code in your document.*

```markdown
---
title: "Week 03 Assignment"
author: "Your Name"
embed-resources: true
---
```

*Please clearly mark which part of your document is answering which question (e.g., using markdown headings) and remove any "scratch work" or failed attempts from your document before submitting it. Finally, note that many essay-style questions will require you to read between the lines and use your imagination (as opposed to just looking up the answer in the lecture).*


## Question 1 (Assignment)

a) Why is it important to give your objects in R good names? {{< hp glyph="." >}}When might poorly thought-out names cause problems for a programmer?

b) Help me fix my code! "I want to calculate the average of three objects corresponding to negative emotion self-reports."

```{r}
#| error: true
#| collapse: true

anger <- 3
sadness <- 4
fear <- 3
negative_emotion <- (anger + sadness + faer) / 3
```

## Question 2 (Functions)

a) In your own words, explain what functions and arguments are (in the context of programming) {{< hp glyph="." >}}and why they are useful in data science. 

b) Use nested functions and operations to achieve all of the following in a **single** line of code: multiply 45 by 3.14, then divide the result by 180, then calculate that value's sine, and then round the result to the nearest even integer.

## Question 3 (Arguments)

a) In your own words, explain what a "default value" is for an argument {{< hp glyph="." >}}and why it might be important for the people who create functions to choose those default values carefully.

b) Help me fix my code! "I want to round 1/16 to three decimal places."

```{r, renv.ignore=TRUE}
#| error: true
#| collapse: true

round(1 / 16; places = 3)
```


## Question 4 (Vectors)

a) What did I mean when I said that "functions typically either transform each element in a vector or summarize all elements in a vector"? {{< hp glyph="." >}}In your answer, provide an example (in text) of when each of these behaviors might be useful.

b) Below is a table of the top 5 countries' production (i.e., [GDP Per Capita at PPP](https://www.worldometers.info/gdp/gdp-by-country/)) in 2017. {{< hp glyph="." >}}Save the numbers only to a vector and then use R to create a new vector that contains the log-transformed versions of each number. Print that new vector.

|Country|Trillions |
|-------|----------|
|USA    |19.485    |
|China  |12.238    |
|Japan  |4.872     |
|Germany|3.693     |
|India  |2.651     |

{{< dstart summary="Hint" >}}
Your output should look something like this:

```{r}
#| echo: false

gdp <- c(19.485, 12.238, 4.872, 3.693, 2.651)
log(gdp)
```

{{< dstop >}}


## Question 5 (Strings)

a) Save the country names from the table in Question 4 to a character vector called `country`. {{< hp glyph="." >}}Calculate the number of characters in each country name and print that.

b) Use a function in R to calculate the average number of characters in each country name (i.e., use the function I taught you in class).


## Question 6 (Packages)

a) R packages can be found in a repository (e.g., CRAN), in a user library, or loaded into an R session. {{< hp glyph="." >}}When would you want an R package to be in each of these places?

b) Install the `praise` package (outside of Quarto). Then load the installed package in your Quarto document and experiment with the `praise()` function by running it several times with no arguments. What does the function seem to do?


---

✅ Done! *Please save and then render your document to an HTML file. Check it over to make sure it is complete (and rendered properly) and then upload this file to Canvas by the deadline listed on the syllabus. Note that, if any of your code chunks has an error in it, the HTML file will not render properly (this is why removing scratch work is important). If you receive an error message upon clicking "Render" about a missing file, this actually means that there is an error in one of your chunks.*
