````{verbatim, lang = "qmd"}
#| echo: true

# Suppress messages
```{r}
library(tidyverse)
```

```{r}
#| message: false

library(tidyverse)
```

# Suppress warnings

```{r}
library(palmerpenguins)
ggplot(penguins, aes(x = body_mass_g)) + geom_histogram()
```

```{r}
#| message: false

library(palmerpenguins)
ggplot(penguins, aes(x = body_mass_g)) + geom_histogram()
```

```{r}
#| message: false
#| warning: false

library(palmerpenguins)
ggplot(penguins, aes(x = body_mass_g)) + geom_histogram()
```

# Include code but don't run it
```{r}
1 + 2
```

```{r}
#| eval: false

1 + 2
```

# Hide code but still run it
```{r}
1 + 2
```

```{r}
#| echo: false

1 + 2
```

# Include a chunk with an error in it
```{r}
mean()
```

```{r}
#| error: true

mean()
```

# Fold code for more compact output
`#| code-fold: true` or `#| code-fold: show`

```{r}
#| code-fold: true

pi * 10^2
```

# Collapse output into the code block

```{r}
#| collapse: true

pi * 10^2
```

# Control figure size
```{r}
ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
```

```{r}
#| fig-width: 10
#| fig-height: 5

ggplot(mpg, aes(x = displ, y = hwy)) + geom_point()
```

# Name a chunk
```{r setup}
library(tidyverse)
```

# Cache (save) a slow chunk
```{r diamondsplot}
#| cache: true

ggplot(diamonds, aes(x = carat, y = price)) + geom_point()
```

````
