---
format: 
  revealjs:
    css: ../../styles.css
    slide-number: true
    show-slide-number: all
    progress: true
    history: true
    hash-type: number
    theme: default
    code-block-background: true
    highlight-style: zenburn
    code-link: false
    code-copy: true
    pagetitle: "Foundations of Data Science"
    author-meta: "Jeffrey Girard"
    date-meta: "2023-03-22"
    semester: "Spring 2023"
    course: "399/800"
    lecture: "10b"
---

::: {.my-title}
# [Foundations of]{.my-subtitle}<br />[Data Science]{.blue}

::: {.my-grey}
[{{< meta semester >}} | Course {{< meta course >}}]{}<br />
[Jeffrey M. Girard | Lecture {{< meta lecture >}}]{}
:::

![](../../img/proud_coder_357EDD.svg){.absolute bottom=0 right=0 width=400}
:::

## Roadmap: Convenience

::: {.columns .pv4}

::: {.column width="60%"}
1. Tidy Selection
  
2. Across

3. Scale Scores

4. Activity
:::

::: {.column .tc .pv4 width="40%"}
{{< li xtkehzkm trigger=loop delay=5000 colors=secondary:#2a76dd class=rc >}}
:::

:::

# Tidy Selection

## Tidy Selection {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   `filter()` lets us retain rows using criteria
    -   A version for columns would be convenient!

::: {.fragment .mt1}
-   This is essentially what tidy selection does
    -   We can `select()` columns using criteria
    -   It's like an extension of the `:` sugar
:::

::: {.fragment .mt1}
-   We will use selection helpers and predicates
    -   `starts_with()`, `contains()`, etc.
    -   `everything()`, `last_col()`
    -   `where()` and `is.*()`

:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li tvyxmjyo trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Tidy Selection Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

library(palmerpenguins)

penguins

# Starts with

penguins |> 
  select(starts_with("bill_"))

# PITFALL

penguins |> 
  starts_with("bill_") # error - don't forget select()
  

# Ends with

penguins |> 
  select(ends_with("_mm"))

# Contains

penguins |> 
  select(contains("_length_"))

# Where

penguins |> 
  select(where(is.factor))

penguins |> 
  select(where(is.numeric))

penguins |> 
  select(where(is.character))

# PITFALL

penguins |> 
  select(where(is.factor())) # error - don't include () after is.*

# Everything

penguins |> 
  select(everything())

# Last Column

penguins |> 
  select(last_col())

penguins |> 
  select(body_mass_g:last_col())

# Negating Criteria

penguins |> 
  select(!ends_with("_mm"))

# Combining Criteria

penguins |> 
  select(where(is.factor) | starts_with("body_"))

penguins |> 
  select(species, where(is.numeric))

penguins |> 
  select(ends_with("_mm") & !contains("_depth_"))
```


# Across

## Across {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   We can apply functions to many columns at once
    -   This is done using the `across()` function
    -   It goes inside `mutate()` and `summarize()`

::: {.fragment .mt1}
-   `across()` pairs very well with tidy selection
    -   e.g., do this to variables starting with "x"
    -   e.g., do this to all numeric variables
:::

::: {.fragment .mt1}
-   Advanced: "Anonymous functions"
    -   Call current col x, then round x to 2 digits
    -   e.g., `\(x) round(x, digits = 2)`
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li krmfspeu trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Across Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

starwars

# Applying the same mutation to multiple variables is a pain
starwars |> 
  mutate(
    hair_color = factor(hair_color),
    skin_color = factor(skin_color),
    eye_color = factor(eye_color),
  )

# dplyr 1.0.0 added across() to make this easier

starwars |> 
  mutate(
    across(
      .cols = c(hair_color, skin_color, eye_color), 
      .fns = factor
    )
  )

# We can also use tidyverse shortcuts like the colon (:)
starwars |> 
  mutate(
    across(
      .cols = c(hair_color:eye_color), 
      .fns = factor
    )
  )

# We can also compress it to take up less space
starwars |> 
  mutate(across(c(hair_color:eye_color), factor))

# We can also use a tidy selection helper when appropriate
starwars |> 
  mutate(
    across(
      .cols = ends_with("color"), 
      .fns = factor
    )
  )

starwars |>  
  mutate(across(.cols = where(is.character), .fns = factor))

# For more control, we use an anonymous function

starwars |> 
  mutate(
    height = height * 1000,
    mass = mass * 1000
  )

starwars |> 
  mutate(
    across(
      .cols = c(height, mass), 
      .fns = \(x) x * 1000
    )
  )

# Across with summarize

starwars |> 
  summarize(
    height = mean(height, na.rm = TRUE),
    mass = mean(mass, na.rm = TRUE),
    birth_year = mean(birth_year, na.rm = TRUE)
  )

starwars |> 
  summarize(
    across(
      .cols = c(height, mass, birth_year), 
      .fns = \(x) mean(x, na.rm = TRUE)
    )
  )
```


# Scale Scores

## Scale Scores {.smaller}

::: {.columns .pv4}

::: {.column width="60%"}
-   In social science, we often use [scale scores]{.b .blue}
    -   Multiple similar items are combined
    -   Sums and means are both common

::: {.fragment .mt1}
-   Calculating scale scores from items is tricky
    -   When each item is a column...
    -   ...we calculate scale scores **per row**
:::

::: {.fragment .mt1}
-   We can either group per row (using `.by`)
    -   Or use special row-based functions
:::
:::

::: {.column .tc .pv4 width="40%"}
{{< li bdfzpkxv trigger=loop delay=3000 colors=secondary:#2a76dd class=rc >}}
:::

:::

## Scale Scores Live Coding

```{r}
#| echo: true
#| eval: false
#| error: true
#| code-line-numbers: false

# Simulate a longitudinal dataset with a five-item scale
df <- 
  tibble(
    subject = rep(101:103, each = 3),
    visit = rep(1:3, times = 3),
    q1 = sample(x = 0:4, size = 9, replace = TRUE),
    q2 = sample(x = 0:4, size = 9, replace = TRUE),
    q3 = sample(x = 0:4, size = 9, replace = TRUE),
    q4 = sample(x = 0:4, size = 9, replace = TRUE),
    q5 = sample(x = 0:4, size = 9, replace = TRUE),
  ) |> 
  print()

# USECASE: Calculate subjects' scores per visit

# Solution 1 (calculate sums and means "by-hand")
df |> 
  mutate(
    total_q = q1 + q2 + q3 + q4 + q5,
    mean_q = (q1 + q2 + q3 + q4 + q5) / 5
  )

# PITFALL: Attempt to use functions for convenience
df |> 
  mutate(
    total_q = sum(c(q1, q2, q3, q4, q5)),
    mean_q = mean(c(q1, q2, q3, q4, q5))
  )
# But the answers aren't what we wanted! These are ungrouped mutates!

# Solution 2 (turn them into grouped mutates)
df |> 
  mutate(
    total_q = sum(c(q1, q2, q3, q4, q5)),
    mean_q = mean(c(q1, q2, q3, q4, q5)),
    .by = c(subject, visit)
  )
# These are the answers we wanted

# USECASE: Can we use tidy selection here?

# The : operator does work if they are in order!
df |> 
  mutate(
    total_q = sum(q1:q5),
    mean_q = mean(q1:q5),
    .by = c(subject, visit)
  )

# But starts_with() doesn't...
df |> 
  mutate(
    total_q = sum(starts_with("q")),
    mean_q = mean(starts_with("q")),
    .by = c(subject, visit)
  )
# Error...

# Solution 3 (use row functions and pick, no .by necessary)
df |> 
  mutate(
    total_q = rowSums(pick(starts_with("q"))),
    mean_q = rowMeans(pick(starts_with("q")))
  )

# PITFALL (don't include your sum in your mean)
df |> 
  mutate(
    q_total = rowSums(pick(starts_with("q"))),
    q_mean = rowMeans(pick(starts_with("q")))
  )
# Note that the means are now much higher because they include q_total
# This is because q_total also starts with q and comes first in mutate()
```
